### 2021-09-23 sample-authordraft.tex
Initial Overleaf Import

%%
%% This is file `sample-manuscript.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `manuscript')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-manuscript.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
%%%% Small single column format, used for CIE, CSUR, DTRAP, JACM, JDIQ, JEA, JERIC, JETC, PACMCGIT, TAAS, TACCESS, TACO, TALG, TALLIP (formerly TALIP), TCPS, TDSCI, TEAC, TECS, TELO, THRI, TIIS, TIOT, TISSEC, TIST, TKDD, TMIS, TOCE, TOCHI, TOCL, TOCS, TOCT, TODAES, TODS, TOIS, TOIT, TOMACS, TOMM (formerly TOMCCAP), TOMPECS, TOMS, TOPC, TOPLAS, TOPS, TOS, TOSEM, TOSN, TQC, TRETS, TSAS, TSC, TSLP, TWEB.
% \documentclass[acmsmall]{acmart}

%%%% Large single column format, used for IMWUT, JOCCH, PACMPL, POMACS, TAP, PACMHCI
% \documentclass[acmlarge,screen]{acmart}

%%%% Large double column format, used for TOG
% \documentclass[acmtog, authorversion]{acmart}

%%%% Generic manuscript mode, required for submission
%%%% and peer review
\documentclass[manuscript,screen,review]{acmart}
%% Fonts used in the template cannot be substituted; margin 
%% adjustments are not allowed.
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{fastReach: Increasing Interaction Speed using Cyberphysical Systems with EEG Readiness Potentials}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ben Trovato}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

\author{Lars Th{\o}rv{\"a}ld}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}

\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}

\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \streetaddress{8600 Datapoint Drive}
  \city{San Antonio}
  \state{Texas}
  \country{USA}
  \postcode{78229}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{%
  \institution{The Kumquat Consortium}
  \city{New York}
  \country{USA}}
\email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{datasets, neural networks, gaze detection, text tagging}

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
\begin{teaserfigure}
  \includegraphics[width=\textwidth]{sampleteaser}
  \caption{Seattle Mariners at Spring Training, 2010.}
  \Description{Enjoying the baseball game from the third-base
  seats. Ichiro Suzuki preparing to bat.}
  \label{fig:teaser}
\end{teaserfigure}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\verb|acmsmall|}: The default journal template style.
\item {\verb|acmlarge|}: Used by JOCCH and TAP.
\item {\verb|acmtog|}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\verb|acmconf|}: The default proceedings template style.
\item{\verb|sigchi|}: Used for SIGCHI conference articles.
\item{\verb|sigchi-a|}: Used for SIGCHI ``Extended Abstract'' articles.
\item{\verb|sigplan|}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\verb|anonymous,review|}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \verb|\acmSubmissionID| command to print the
  submission's unique ID on each page of the work.
\item{\verb|authorversion|}: Produces a version of the work suitable
  for posting by the author.
\item{\verb|screen|}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[sigconf,authordraft]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification. Multiple authors may share one affiliation. Authors'
names should not be abbreviated; use full first names wherever
possible. Include authors' e-mail addresses whenever possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}. Artifacts:
  \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{itemize}
\item {\verb|sidebar|}:  Place formatted text in the margin.
\item {\verb|marginfigure|}: Place a figure in the margin.
\item {\verb|margintable|}: Place a table in the margin.
\end{itemize}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-authordraft.tex'.

### 2021-09-23 processing/data_exploration.m
exploring readiness potential

@@ -1,0 +1,204 @@
+EEG = pop_loadset('filename','sub-1_preprocessed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/3_EEG-preprocessing/sub-1/');
+motion = pop_loadset('filename','sub-1_motion_processed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/6_single-subject-motion-analysis/sub-1/');
+physio = pop_loadset('filename','sub-1__PHYSIO.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/2_raw-EEGLAB/sub-1/');
+
+events = EEG.event;
+motion.event = events;
+physio.event = events;
+
+fR.p.movement_class = {'trial_start', [0, 10]};
+fR.p.idle_class = {'trial_end', [1, 2]};
+
+% ignore first second after free to move epoch start
+fR.p.ignore_early_samples = EEG.srate;
+
+% movement detector for both hand velocity and emg
+fR.p.raw_thresh = .7;
+fR.p.fine_thresh = .05;
+fR.p.window_for_fine_thresh = EEG.srate/2;
+
+% feature windows
+fR.p.feature.window_size = EEG.srate;
+fR.p.feature.nr_windowed_means = 10;
+
+%% signal processing EEG and Physio data
+
+EEG = pop_eegfiltnew(EEG, .1, 20);
+physio = pop_eegfiltnew(physio, .1, 10);
+
+% motion: overwrite random channel with magnitude, only need that!
+motion.data(1,:) = sqrt(motion.data(4,:).^2 + motion.data(5,:).^2 + motion.data(6,:).^2);
+
+%% MOVEMENT Class: extract movement onsets from Velocity and EMG
+
+%%% get hand movement onsets and epochs from velocity -> defines movement class for classifier
+tmp = pop_epoch(motion, fR.p.movement_class(1), fR.p.movement_class{2});
+fR.move.vel.mag = squeeze(tmp.data(1,:,:));
+for i = 1:size(fR.move.vel.mag,2)
+    if fR.p.ignore_early_samples
[…]

### 2021-09-23 processing/extract_EEG_features.m

@@ -1,0 +1,63 @@
+function [sp] = extract_EEG_features(EEG, onsets, event_str, epoch_classes)
+%EXTRACT_EEG_FEATURES Summary of this function goes here
+%   Detailed explanation goes her
+
+%% signal processing
+
+EEG = pop_eegfiltnew(EEG, .1, 20);
+
+%% select sample in pre-movement class
+
+EEG_mov = pop_epoch( EEG, {event_str}, epoch_classes{1});
+
+for i = 1:numel(onsets)
+    premovement_class(:,:,i) = EEG_mov.data(:,onsets-249:onsets,i);
+end
+
+%% select samples in idle class
+
+EEG_idle = pop_epoch( EEG, {event_str}, epoch_classes{2});
+
+%% baseline correct
+
+premovement_class = premovement_class - mean(premovement_class(:,1:25,:),2);
+EEG_idle.data = EEG_idle.data - mean(EEG_idle.data(:,1:25,:),2);
+
+%% select discriminatory channels: which channels show negative going RP
+% the signal in the last 200ms is smaller than in the first 200ms
+
+for c = 1:numel(EEG.chanlocs)
+    sp.rp_move(c,:) = squeezemean(premovement_class(c,end-49:end,:),2) - squeezemean(premovement_class(c,1:50,:),2);
+    [H(c),P(c),CI,STATS] = ttest(sp.rp_move(c,:));
+    t(c) = STATS.tstat;
+end
+
+sp.selected_chans_rp = intersect(find(t<0), find(P < .05));
+% {EEG.chanlocs(sp.selected_chans_rp).labels}
+
+%% which channels show the negative going deflection only in the movement class
+
+for c = 1:numel(EEG.chanlocs)
+    sp.rp_idle(c,:) = squeezemean(EEG_idle.data(c,end-49:end,:),2) - squeezemean(EEG_idle.data(c,1:50,:),2);
+    [H(c),P(c),CI,STATS] = ttest(sp.rp_idle(c,:));
+    t(c) = STATS.tstat;
+end
+
+sp.selected_chans_rp_class_dep = find(mean(sp.rp_move,2) < mean(sp.rp_idle,2));
+% {EEG.chanlocs(sp.selected_chans_rp_class_dep).labels}
+
+%% merge selected channels
+
+sp.selected_channels = intersect(sp.selected_chans_rp, sp.selected_chans_rp_class_dep);
[…]

### 2021-09-23 processing/extract_motion_features.m

@@ -1,0 +1,22 @@
+function [sp] = extract_motion_features(events, motion, event_str, epoch, buffer)
+%EXTRACT_MOTION_FEATURES
+
+%% detect motion onset samples to build pre-movement class
+
+motion.data(1,:) = sqrt(motion.data(4,:).^2 + motion.data(5,:).^2 + motion.data(6,:).^2);
+motion = pop_epoch( motion, {event_str}, epoch);
+
+sp.mag = squeeze(motion.data(1,:,:));
+
+for i = 1:size(sp.mag,2)
+    
+    if buffer
+        sp.onset_sample(i) = fR_movement_onset_detector(sp.mag(buffer:end,i), .7, 125, .05);
+        sp.onset_sample(i) = sp.onset_sample(i) + buffer;        
+    else
+        sp.onset_sample(i) = fR_movement_onset_detector(sp.mag(:,i), .7, 125, .05);
+    end 
+end
+
+end
+

### 2021-09-23 processing/extract_physio_features.m

@@ -1,0 +1,28 @@
+function [sp] = extract_physio_features(events, physio, event_str, epoch, buffer)
+
+%% signal processing
+physio = pop_eegfiltnew(physio, .1, 10);
+
+%% events
+physio.event = events;
+physio = eeg_checkset(physio);
+physio = pop_epoch( physio, {event_str}, epoch);
+
+%% EMG
+for i = 1:size(physio.data,3)
+    sp.emg.data(i,:) = squeeze(physio.data(2,:,i));
+    [~,sp.emg.ylower(i,:)] = envelope(sp.emg.data(i,:),1,'peak');
+    
+    if buffer
+        sp.emg.onset_sample(i) = fR_movement_onset_detector(sp.emg.ylower(i,buffer:end)*-1, .7, 125, .05);
+        sp.emg.onset_sample(i) = sp.emg.onset_sample(i) + buffer;        
+    else
+        sp.emg.onset_sample(i) = fR_movement_onset_detector(sp.emg.ylower(i,:)*-1, .7, 125, .05);
+    end 
+    
+end
+
+%% EDA
+
+end
+

### 2021-09-23 processing/fR_movement_onset_detector.m

@@ -1,0 +1,28 @@
+function [onset_sample] = fR_movement_onset_detector(motion_segment, raw_threshold, pre_raw_window, fine_threshold)
+%FR_MOVEMENT_ONSET_DETECTOR
+
+dist = max(motion_segment) - min(motion_segment);
+first_thresh = min(motion_segment) + dist*raw_threshold;
+movement_detected = min(find(motion_segment>first_thresh));
+ 
+% if size(motion_segment,1) > size(motion_segment,2)
+%     motion_segment = motion_segment';
+% end
+
+try
+    mov_rev = fliplr(motion_segment(movement_detected-pre_raw_window:movement_detected));
+    dist = max(mov_rev) - min(mov_rev);
+    second_thresh = min(mov_rev) + dist*fine_threshold;
+
+    movement_detected_fine = min(find(mov_rev<second_thresh));
+    onset_sample = movement_detected - movement_detected_fine;
+catch
+    warning('event occurred too early');
+    onset_sample = [];
+
+if isempty(onset_sample)
+    onset_sample = 0;
+end
+
+end
+

### 2021-09-23 processing/fastReach_bemobil_bids_metadata.m

@@ -1,0 +1,80 @@
+% general metadata shared across all modalities
+%--------------------------------------------------------------------------
+%--------------------------------------------------------------------------
+general_info = [];
+
+% root directory (where you want your bids data to be saved)
+general_info.bidsroot                                = fullfile(bemobil_config.study_folder, bemobil_config.bids_data_folder); 
+
+% required for dataset_description.json
+general_info.dataset_description.Name                = 'Walking task in the young and old';
+general_info.dataset_description.BIDSVersion         = 'unofficial extension';
+
+% % optional for dataset_description.json
+% general_info.dataset_description.License             = 'CC BY 4.0';
+% general_info.dataset_description.Authors             = 'Janna Protzak and Klaus Gramann';
+% general_info.dataset_description.Acknowledgements    = 'n/a';
+% general_info.dataset_description.Funding             = 'This Study was realized by funding from the Federal Ministry of Education and Research (BMBF)';
+% general_info.dataset_description.ReferencesAndLinks  = 'n/a';
+% general_info.dataset_description.DatasetDOI          = 'n/a';
+% 
+% % general information shared across modality specific json files 
+% general_info.InstitutionName                         = 'Technische Universitaet Berlin';
+% general_info.InstitutionalDepartmentName             = 'Junior research group FANS (Pedestrian Assistance System for Older Road Users), Department of Psychology and Ergonomics';
+% general_info.InstitutionAddress                      = 'Marchstr. 23, 10587, Berlin, Germany';
+% general_info.TaskDescription                         = 'Younger and older adults performed a visual discrimination task (button presses to peripheral presented LED flashes) during walking. Visual targets were either presented with or without preceding vibro-tactile cues';
+general_info.task                                    = bemobil_config.bids_task_label;  
+% 
+% % information about the EEG recording system 
+% %--------------------------------------------------------------------------
+% %--------------------------------------------------------------------------
+% coordinate system
+eeg_info.eeg.PowerLineFrequency                       = 50;
+% eeg_info.eeg.EEGReference                            = 'FCz'; 
+% 
+% % information about the motion recording system 
+% %--------------------------------------------------------------------------
+% %--------------------------------------------------------------------------
+% % data type and acquisition label
+% % motion_info.acq                                     = 'PhS';
+% % 
+% % % motion specific fields in json
[…]

### 2021-09-23 processing/fastReach_bemobil_config.m

@@ -1,0 +1,217 @@
+clear bemobil_config
+
+%% General Setup
+bemobil_config.study_folder = '/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/'; %(NEEDS to have a filesep at the end, sorry!) 
+bemobil_config.filename_prefix = 'sub-';
+
+% foldernames (NEED to have a filesep at the end, sorry!) 
+bemobil_config.source_data_folder = '0_source-data/';
+bemobil_config.bids_data_folder = '1_BIDS-data/'; 
+bemobil_config.raw_EEGLAB_data_folder = '2_raw-EEGLAB/';
+bemobil_config.EEG_preprocessing_data_folder = '3_EEG-preprocessing/';
+bemobil_config.spatial_filters_folder = '4_spatial-filters/';
+bemobil_config.spatial_filters_folder_AMICA = '4-1_AMICA/';
+bemobil_config.single_subject_analysis_folder = '5_single-subject-EEG-analysis/';
+bemobil_config.motion_analysis_folder = '6_single-subject-motion-analysis/';
+
+% filenames
+bemobil_config.merged_filename = '_EEG.set';
+bemobil_config.basic_prepared_filename = 'basic_prepared.set';
+bemobil_config.preprocessed_filename = 'preprocessed.set';
+bemobil_config.filtered_filename = 'filtered.set';
+bemobil_config.amica_filename_output = 'AMICA.set';
+bemobil_config.dipfitted_filename = 'dipfitted.set';
+bemobil_config.preprocessed_and_ICA_filename = 'preprocessed_and_ICA.set';
+bemobil_config.single_subject_cleaned_ICA_filename = 'cleaned_with_ICA.set';
+bemobil_config.processed_motion_filename = 'motion_processed.set';
+
+%% Import Settings 
+% Fields below do not have default values or important for data reading
+% configuration is necessary
+%--------------------------------------------------------------------------
+
+% scripts will find files containing respective string, sort them incrementally 
+% and output files merged within and between sessions. 
+% _ or - character not recommended (incompatible with bids, potentially breaks parsing)
+bemobil_config.session_names = {''};
+
+% name of the task that is common to all sessions - only relevant for data in BIDS
+bemobil_config.bids_task_label = 'fastReach'; 
+
+% a keyword contained in the EEG stream that is unique to EEG. 
+% Non-continuous (marker) streams will not be mixed up even if it contains this string.
[…]

### 2021-09-23 processing/fastReach_bemobil_process_all_EEG_data.m

@@ -1,0 +1,94 @@
+
+% initialize EEGLAB 
+if ~exist('ALLEEG','var')
+	eeglab;
+end
+
+% initialize FieldTrip 
+ft_defaults
+
+% load configuration 
+fastReach_bemobil_config;
+
+% if present, load additional metadata saved in example_bemobil_bids_metadata.m
+% this is primarily for enhancing documentation and data sharing 
+% (no influence on processing if this step is skipped)
+fastReach_bemobil_bids_metadata; 
+
+% enter all subjects to process here (you can split it up in more MATLAB instances if you have more CPU power and RAM)
+subjects = [1]; 
+
+% set to 1 if all files should be computed, independently of whether they are present on disk or not
+force_recompute = 0;
+
+%% Import 
+% (no looping over subjects - enter the whole array of IDs)
+
+% step 1 : convert .xdf to bids
+% bemobil_xdf2bids(bemobil_config, subjects) for minimal use 
+bemobil_xdf2bids(bemobil_config, subjects);
+
+% bemobil_xdf2bids(bemobil_config, subjects, 'general_metadata', general_info, 'motion_metadata', motion_info, 'eeg_metadata',...
+%     eeg_info, 'participant_metadata', subject_info)
+
+% step 2 : convert bids to .set
+bemobil_bids2set(bemobil_config, subjects);
+
+
+%% processing loop
+
+for subject = subjects
+    
+    %% prepare filepaths and check if already done
+    
+	disp(['Subject #' num2str(subject)]);
+    
+	STUDY = []; CURRENTSTUDY = 0; ALLEEG = [];  CURRENTSET=[]; EEG=[]; EEG_interp_avref = []; EEG_single_subject_final = [];
+	
+	input_filepath = [bemobil_config.study_folder bemobil_config.raw_EEGLAB_data_folder bemobil_config.filename_prefix num2str(subject)];
+	output_filepath = [bemobil_config.study_folder bemobil_config.single_subject_analysis_folder bemobil_config.filename_prefix num2str(subject)];
+	
+	try
+		% load completely processed file
+		EEG_single_subject_final = pop_loadset('filename', [ bemobil_config.filename_prefix num2str(subject)...
+			'_' bemobil_config.single_subject_cleaned_ICA_filename], 'filepath', output_filepath);
+    catch
+        disp('...failed. Computing now.')
+	end
+	
+	if ~force_recompute && exist('EEG_single_subject_final','var') && ~isempty(EEG_single_subject_final)
[…]

### 2021-09-23 processing/fastReach_bemobil_process_all_motion_data.m

@@ -1,0 +1,73 @@
+
+% initialize EEGLAB 
+if ~exist('ALLEEG','var')
+	eeglab;
+end
+
+% load configuration 
+fastReach_bemobil_config;
+
+% if present, load additional metadata saved in example_bemobil_bids_metadata.m
+% this is primarily for enhancing documentation and data sharing 
+% (no influence on processing if this step is skipped)
+fastReach_bemobil_bids_metadata; 
+
+% enter all subjects to process here (you can split it up in more MATLAB instances if you have more CPU power and RAM)
+subjects = [1]; 
+
+% set to 1 if all files should be computed, independently of whether they are present on disk or not
+force_recompute = 0;
+
+%% processing loop
+
+for subject = subjects
+    
+    %% prepare filepaths and check if already done
+    
+    disp(['Subject #' num2str(subject)]);
+    
+    STUDY = []; ALLEEG = [];  CURRENTSET=[]; motion_processed = [];
+    
+    input_filepath = [bemobil_config.study_folder bemobil_config.raw_EEGLAB_data_folder bemobil_config.filename_prefix num2str(subject)];
+    output_filepath = [bemobil_config.study_folder bemobil_config.motion_analysis_folder bemobil_config.filename_prefix num2str(subject)];
+    
+    try
+        % load completely processed file
+        motion_processed = pop_loadset('filename', [ bemobil_config.filename_prefix num2str(subject)...
+            '_' bemobil_config.processed_motion_filename], 'filepath', output_filepath);
+    catch
+        disp('...failed.')
+    end
+    
+    if ~force_recompute && exist('motion_processed','var') && ~isempty(motion_processed)
+        clear motion_processed
+        disp('Subject is completely preprocessed already.')
+        continue
+    end
+    
+    %% load data that is provided by the BIDS importer
+    
+    % make sure the data is stored in double precision, large datafiles are supported, and no memory mapped objects are
+    % used but data is processed locally
+    try
+        pop_editoptions( 'option_saveversion6', 0, 'option_single', 0, 'option_memmapdata', 0);
+    catch
+        warning('Could NOT edit EEGLAB memory options!!');
+    end
+    
+    % load files that were created from xdf to BIDS to EEGLAB
+    EEG = pop_loadset('filename',[ bemobil_config.filename_prefix num2str(subject) '__MOTION.set'],'filepath',input_filepath);
+    
[…]

### 2021-09-23 processing/plot_vel_emg_envelope.m

@@ -1,0 +1,105 @@
+motion = pop_loadset('filename','sub-1_motion_processed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/6_single-subject-motion-analysis/sub-1/');
+EEG = pop_loadset('filename','sub-1_preprocessed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/3_EEG-preprocessing/sub-1/');
+physio = pop_loadset('filename','sub-1__PHYSIO.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/2_raw-EEGLAB/sub-1/');
+
+ignore_early_samples = 250;
+events = EEG.event;
+
+%% signal processing
+sp_move = extract_motion_features(events,motion,'trial_start',[0, 10],ignore_early_samples);
+
+sp_eeg = extract_EEG_features(EEG,sp_move.onset_sample,'trial_start',{[0, 10],[-3, -2]});
+
+sp_physio = extract_physio_features(events,physio,'trial_start',[0, 10],ignore_early_samples);
+
+%% plot movement onset detection using EMG envelope and Velocity data
+
+sp_physio.emg.data = sp_physio.emg.data';
+sp_physio.emg.ylower = sp_physio.emg.ylower';
+
+%% align data using velocity onsets
+
+for i = 1:size(sp_move.mag,2)
+    interval = sp_move.onset_sample(i)-ceil(EEG.srate/4):sp_move.onset_sample(i)+(EEG.srate/4);
+    vel(:,i) = sp_move.mag(interval,i) - sp_move.mag(interval(1),i);
+    emg(:,i) = sp_physio.emg.data(interval,i) - sp_physio.emg.data(interval(1),i);
+    env(:,i) = sp_physio.emg.ylower(interval,i) - sp_physio.emg.ylower(interval(1),i);
+end
+
+vel = vel';
+env = env';
+emg = emg';
+
+%% plot single trial
+
+normal;
+figure('visible','on', 'Renderer', 'painters', 'Position', [10 10 600 500]);
+
+plot(vel(1,:),'LineWidth',2); hold on;
+plot(env(1,:),'LineWidth',2);
+plot(emg(1,:),'LineWidth',2);
+legend('vel','emg envelope','emg');
+
+ylim([-.2,.2]);
+
+% add markers
[…]

### 2022-01-14 processing/data_exploration.m
updates for Kristian and Konstantin

@@ -1,3 +1,16 @@
+
+% check in with Sein Yeung or Marius Klug to get these two functions below
+% to work to read in the .xdf datasets recoreded
+
+% bemobil_xdf2bids
+% bemobil_bids2set
+
+% you'll find them in this repo on the dev branch
+% https://github.com/BeMoBIL/bemobil-pipeline/tree/dev/import
+
+
+%%
+
 EEG = pop_loadset('filename','sub-1_preprocessed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/3_EEG-preprocessing/sub-1/');
 motion = pop_loadset('filename','sub-1_motion_processed.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/6_single-subject-motion-analysis/sub-1/');
 physio = pop_loadset('filename','sub-1__PHYSIO.set','filepath','/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/2_raw-EEGLAB/sub-1/');
@@ -23,7 +36,7 @@ fR.p.feature.nr_windowed_means = 10;
 
 %% signal processing EEG and Physio data
 
-EEG = pop_eegfiltnew(EEG, .1, 20);
+EEG_filt = pop_eegfiltnew(EEG, .1, 20);
 physio = pop_eegfiltnew(physio, .1, 10);
 
 % motion: overwrite random channel with magnitude, only need that!
@@ -65,7 +78,9 @@ end
 
 %%% EEG epochs
 tmp = pop_epoch(EEG, fR.p.movement_class(1), fR.p.movement_class{2});
-fR.move.eeg.data = tmp.data;
+fR.move.eeg.raw_data = tmp.data;
+tmp = pop_epoch(EEG_filt, fR.p.movement_class(1), fR.p.movement_class{2});
+fR.move.eeg.filt_data = tmp.data;
 
 %% IDLE Class
 
@@ -80,7 +95,12 @@ for i = 1:size(tmp.data,3)
 end
 
 tmp = pop_epoch(EEG, fR.p.idle_class(1), fR.p.idle_class{2});
-fR.idle.eeg.feat = tmp.data - tmp.data(:,1,:); % baseline correction
+fR.idle.eeg.raw_data = tmp.data; % - tmp.data(:,1,:); % baseline correction
[…]

### 2022-01-14 processing/extract_EEG_features.m

@@ -2,10 +2,6 @@ function [sp] = extract_EEG_features(EEG, onsets, event_str, epoch_classes)
 %EXTRACT_EEG_FEATURES Summary of this function goes here
 %   Detailed explanation goes her
 
-%% signal processing
-
-EEG = pop_eegfiltnew(EEG, .1, 20);
-
 %% select sample in pre-movement class
 
 EEG_mov = pop_epoch( EEG, {event_str}, epoch_classes{1});

### 2022-01-14 processing/fastReach_bemobil_config.m

@@ -1,8 +1,11 @@
 clear bemobil_config
 
 %% General Setup
-bemobil_config.study_folder = '/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/'; %(NEEDS to have a filesep at the end, sorry!) 
-bemobil_config.filename_prefix = 'sub-';
+% bemobil_config.study_folder = '/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/data/'; %(NEEDS to have a filesep at the end, sorry!) 
+% bemobil_config.filename_prefix = 'sub-';
+
+bemobil_config.study_folder = '/Users/lukasgehrke/Documents/publications/2021-CHI-fastReach/classifier-poc/pre-study/data'; %(NEEDS to have a filesep at the end, sorry!) 
+bemobil_config.filename_prefix = 'p';
 
 % foldernames (NEED to have a filesep at the end, sorry!) 
 bemobil_config.source_data_folder = '0_source-data/';

### 2022-02-07 input/1_teaser_figure.tex
add new latex project template

% This is what most people (90+%) will see from your paper!

% - [ ]  use hemingwayapp.com (+SEO if hardcore) to make figure captions super precise
% - [ ]  are all the axes labels readable, font size 10 or up
% - [ ]  check for missing information: give someone the figure with captions to someone neutral to check if everything is clear from the figure alone
% - [ ]  for digital version: what is alt-text of figure, i.e. when hovering with the mouse over the figure what text appears next to the cursor
% - [ ]  for accessibility: write figure caption for the blind
%     - [ ]  explain what the figure shows
%     - [ ]  refactor using hemingwayapp.com

\begin{teaserfigure}
  \includegraphics[width=\textwidth]{tex_sample_files/sampleteaser.pdf}
  \caption{Seattle Mariners at Spring Training, 2010.}
  \Description{Enjoying the baseball game from the third-base
  seats. Ichiro Suzuki preparing to bat.}
  \label{fig:teaser}
\end{teaserfigure}

### 2022-02-07 input/2_introduction.tex
add new latex project template

\section{Introduction}
% Lukas note:
% I like to use SCQA (Situation, Complication, Question, Answer) 'system' to write the introduction, see one of the two links below
% https://analytic-storytelling.com/scqa-what-is-it-how-does-it-work-and-how-can-it-help-me/
% https://corporatefinanceinstitute.com/resources/careers/how-to-job-guides/scqa/
% https://karrierebibel.de/scqa-methode/

% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.

% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short

\subsection{Main contribution: Our proposed solution}
% 4. (long) Anwser with teaser image
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% Now extend into a longer paragraph with a teaser image explaining the system in the following subsection

% for citations use use "~\cite{}" instead of " \cite{}" → no white space, copy bibtex items to cite into input/bibliography.bib
% example:
some text~\cite{Gehrke2019-og}

\input{input/1_teaser_figure}

### 2022-02-07 input/4_user_study_and_methods.tex
add new latex project template

% Gender-inclusive: Relevant throughout the document but frequently occuring here:
% - [ ]  write gender-inclusive: do not use he or she but use they instead: "the participant was asking..., he liked ... → they liked (in singular)
% check for the following:
% - [ ]  Have you used “man” or “men” or words containing them to refer to people who may not be men?
% - [ ]  Have you used “he,” “him,” “his,” or “himself” to refer to people who may not be men?
% - [ ]  If you have mentioned someone’s sex or gender, was it necessary to do so?
% - [ ]  Do you use any occupational (or other) stereotypes?
% - [ ]  Do you provide the same kinds of information and descriptions when writing about people of different genders?

\section{User Study}

\section{Methods}

### 2022-09-02 app/fastReach/fastReach.py
add intentional binding task LSL markers

@@ -58,7 +58,8 @@ class fastReach:
             do_reg = False
             self.eeg = Classifier('eeg_classifier', (buffer_feat_comp_size_samples/windowed_mean_size_samples)-1,
                 model_path_eeg, "eeg", target_class, chans, threshold, windowed_mean_size_samples, buffer_feat_comp_size_samples, do_reg)
-            self.eeg.start()
+            
+            # self.eeg.start()
 
             model_path_motion = '/Users/lukasgehrke/Documents/publications/2021-fastReach/data/study/eeglab2python/'+str(self.pID)+'/model_'+str(self.pID)+'_motion.sav'
             target_class = 1
@@ -179,9 +180,8 @@ class fastReach:
 
                     elif run_ib_task == True:
 
-                        # TODO now do the IB test here
-                        # after 2s after button press, play a sound and let participants judge whether they think this was longer than their reach to press lasted
-                        # # print the adjusted IB so I can then add it to the live phase
+                        # after Xs after button press, play a sound and let participants judge whether they think this was longer than their reach to press lasted
+                        # print the adjusted IB so I can then add it to the live phase
                         
                         if time.time() - reach_end_time < 2:
                             # self.instruct(self.instruction["start"]+'\n\n'+''.join(name))
@@ -190,9 +190,11 @@ class fastReach:
                         # some interval
                         if time.time() - reach_end_time > 2 and time.time() - reach_end_time < 7:
                             self.instruct(self.instruction["ib task wait"])
+                            self.lsl.send(self.markers["ib_task"]+"start",1)
 
                         if time.time() - reach_end_time > 4 and sound_played == False:
                             print(self.IB_DURATION)
+                            self.lsl.send(self.markers["ib_task"]+"tone",1)
                             self.play_sound(self.IB_DURATION)
                             sound_played = True
 
@@ -261,9 +263,11 @@ class fastReach:
 
                     if event.key == pg.K_UP:
                         self.IB_DURATION += self.IB_stepsize
+                        self.lsl.send(self.markers["ib_task"]+self.IB_stepsize,1)
                         ib_answered = True
                     if event.key == pg.K_DOWN:
                         self.IB_DURATION -= self.IB_stepsize              
+                        self.lsl.send(self.markers["ib_task"]+self.IB_stepsize,1)
                         ib_answered = True
 
[…]

### 2022-09-04 app/fastReach/fastReach.py
Added relay to discharge EMS intro 1k resistor in idle state

@@ -44,6 +44,7 @@ class fastReach:
         if ems_on == True:
             
             self.ems = serial.Serial(port=arduino_port, baudrate=9600, timeout=.1)
+            self.ems.write("r".encode('utf-8')) # discharge into resistor
             # self.ems_resetter = EMSResetter(self.ems, self.lsl, self.markers)
             # self.ems_resetter.start()
 
@@ -68,6 +69,7 @@ class fastReach:
             buffer_feat_comp_size_samples = 2
             windowed_mean_size_samples = 2
             do_reg = False
+
             # self.motion = Classifier('motion_classifier', 45, model_path_motion, "motion", target_class,
                 # chans, threshold, windowed_mean_size_samples, buffer_feat_comp_size_samples, do_reg)
             # self.motion.start()
@@ -235,7 +237,10 @@ class fastReach:
                 ems_duration = time.time() - ems_time
                 if ems_duration > .5 and ems_sent == True:
                     ems_sent = False
-                    self.ems.write("p".encode('utf-8'))
+                    
+                    self.ems.write("r".encode('utf-8'))
+                    self.ems.write("e".encode('utf-8'))
+
                     self.lsl.send(self.markers["ems off"],1)
 
                 if elapsed > trial_dur and ems_counter == trial_counter and trial_marker_sent == False:
@@ -247,7 +252,10 @@ class fastReach:
                         self.lsl.send(self.markers["deactivate EMS due to movement"],1)
 
                     if self.eeg.state == True and ems_sent == False and self.motion.state == False: # and self.ems_resetter.state == False:
-                        self.ems.write("p".encode('utf-8'))
+    
+                        self.ems.write("r".encode('utf-8'))
+                        self.ems.write("e".encode('utf-8'))
+
                         self.lsl.send(self.markers["ems on"],1)
                         ems_sent = True
                         ems_counter += 1
@@ -293,7 +301,10 @@ class fastReach:
 
             if ems_duration > .5 and ems_sent == True:
                 ems_sent = False
-                self.ems.write("p".encode('utf-8'))
+
+                self.ems.write("r".encode('utf-8'))
+                self.ems.write("e".encode('utf-8'))
+
                 self.lsl.send(self.markers["ems off"],1)
             
             if ems_duration > 4:
@@ -301,7 +312,10 @@ class fastReach:
         
             if mode == 'eeg':
                 if self.eeg.state == True and ems_sent == False and self.motion.state == False and ems_timeout == False: # and self.ems_resetter.state == False:
[…]

### 2022-09-04 input/2_introduction.tex
Updates from Overleaf

@@
 
+Hardware that allows physically moving a user's body becomes increasingly lightweight and seamless.
+Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
+
+Such technologies have primarily found application in rendering haptic sensations in virtual realities.
+However, with the potential to move the user's body independent of their intent, applications augmenting the user and its environment are feasible.
+
 % 2. What is the complication, challenge identified
@@
 
+if something happens for you in an interaction there is a cost to that:
+- impacts sense of agency which in turn makes user's engage less
+- impacts precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control
+
 % 3. Question
 % Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
+
+But how can we build a physical augmentation system that keeps user's engaged in the action, preserving sense of agency? In other words how can we design for users to experience natural augmentation?
 
@@
 
-\subsection{Main contribution: Our proposed solution}
+In this paper, we present an augmentation that maintains the user's sense of agency.
+Our system establishes a fast communication channel between the user's thoughts and a physical end effector, here EMS.
+The system controls the user's muscles through EMS at the time of the user's intent to interact, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
+In our user study we then investigated whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.
+
+\subsection{Human-Computer Collaboration for Natural Augmentation: Triggering Reaching Movements at the Time of Intention}
+
+\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thoght to EMS trigger on the Arm.}
+
 % 4. (long) Anwser with teaser image
@@
 % example:
-some text~\cite{Gehrke2019-og}
 

### 2022-09-04 input/4_user_study_and_methods.tex
Updates from Overleaf

@@
 
-\section{User Study}
+\section{User Study \& Methods}
 
-\section{Methods}
+With this study we wanted to check user's readiness for collaboration with a computer during preconscious interaction intent.
+
+We interviewed user's after they used a system that controlled their right forearm muscles to cue reaching movements.
+Our system uses electrical muscle stimulation (EMS) that is controlled by the output of a classifier based on the Electroencephalogram (EEG).
+
+\subsection{Participants}
+
+\todo{add actual number} Eight participants (\todo{add mean age and sd}) from our local institution were interviewed after they completed a writing task using only their right index finger.
+All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
+Participants received 12 Euro per hour.
+Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form approved by a fast track process of the local ethics committee.
+
+\subsection{Task}
+
+\todo{adapt this to the current task} Participants performed a 3D object selection task in VR.
+The interaction flow of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants moved their hands from the \textit{resting position} to the \textit{ready position}, to indicate they were ready to start the next trial; (2) participants waited for a new target to appear (the time of a new target spawning was randomized between 1-2 s); (3) then, the target (a cube) would appear in one of three possible positions (center, left, right), all equidistant from the participant's \textit{ready position}; (4) then, participants acquired the target by moving and touching the target with their index finger.
+(5) After a target was acquired, participants moved back to the \textit{resting position}.
+Here, they could take a break before the next trial.
+
+\missingfigure{add a infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
+
+\subsection{Interview \& Questionnaire}
+\todo{how where the results analyzed: sentiment analyses and statistics}
+
+We decided against asking questions at the end of each trial, to not break the user's immersion in the task.
+We aimed to learn about the feelings of cooperation between users and our system.
+The focus was not on correlating the accuracy of our system with the user's impression of it.
+This would have required to obtain labels for every trial, significantly disturbing the user in the flow of the interaction.
+
+\subsection{Recordings: Motion Capture and EEG}
+EEG was recorded using 64 active Ag/AgCl electrodes placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
+The electrode at position \todo{which one actually?}FP2 was detached from the cap and placed under the left eye to provide additional information about eye movements (EOG).
+Impedance was kept under 5k$\Omega$ where possible and the EEG was sampled at 250 Hz and amplified using BrainAmp DC amplifiers (Brainproducts GmbH, Gilching, Germany).
+Hand movements were sampled at 90 Hz.
+We repurposed the HTC Vive VR System for motion capture purposes.
+Through a Unity3D routine we streamed motion capture data of a HTC Vive Tracker Puck that was attached to the right hand, see \todo{add figure ref}.
+EEG, motion capture and an experiment marker stream were recorded and synchronized using labstreaminglayer \footnote{https://github.com/sccn/labstreaminglayer}.
+
+\missingfigure{show participant with measurement system and description next to each element}
+
+\subsubsection{Reproducing Results and Data Availability}
+\todo{Anonymize for submission} Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation \todo{add the correct links when uploaded} (OSF)\footnote{https://osf.io/x7hnm/}.
+BIDS formatted data is hosted on openneuro \cite{ds003846:1.0.0}.
+
+\subsection{Classifying EEG Signals to Control Muscle Stimulation}
+
+% event labels
+
+\subsection{Labelling Pre-movement Segments}
+For classification of single-trial ERPs, a regularized linear discriminant analysis (LDA) classifier was trained per participant with all idle data segments constituting the first class and pre-move data segments making up the second class.
+
+We developed a movement onset detector to label pre-movement segments of the hand movement time series.
+For this, the raw hand motion data was filtered with a 6Hz low-pass filter and re-sampled to match the EEG sample rate using \todo{reference bemobil pipeline}.
+Subsequently the first derivative was computed and velocity was extracted.
+
+Using the LDA implementation in the python toolbox scikit-learn, the classifier was trained on windowed means as features.
+First, EEG data were re-sampled to 100 Hz and band-pass filtered from 0.1 to 15 Hz.
+\todo{Refer to schultze-kraft paper here and adapt the following sentences accordingly, this is just copied from jne paper}Average amplitudes of 20 channels in 20 sequential 50 ms time windows between 0 and 400 ms after the cube was tapped were extracted as the windowed means feature vectors.
+A mean baseline taken in the -50 to 0 ms window was subtracted in order to compensate for event classes, match and mismatch, occurring at different stages of the ongoing movement.
+
+We then proceeded to a 20 best channels selection to reduce the computation time and increase the relevancy of the data.
+We first calculated the mean potential for each channel averaged on all the intention phases, then the selection was done by calculating the potential difference between the mean of the 25 first samples and the 25 last samples in each channel.
+The best ones were considered as the 20 channels presenting the largest difference between their first and last means.
+Once we extracted all the epochs and selected our best channels, we finally proceeded to a baseline correction of the signal.
+Indeed, the EEG potential may shift away from its initial position along the recording without any biophysical change occurring.
+
+After this processing, we extracted the mean each 25 samples during the ~\textit{idle} and ~\textit{intention} phases (1s duration at 250Hz) on the selected channel.
+Thus we obtained 10 features per channel on each trial for a total of 200 features.
+
+For robust performance estimation, a 5 x 5 nested cross-validation was used to calculate the shrinkage regularization parameter and assess the classifiers performance.
+
+\subsection{Experiment Design}
+
+% training and live phase with follow up interview

### 2022-09-04 input/3_related_work.tex

\section{Related Work}

\subsection{Disrupted Sense of Agency in Human Augmentation}

This defines the concept of \textbf{natural augmentation} which pairs the idea of wearable designs with reduced or non-existent physical and cognitive burden when using the interface. This notion was illustrated by~\citet{Eden2021-og} which defined natural/true degree of freedom augmentation for movement as a usable additional interface that would be controlled by the user without sacrificing other movements. For example, the control of a robotic arm via a remote control is sacrificing the degree of freedom of our hands. However, a robotic arm that is controlled as if it was an additional identical separate limb would not impair our initial range of potential movements, thus it would be a true augmentation. We could extend this definition saying that any true augmentation would be an enhancement in the way to interact with the environment without impairing any other physical or mental function.

\subsection{Single-trial Classification of the Intent to Interact}

Our idea to overcome the disruption in agency is to use EEG signal reflecting (inter)action intent.

One possible usage scenario of BCIs can be realized by using the Readiness Potential (RP) to de- tect motion intent and trigger actions accordingly. The is a type of potential that has been known for decades (Kornhuber & Deecke, 1965), with research continuously providing evidence of its role in volitional movement preparation (Libet et al., 1993). While it is similar to an event-related potential (ERP), it does not get generated after a stimulus but before the conscious movement starts (Smulders et al., 2012). RP comprises two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly af- ter. The second stage manifests contralaterally in the primary motor cortex (Shibasaki & Hallett, 2006).
A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preced- ing movements that people cannot explicitly suppress (Schultze-Kraft et al., 2021). The inability to suppress the RP indicates that it can always be present in the brain waves before motion onset, and it can serve as a reliable predictor of movement.
Having such a predictor for movement can open up new opportunities for the design of BCIs. Shih et al. (2012) defines BCIs as computer-based systems that acquire brain signals, analyze them, and translate them into commands relayed to an output device to carry out the desired action. The purpose of these systems is to detect and quantify the features of brain signals that indicate the user’s intention and translate these features in real-time into device commands that accomplish the user’s intent (Shih et al., 2012). Therefore, using the RP as a feature for prediction motion intent seems a good design for a BCI. Bai et al. (2011) have been the first to successfully set up such an interface and confirmed its feasibility, and Gerrits (2017) has replicated the setup in a follow-up experiment.

\subsection{Levels of Consciousness in HCI}

Along with this idea,~\citet{Jain2020-og} highlighted that numerous types of HCIs for human augmentation remain unexplored. Following their thoughts, they proposed a finer classification of the traditionally defined levels of consciousness (i.e. unconscious, conscious and meta-conscious) by splitting the unconscious into subliminal and pre-conscious and the meta-conscious into meta-cognitive and meta-somatic (Figure \ref{fig:consciousness}). The subliminal level corresponds to what is outside the range of perception of the user and thus a subliminal stimulus do not increase the cognitive load (e.g effect of a vitamin pill intake). A pre-conscious stimulus is below the awareness of the user but it becomes perceivable if the user change its focus toward it (e.g. the side of the road while driving). The conscious level occupies what the user is currently focusing on and thus most of the cognitive load comes from it. The meta-cognitive process is the awareness of the current conscious state (e.g. reflecting on the speed when walking). Finally, the meta-somatic pathway corresponds to the automatic response of the body to external stimulus, thus without intervention of the user (e.g. chills in response to a cold breeze).
This extended classification puts the spot on the new aspects on which a HCI can be designed and how they could use these pathways as new alternatives to contribute to natural human augmentation. Indeed, while numerous HCIs depend on the conscious state requiring the full user's awareness, the achievement of natural enhanced user experience may rely on the exploration of the interaction with other levels of consciousness.

### 2022-09-04 input/5_results.tex

\section{Results}

\subsection{Human-Computer Collaboration for Human Augmentation}
\missingfigure{A: Boxplot of questionnaire scores, are they significantly better than 0? Are they significantly better then after the training? -> frame the training the same as the experiment, B: Sentiment Word Clouds}

% system usability score

% sentiment analysis ttest the counts?

% Did they get faster? Calculate time between hand motion onset and button press. If yes, evidence that there was acceptance for the help/augmentation the system provided

\subsection{The BCI}

\missingfigure{!: grand average ERP at Cz comparing idle vs. movement intention with scalp map of voltage across all channels in the last [-100 0] interval of both idle and pre-move; B: What did the windows mean features look like. Boxplot with idle vs. pre-move and single observations}

% ttest of amplitudes in last time window

% classification accuracy

\missingfigure{confusion matrix!, AUC, ROC -> in discussion mention that we want to do that continuously and not only look at the times preceding hand movement but at all time points}

### 2022-09-05 input/4_user_study_and_methods.tex
Update 4_user_study_and_methods.tex

@@
 % - [ ]  Have you used “he,” “him,” “his,” or “himself” to refer to people who may not be men?
-% - [ ]  If you have mentioned someone’s sex or gender, was it necessary to do so?
+% - [ ]  If you have mentioned someone's sex or gender, was it necessary to do so?
 % - [ ]  Do you use any occupational (or other) stereotypes?
@@
 
-With this study we wanted to check user's readiness for collaboration with a computer during preconscious interaction intent.
+With this study we wanted to find out whether agency can be preseverd during physical actuation of the user's body when their own brain signals serve as the control signal.
+Specifically, our system leverages the brain signals known to reflect the volitional intent to interact, i.e.
+the EEG readiness potential, to trigger electrical muscle stimulation to initiate a reaching movement.
 
-We interviewed user's after they used a system that controlled their right forearm muscles to cue reaching movements.
-Our system uses electrical muscle stimulation (EMS) that is controlled by the output of a classifier based on the Electroencephalogram (EEG).
+To assess if agency was preserved we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitatitve interviews.
+Our interviews followed a phenomenelogical structure \todo{add reference from AH review paper mueller}.
 
-\subsection{Participants}
+\subsection{Apparatus}
+The experimental setup, depicted in Figure 2, comprised: (1) a wrist-mounted wearable HTC VIVE tracker leveraged for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
+To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
 
-\todo{add actual number} Eight participants (\todo{add mean age and sd}) from our local institution were interviewed after they completed a writing task using only their right index finger.
-All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
-Participants received 12 Euro per hour.
-Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form approved by a fast track process of the local ethics committee.
+\indent\textbf{1} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand.
+\indent\textbf{2} EEG data was recorded from 64 actively amplified electrodes using BrainAmp DC amplifiers from BrainProducts.
+Electrodes were placed according to the extended 10\% system \todo{ref}.
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10kOhm for all electrodes.
+EEG data was recorded with a sampling rate of 250 Hz.
+We used labstreaminglayer\footnote{Ref lsl} to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure using labstreaminglayer\footnote{Ref lsl}.
+
+\indent\textbf{3} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
+We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles.
+This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (\todo{EMS system}), which provides a maximum of 100mA.
+The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer.
+The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.
 
 \subsection{Task}
+\missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
+Participants performed a simple reaching task: resting their hand on the table, participants were instructed to press a specified key on a notebook in front of them.
+The interaction flow for one trial of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key.
+In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move".
+(4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone.
+(5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement.
+Following an inter-trial-interval the next trial started.
 
-\todo{adapt this to the current task} Participants performed a 3D object selection task in VR.
-The interaction flow of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants moved their hands from the \textit{resting position} to the \textit{ready position}, to indicate they were ready to start the next trial; (2) participants waited for a new target to appear (the time of a new target spawning was randomized between 1-2 s); (3) then, the target (a cube) would appear in one of three possible positions (center, left, right), all equidistant from the participant's \textit{ready position}; (4) then, participants acquired the target by moving and touching the target with their index finger.
-(5) After a target was acquired, participants moved back to the \textit{resting position}.
-Here, they could take a break before the next trial.
+\subsubsection{Experiment Design and Procedure}
+Participants completed two experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
 
-\missingfigure{add a infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
+In the second block, participants completed 80 trials of the same task and trial order as in the first block.
+Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background.
+After the target letter was shown to the participants and they were preparing to reach out and press the corresponding key, the BCI was live.
+Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.
+The second block was then followed by the mixed methods data collection of participants perceived agency.
 
-\subsection{Interview \& Questionnaire}
-\todo{how where the results analyzed: sentiment analyses and statistics}
+In total, the experiment consisted of four phases: (1) a setup phase; (2) a first experimental block: the training phase; (3) a second experiment block: the test phase; and (4) a test battery of questionnaire and interview to inquire about their experienced agency.
 
-We decided against asking questions at the end of each trial, to not break the user's immersion in the task.
-We aimed to learn about the feelings of cooperation between users and our system.
-The focus was not on correlating the accuracy of our system with the user's impression of it.
-This would have required to obtain labels for every trial, significantly disturbing the user in the flow of the interaction.
+\subsection{Brain-Computer Interface to Detect EEG Readiness Potentials }
 
-\subsection{Recordings: Motion Capture and EEG}
-EEG was recorded using 64 active Ag/AgCl electrodes placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
-The electrode at position \todo{which one actually?}FP2 was detached from the cap and placed under the left eye to provide additional information about eye movements (EOG).
-Impedance was kept under 5k$\Omega$ where possible and the EEG was sampled at 250 Hz and amplified using BrainAmp DC amplifiers (Brainproducts GmbH, Gilching, Germany).
-Hand movements were sampled at 90 Hz.
-We repurposed the HTC Vive VR System for motion capture purposes.
-Through a Unity3D routine we streamed motion capture data of a HTC Vive Tracker Puck that was attached to the right hand, see \todo{add figure ref}.
-EEG, motion capture and an experiment marker stream were recorded and synchronized using labstreaminglayer \footnote{https://github.com/sccn/labstreaminglayer}.
+\subsubsection{Detecting EEG Readiness Potentials}
+% Classifiying EEG Signals of the Intent to Interact
 
-\missingfigure{show participant with measurement system and description next to each element}
+To classify single-trial EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually.
+Using the data from the training phase of the experiment, we trained the classifier on 10 features from 20 select EEG channels for two classes: an idle class representing EEG background activity and a pre-movement class, representing the intent to interact.
 
-\subsubsection{Reproducing Results and Data Availability}
-\todo{Anonymize for submission} Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation \todo{add the correct links when uploaded} (OSF)\footnote{https://osf.io/x7hnm/}.
-BIDS formatted data is hosted on openneuro \cite{ds003846:1.0.0}.
+\missingfigure{some picture from the setup and prototype}
 
-\subsection{Classifying EEG Signals to Control Muscle Stimulation}
+\subsubsection{Best Channel Selection}
 
-% event labels
-
-\subsection{Labelling Pre-movement Segments}
-For classification of single-trial ERPs, a regularized linear discriminant analysis (LDA) classifier was trained per participant with all idle data segments constituting the first class and pre-move data segments making up the second class.
-
+\subsubsection{Idle and Pre-movement Data: Labelling Pre-movement Segments}
 We developed a movement onset detector to label pre-movement segments of the hand movement time series.
@@
 
-\subsection{Experiment Design}
+\subsubsection{EMS system and interface}
 
-% training and live phase with follow up interview
+% solid state relays hooked up to an arduino uno controlled using python
+
+\subsection{Participants}
+
+\todo{add actual number} Eight participants (\todo{add mean age and sd}) from our local institution participated in our study.
+All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
+Participants were compensated with 12 Euro per hour of study participation.
+Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form approved by a fast track process of the local ethics committee.
+
+\subsection{Interview \& Questionnaire}
+\todo{how where the results analyzed: sentiment analyses and statistics}
+
+We decided against asking questions at the end of each trial, to not break the user's immersion in the task.
+We aimed to learn about the feelings of cooperation between users and our system.
+The focus was not on correlating the accuracy of our system with the user's impression of it.
+This would have required to obtain labels for every trial, significantly disturbing the user in the flow of the interaction.
+
+% Questionnaire taken from Hornbaek
+
+\subsection{Recordings: Motion Capture and EEG}
+EEG was recorded using 64 active Ag/AgCl electrodes placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
+The electrode at position \todo{which one actually?}FP2 was detached from the cap and placed under the left eye to provide additional information about eye movements (EOG).
+Impedance was kept under 5k$\Omega$ where possible and the EEG was sampled at 250 Hz and amplified using BrainAmp DC amplifiers (Brainproducts GmbH, Gilching, Germany).
+Hand movements were sampled at 90 Hz.
+We repurposed the HTC Vive VR System for motion capture purposes.
+Through a Unity3D routine we streamed motion capture data of a HTC Vive Tracker Puck that was attached to the right hand, see \todo{add figure ref}.
+EEG, motion capture and an experiment marker stream were recorded and synchronized using labstreaminglayer \footnote{https://github.com/sccn/labstreaminglayer}.
+
+\missingfigure{show participant with measurement system and description next to each element}
+
+\subsubsection{Reproducing Results and Data Availability}
+\todo{Anonymize for submission} Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation \todo{add the correct links when uploaded} (OSF)\footnote{https://osf.io/x7hnm/}.
+BIDS formatted data is hosted on openneuro \cite{ds003846:1.0.0}.

### 2022-09-07 input/2_introduction.tex
methods section written

@@
 
-Hardware that allows physically moving a user's body becomes increasingly lightweight and seamless.
+The sense of agency refers to the experience of being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
+This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ubd} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
+
+When integrating with computers, the key design goal is to maintain this pre-reflective experience of agency.
+With hardware that allows physically moving a user's body becoming increasingly lightweight and seamless.
 Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
-
-Such technologies have primarily found application in rendering haptic sensations in virtual realities.
-However, with the potential to move the user's body independent of their intent, applications augmenting the user and its environment are feasible.
 

### 2022-09-07 input/4_user_study_and_methods.tex
methods section written

--- 2022-02-07 input/4_user_study_and_methods.tex
@@
-% Gender-inclusive: Relevant throughout the document but frequently occuring here:
-% - [ ]  write gender-inclusive: do not use he or she but use they instead: "the participant was asking..., he liked ... → they liked (in singular)
-% check for the following:
-% - [ ]  Have you used “man” or “men” or words containing them to refer to people who may not be men?
-% - [ ]  Have you used “he,” “him,” “his,” or “himself” to refer to people who may not be men?
-% - [ ]  If you have mentioned someone’s sex or gender, was it necessary to do so?
-% - [ ]  Do you use any occupational (or other) stereotypes?
-% - [ ]  Do you provide the same kinds of information and descriptions when writing about people of different genders?
+\section{Methods}
 
-\section{User Study}
+With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
+Our system leverages the brain signals known to reflect the volitional intent to interact, i.e.
+the EEG readiness potential, to trigger electrical muscle stimulation to cue a reaching movement.
 
-\section{Methods}
+To assess if agency was preserved, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews.
+Our interviews followed a phenomenological structure \todo{add reference from AH review paper mueller}.
+
+\subsection{Participants}
+
+\todo{add actual number} Eight participants (\todo{add mean age and sd}) from our local institution participated in our study.
+All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
+Participants were compensated with 12 Euro per hour of study participation.
+Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form approved by a fast track process of the local ethics committee.
+
+\subsection{Apparatus}
+The experimental setup, depicted in Figure~\ref{}, comprised: (1) a wrist-mounted wearable HTC VIVE tracker for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
+To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
+
+\indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz.
+Therefore, we ran a simple Unity3D scene and streamed the data via the labstreaminglatyer (LSL)\footnote{https://github.com/sccn/labstreaminglayer}.
+
+\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
+Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
+One electrode was placed under the left eye to provide additional information about eye movements (vEOG).
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes.
+EEG data was recorded with a sampling rate of 250 Hz.
+We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.
+
+\missingfigure{show participant with measurement system and description next to each element}
+
+\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
+We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles.
+This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus).
+The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer.
+The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.
+
+\subsection{Task}
+\missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
+Participants performed a simple reaching task: resting their hand on the table, participants were instructed to press a specified key on a notebook in front of them.
+For each trial, a key was randomly selected from one of two groups of letters: (1) 'a', 's', 'd' or (2) 'j', 'k', 'l'.
+The selected group was then changed with every trial.
+This was done to keep the task engaging while maintaining a comparable reaching movement between trials.
+
+The interaction flow for one trial of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key.
+In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move".
+(4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone.
+(5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement.
+Following an inter-trial-interval the next trial started.
+
+\subsubsection{Experiment Design and Procedure}
+Participants completed two experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
+
+In the second block, participants completed 80 trials of the same task and trial order as in the first block.
+Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background.
+After the target letter was shown to the participants and they were preparing to reach out and press the corresponding key, the BCI was live.
+Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.
+The second block was then followed by the mixed methods data collection of participants perceived agency.
+
+In summary, the experiment consisted of four phases: (1) a setup phase; (2) a first experimental block: the training phase; (3) a second experiment block: the test phase; and (4) a test battery of questionnaire and interview to inquire about their experienced agency.
+
+\subsection{Training of Classifiers from Training Phase Data}
+% Brain-Computer Interface to Detect EEG Readiness Potentials }
+% Classifiying EEG Signals of the Intent to Interact
+
+Before the test phase of our experiment, the data recorded during the training phase was processed in three consecutive steps: (1) we extracted movement onsets in each trial to construct three event classes: an \textit{idle} class representing resting background activity, a \textit{movement} class and a \textit{pre-movement} class, representing the intent to interact.
+Next, (2) a motion classifier was trained on \textit{movement} and \textit{idle} data segments and (3) an EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels.
+To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually.
+The EEG classifier was then used in the test phase to detect RPs and trigger EMS.
+The motion classifier was used to make the interaction more robust.
+When it detected the participants hand to be moving, it deactivated the control loop from RP to EMS.
+
+\missingfigure{some picture from the setup and prototype}
+
+\indent\textbf{(1) Labelling Pre-movement, Movement and Idle Data Segments.}
+In order to obtain training data segments for the \textit{pre-movement} class, a detection algorithm was applied on the hand velocity time series.
+From the training data, the raw hand motion data was filtered with a 6Hz low-pass filter and re-sampled to match the EEG sample rate using BeMoBIL Pipeline functions \textit{xdf2bids} and \textit{bids2set}\footnote{https://github.com/BeMoBIL/bemobil-pipeline}.
+Subsequently the first derivative was computed and velocity was extracted.
+Then, for each trial a two step thresholding process was applied to extract the exact time of movement onset: (1) the first time point when the velocity exceeded 70\% of the maximum velocity in the trial was selected, then (2) the signal was flipped at that time point and the time point where the flipped signal first fell below 10\% of the maximum in the remaining data was selected as the movement onset.
+
+The three event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets, \textit{movement} from 0 to 1000 ms succeeding the movement onsets and \textit{idle}, a one second data segment that fell exactly between two succeeding trials, i.e.
+time-locked to the inter-trial-interval.
+Here, participants where looking at a fixation cross on the screen with their hand resting on the table.
+
+\indent\textbf{(2) Training of Hand Motion Classifier.}
+An LDA was trained using the mean velocity per trial and movement and idle time segment.
+Hence, 60 values for each class were subjected to train the motion classifier.
+Since in the idle time segments almost no movement occurred we added some artificial noise.
+To this end, the idle features were multiplied by a factor of 3, chosen by simple trial and error.
+In the real-time feedback, this rendered the classifier less susceptible to false positives, i.e.
+detecting a movement when there was none.
+
+\indent\textbf{(3) Channel Selection \& Training of EEG Classifier.}
+For the EEG classifier, we followed the approach outlined by Schultze-Kraft et al.\cite{} with slight modifications.
+First, the 20 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure.
+For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment.
+The resulting value thus indicated how much the signal had changed in the course of the 1s segments.
+In line with the literature there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment.
+Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments.
+The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment.
+The ranks were joined and the 20 best channels were selected.
+In any case, channels C3, C4 and Cz were kept for every participant as these are most frequently reported in studies on the RP.
+
+To train the LDA EEG classifier, we extracted spatiotemporal features from the EEG; 10 windowed means were extracted for each segment and channel.
+These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment.
+The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.
+
+\subsubsection{Real-time Feedback}
+
+During the test phase, both classifiers were applied to the real-time data.
+With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier.
+For the EEG data streaming at 250 Hz, the EEG data of the last 1000ms was buffered for the selected best channels and 10 windowed means per channel were computed.
+Identical to the training data, these features were baseline corrected by subtracting the mean from the last 100 ms preceding the last second, i.e.
+-1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector.
+Hence, both classifier yielded one output value at each sample point.
+
+The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 certainty.
+Additionally, EMS was only triggered when the hand was \textit{not} moving.
+In each trial, as soon as the hand was in motion once after trial start, the EEG-EMS control was deactivated for the rest of the trial.
+This guaranteed that participants experienced EMS only during movement preparation.
+
+\subsection{Interview \& Questionnaire}
+
+To assess if agency was preserved, three measures were recorded during and after the test phase.
+
+\indent\textbf{(1) Intentional Binding.}
+At the end of each trial participants were tasked to compare the duration of their arm movement to a tone and indicate whether the tone was longer or shorter.
+We set the initial duration at 500 ms for each participant.
+When participants indicated the tone to be longer or shorter by pressing either the up or down key we added, or respectively subtracted, 100 ms from the tone duration.
+This new duration was then used for the next trial and so on.
+With 60 trials in the training phase, we hypothesized participants would oscillate 50-50 around a target duration after about 2/3 of the training trials.
+The final duration after all training trials was then used again as the start duration during the test phase.
+In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS.
+A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.
+
+\indent\textbf{(2) Questionnaire.}
+In line with~\cite{Hornbaek} we prompted participants with three statements and a 7-point Likert scale centered at 0.
+The first one "I felt like the device I was using was part of my body" was in relation to body ownership.
+The following two statements "It felt like I was in control of the movements during the task." and "What is the degree of control you felt?" prompted participants on their experienced agency.
+
+\indent\textbf{(1) Interview.}
+We interviewed user's about the experience with the EMS integration.
+We asked open how questions in a collaborative interview inquiring about the user's lived experience.
+After prompting user's to recall their experience we started the interview with: "how did it feel like?".
+If user's made reference themselves to the EMS device we followed up by inquiring: "why do you think it (the EMS trigger) happened?", "What do you think triggered it?".
+The goal was to learn about the source attribution user's made about the EMS trigger.
+Since we leveraged the RP, we hypothesized that user's would attribute, at least in part, the EMS trigger to themselves, i.e.
+their movement intention.
+Next, we pressed deeper into learning about the cooperative nature between user and system.
+We asked "Do you believe you did it alone?" and when they made reference to a specific situation where the EMS was triggered we followed up with "What did it feel like, did you had the feeling the system controlled you or did you cooperate?" and "What would have to be there for it to feel like a cooperation?".
+We closed the interview with two additional exploratory questions: "Did the system influence your performance?
+If yes, how and why?" as well as asking them about any application scenarios they have in mind.
+
+To not break the user's immersion by disturbing the flow of the interaction we decided against asking qualitative questions at the end of each and every trial.
+
+\subsection{Statistical Analyses}
+\todo{how where the results analyzed: sentiment analyses and statistics of IB data and questionnaire results}
+
+% ERP with significance mask for Cz? Assessing significance of classifier
+
+% IB Task: ttest of final ib time between with and without EMS -> # subjects input values
+% did they increase the IB task in more than 50% of the trials after EMS?
+
+% questionnaires: ttest against zero to report direction
+
+% interviews: anecdotal summary and sentiment analyses with two groups splitting questionnaires at 0
+
+% correlation with EEG classifier accuracy?
+
+% evaluation of EEG classifier: can report false negatives, how often was it missing to predict a movement onset?
+hand was moving without EMS being triggered occurred how often?
+Not possible to evlaute false positives as no ground truth available
+
+\subsection{Reproducing Results and Data Availability}
+Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}.
+BIDS formatted data is hosted on openneuro \cite{}.

### 2022-09-07 input/2_introduction.tex
Update 2_introduction.tex

@@
 
+Technological advances in hardware that allow physically moving a user's body have reignited the dreams of overcoming human limitations, recorvering lost abilities as well as simplifying skill acquisition.
+Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
+These technologies are becoming increasingly seamless to use.
+However, augmented user's frequently report dissociative experiences following augmented action.
+They do not experience agency.
+
 The sense of agency refers to the experience of being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
 This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ubd} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
+The key challenge in human \textit{physical augmentation} is to create an experience of pre-reflective agency, so user's feel as though they are in the "driving seat" once again.
 
-When integrating with computers, the key design goal is to maintain this pre-reflective experience of agency.
-With hardware that allows physically moving a user's body becoming increasingly lightweight and seamless.
-Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
+% What have others done to address this and what do we propose "instead"
 

### 2022-09-08 app/fastReach/fastReach.py
added code documentation

@@ -16,6 +16,17 @@ import matplotlib.pyplot as plt
 EMS_RESET_TIME = time.time()
 
 class fastReach:
+    """Experiment triggering electrical muscle stimulation EMS directly from the output of a brain-computer interface (BCI)
+    This class initializes the different components, starts an EEG and motion classifier in separate threads, and draws the experiment to the screen.
+
+    The experiment shows different letters on screen and waits for them to be pressed, then an intentional binding task follows.
+    This plays a sound and asks the participant to estimate the duration of the sound.
+
+    Args:
+        pID (integer): participant ID used to save data
+        ems_on (boolean): whether to run with or without electrical muscle stimulation (EMS)
+        arduino_port (string): port of connected arduino device
+    """
 
     def __init__(self, pID, ems_on, arduino_port) -> None:
         self.lsl = LSL('fastReach')
@@ -75,6 +86,11 @@ class fastReach:
             # self.motion.start()
 
     def init_screen(self, fullscreen):
+        """Initialize screen object using pygame
+
+        Args:
+            fullscreen (boolean): Whether to run in fullscreen mode
+        """
         info = pg.display.Info()
         if fullscreen == True:
             self.screen = pg.display.set_mode((info.current_w, info.current_h), pg.FULLSCREEN) 
@@ -87,17 +103,35 @@ class fastReach:
         self.font = pg.font.Font(None, 45)
 
     def play_sound(self, duration):
+        """Plays a sound using pygame
+
+        Args:
+            duration (integer): Duration in seconds that the sound is played
+        """
         self.sound.play(-1)
         pg.time.delay(duration)
         self.sound.stop()
 
     def instruct(self, text, col = (0,0,0)):
+        """Writes text to the screen object
+
+        Args:
+            text (string): The text to draw
+            col (tuple, optional): Color of the text. Defaults to (0,0,0).
+        """
 
         self.screen.fill((240, 240, 240))
         ptext.draw(text, topleft=(200, 200), color=col, fontsize=40)  # Recognizes newline characters.
         pg.display.flip()
  
     def start(self, trial_type, num_trials, ems_on):
+        """_summary_
+
+        Args:
+            trial_type (string): Can be "training" or "experiment", "training" is always without EMS
+            num_trials (integer): Number of trials to complete
+            ems_on (boolean): Whether to EMS is presented or not
+        """
         
         while True:
             for event in pg.event.get():
@@ -109,6 +143,14 @@ class fastReach:
                         self.app(trial_type, num_trials, ems_on, start)
 
[…]

### 2022-09-14 input/2_introduction.tex

@@
 \section{Introduction}
-% Lukas note:
-% I like to use SCQA (Situation, Complication, Question, Answer) 'system' to write the introduction, see one of the two links below
-% https://analytic-storytelling.com/scqa-what-is-it-how-does-it-work-and-how-can-it-help-me/
-% https://corporatefinanceinstitute.com/resources/careers/how-to-job-guides/scqa/
-% https://karrierebibel.de/scqa-methode/
-
 % 1. Describe the Current Situtation
 % Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
-
+% 2. What is the complication, challenge identified
+% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
 Technological advances in hardware that allow physically moving a user's body have reignited the dreams of overcoming human limitations, recorvering lost abilities as well as simplifying skill acquisition.
@@
 The key challenge in human \textit{physical augmentation} is to create an experience of pre-reflective agency, so user's feel as though they are in the "driving seat" once again.
+% if something happens for you in an interaction there is a cost to that:
+% - impacts sense of agency which in turn makes user's engage less
+% - impacts precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control
 
 % What have others done to address this and what do we propose "instead"
-
-% 2. What is the complication, challenge identified
-% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
-
-if something happens for you in an interaction there is a cost to that:
-- impacts sense of agency which in turn makes user's engage less
-- impacts precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control
+% summarize Kasahara's work
+This challenge has driven researcher
 
@@
 % Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
-
 But how can we build a physical augmentation system that keeps user's engaged in the action, preserving sense of agency? In other words how can we design for users to experience natural augmentation?
@@
 % keep this short
-
 In this paper, we present an augmentation that maintains the user's sense of agency.
@@
 
-\subsection{Human-Computer Collaboration for Natural Augmentation: Triggering Reaching Movements at the Time of Intention}
-
-\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thoght to EMS trigger on the Arm.}
-
+\subsection{Preserving Agency using Brain Signal of the Intent to Interact}
 % 4. (long) Anwser with teaser image
@@
 
-% for citations use use "~\cite{}" instead of " \cite{}" → no white space, copy bibtex items to cite into input/bibliography.bib
-% example:
+% not only physical integration but also other adpative interfaces, or control of mobile phone
 
+\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.}
 \input{input/1_teaser_figure}

### 2023-03-06 input/2_introduction.tex

@@
 % Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
-Technological advances in hardware that allow physically moving a user's body have reignited the dreams of overcoming human limitations, recorvering lost abilities as well as simplifying skill acquisition.
+Technological advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition.
 Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
@@
 The sense of agency refers to the experience of being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
-This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ubd} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
+This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
 The key challenge in human \textit{physical augmentation} is to create an experience of pre-reflective agency, so user's feel as though they are in the "driving seat" once again.
@@
 
-% What have others done to address this and what do we propose "instead"
-% summarize Kasahara's work
-This challenge has driven researcher
-
 % 3. Question
 % Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
-But how can we build a physical augmentation system that keeps user's engaged in the action, preserving sense of agency? In other words how can we design for users to experience natural augmentation?
+% What have others done to address this and what do we propose "instead"
+In physical augmentation, this experience of agency positively impacts performance.~\citet{Kasahara2019-sk} have shown that reaction times when tapping targets on a touchscreen can be increased by augmenting the user.
+Critically they found an optimal increase in reaction time, \textit{pre-emptive} gain, that preserves agency.
+At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.
+However, setting this gain factor is dependent on each specific task and requires fitting the parameters over and over again.
+A key question remains: How can we maintain agency during physical augmentation without training and over-fitting parameters?
+In other words, how can we create closed systems for users to experience natural augmentation?
 
@@
 % keep this short
-In this paper, we present an augmentation that maintains the user's sense of agency.
-Our system establishes a fast communication channel between the user's thoughts and a physical end effector, here EMS.
+In this paper, we present a system that establishes a fast communication channel between the user's brain signals and a physical end effector, here EMS.
 The system controls the user's muscles through EMS at the time of the user's intent to interact, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
-In our user study we then investigated whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.
+In our user study we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.
 
-\subsection{Preserving Agency using Brain Signal of the Intent to Interact}
+\subsection{Preserving Agency using Brain Signals of the Intent to (Inter)act}
 % 4. (long) Anwser with teaser image
 % Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
-% Now extend into a longer paragraph with a teaser image explaining the system in the following subsection
 
-% not only physical integration but also other adpative interfaces, or control of mobile phone
-
-\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.}
+%- not only physical integration but also other adpative interfaces, or control of mobile phone
+\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.
+<- better use this for the teaser image.
+Here better present the main results, i.e.
+sense of agency scores?}
 \input{input/1_teaser_figure}

### 2023-03-06 input/3_related_work.tex
Updates from Overleaf

@@
 
-\subsection{Disrupted Sense of Agency in Human Augmentation}
+We present a bodily integration experience moving the finger of a user after the user intends to do so, but before the user directs its muscles to move.
 
-This defines the concept of \textbf{natural augmentation} which pairs the idea of wearable designs with reduced or non-existent physical and cognitive burden when using the interface.
-This notion was illustrated by~\citet{Eden2021-og} which defined natural/true degree of freedom augmentation for movement as a usable additional interface that would be controlled by the user without sacrificing other movements.
-For example, the control of a robotic arm via a remote control is sacrificing the degree of freedom of our hands.
-However, a robotic arm that is controlled as if it was an additional identical separate limb would not impair our initial range of potential movements, thus it would be a true augmentation.
-We could extend this definition saying that any true augmentation would be an enhancement in the way to interact with the environment without impairing any other physical or mental function.
+\subsection{The Disruption in Sense of Agency in Augmented Humans}
 
-\subsection{Single-trial Classification of the Intent to Interact}
+\subsection{Assessing the Sense of Agency when Integrating with Computers}
+% how can we measure our outcome measure sense of agency?
+% - questionnaires
+% - intentional binding
+% - phenomenological interview
 
-Our idea to overcome the disruption in agency is to use EEG signal reflecting (inter)action intent.
-
-One possible usage scenario of BCIs can be realized by using the Readiness Potential (RP) to de- tect motion intent and trigger actions accordingly.
-The is a type of potential that has been known for decades (Kornhuber & Deecke, 1965), with research continuously providing evidence of its role in volitional movement preparation (Libet et al., 1993).
-While it is similar to an event-related potential (ERP), it does not get generated after a stimulus but before the conscious movement starts (Smulders et al., 2012).
-RP comprises two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
-The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly af- ter.
-The second stage manifests contralaterally in the primary motor cortex (Shibasaki & Hallett, 2006).
-A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preced- ing movements that people cannot explicitly suppress (Schultze-Kraft et al., 2021).
-The inability to suppress the RP indicates that it can always be present in the brain waves before motion onset, and it can serve as a reliable predictor of movement.
-Having such a predictor for movement can open up new opportunities for the design of BCIs.
-Shih et al.
-(2012) defines BCIs as computer-based systems that acquire brain signals, analyze them, and translate them into commands relayed to an output device to carry out the desired action.
-The purpose of these systems is to detect and quantify the features of brain signals that indicate the user’s intention and translate these features in real-time into device commands that accomplish the user’s intent (Shih et al., 2012).
-Therefore, using the RP as a feature for prediction motion intent seems a good design for a BCI.
-Bai et al.
-(2011) have been the first to successfully set up such an interface and confirmed its feasibility, and Gerrits (2017) has replicated the setup in a follow-up experiment.
-
-\subsection{Levels of Consciousness in HCI}
-
-Along with this idea,~\citet{Jain2020-og} highlighted that numerous types of HCIs for human augmentation remain unexplored.
-Following their thoughts, they proposed a finer classification of the traditionally defined levels of consciousness (i.e.
-unconscious, conscious and meta-conscious) by splitting the unconscious into subliminal and pre-conscious and the meta-conscious into meta-cognitive and meta-somatic (Figure \ref{fig:consciousness}).
-The subliminal level corresponds to what is outside the range of perception of the user and thus a subliminal stimulus do not increase the cognitive load (e.g effect of a vitamin pill intake).
-A pre-conscious stimulus is below the awareness of the user but it becomes perceivable if the user change its focus toward it (e.g.
-the side of the road while driving).
-The conscious level occupies what the user is currently focusing on and thus most of the cognitive load comes from it.
-The meta-cognitive process is the awareness of the current conscious state (e.g.
-reflecting on the speed when walking).
-Finally, the meta-somatic pathway corresponds to the automatic response of the body to external stimulus, thus without intervention of the user (e.g.
-chills in response to a cold breeze).
-This extended classification puts the spot on the new aspects on which a HCI can be designed and how they could use these pathways as new alternatives to contribute to natural human augmentation.
-Indeed, while numerous HCIs depend on the conscious state requiring the full user's awareness, the achievement of natural enhanced user experience may rely on the exploration of the interaction with other levels of consciousness.
+\subsection{Brain-Computer Interfaces for Intent Prediction}
+% Single-trial Classification of the Intent to Interact
+Our idea to overcome the disruption in agency is to use EEG signals reflecting (inter)action intent.
+One such signal is the Readiness Potential (RP), or \textit{lateralized readiness potential}.
+It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
+It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand.
+Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
+However, evidence abounds for its role in action preparation.
+An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
+The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after.
+The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}.
+A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}.
+In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible.
+After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for.
+They found participants \textit{not} to be able to suppress their RP.
+This inability to suppress the RP renders it a reliable feature for classification.
+While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.

### 2023-03-06 input/4_user_study_and_methods.tex
Updates from Overleaf

@@
 With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
-Our system leverages the brain signals known to reflect the volitional intent to interact, i.e.
+To this end, we build a system that leverages the brain signals known to reflect the volitional intent to interact, i.e.
 the EEG readiness potential, to trigger electrical muscle stimulation to cue a reaching movement.
@@
 Our interviews followed a phenomenological structure \todo{add reference from AH review paper mueller}.
+We t
 
@@
 
-\todo{add actual number} Eight participants (\todo{add mean age and sd}) from our local institution participated in our study.
+\todo{add actual number} Ten participants (\todo{add mean age and sd}) from our local institution participated in our study.
 All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
 Participants were compensated with 12 Euro per hour of study participation.
-Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form approved by a fast track process of the local ethics committee.
+Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form.
 
@@
 \subsubsection{Experiment Design and Procedure}
-Participants completed two experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
+Participants completed three experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
 
-In the second block, participants completed 80 trials of the same task and trial order as in the first block.
+%The condition of the second and third block, see below, were alternated between participants.
+
+In a second block, participants completed 80 trials of the same task and trial order as in the first block.
 Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background.
@@
 Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.
-The second block was then followed by the mixed methods data collection of participants perceived agency.
+
+Both experimental blocks two and three were followed by a mixed methods data collection of participants' perceived agency.
 
@@
 -1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector.
-Hence, both classifier yielded one output value at each sample point.
+Hence, both classifiers yielded one output value at each sample point.
 
-The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 certainty.
+The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability.
 Additionally, EMS was only triggered when the hand was \textit{not} moving.
@@
 In line with~\cite{Hornbaek} we prompted participants with three statements and a 7-point Likert scale centered at 0.
-The first one "I felt like the device I was using was part of my body" was in relation to body ownership.
-The following two statements "It felt like I was in control of the movements during the task." and "What is the degree of control you felt?" prompted participants on their experienced agency.
+The first one ``I felt like the device I was using was part of my body'' was in relation to body ownership.
+The following two statements ``It felt like I was in control of the movements during the task.'' and ``What is the degree of control you felt?'' prompted participants on their experienced agency.
 
-\indent\textbf{(1) Interview.}
+\indent\textbf{(3) Interview.}
 We interviewed user's about the experience with the EMS integration.

### 2023-03-06 signal_processing/pi_bemobil_config.m
changes to run on lab computers

@@ -2,7 +2,8 @@ clear bemobil_config
 
 %% General Setup
 
-path = '/Users/lukasgehrke/Documents/publications/2021-fastReach';
+path = 'P:\Lukas_Gehrke\2021-fastReach';
+%'/Users/lukasgehrke/Documents/publications/2021-fastReach';
 
 bemobil_config.study_folder = [path filesep 'data' filesep]; %(NEEDS to have a filesep at the end, sorry!) 
 bemobil_config.filename_prefix = 'sub-';

### 2023-03-06 signal_processing/pi_main.m

@@ -2,11 +2,13 @@
 pID = 1;
 
 %% config
-current_sys = "mac";
-eeglab_ver(current_sys);
+current_sys = "win";
+eeglab
+%eeglab_ver(current_sys);
 
-addpath('/Users/lukasgehrke/Documents/publications/2021-fastReach/signal_processing');
-addpath('/Users/lukasgehrke/Documents/code.nosync/signal-processing-motor-intent');
+addpath(genpath('D:\Lukas\signal-processing-motor-intent'));
+%addpath('/Users/lukasgehrke/Documents/publications/2021-fastReach/signal_processing');
+%addpath('/Users/lukasgehrke/Documents/code.nosync/signal-processing-motor-intent');
 
 pi_bemobil_config;
 
@@ -58,8 +60,14 @@ idle_erp = pop_epoch(idle_data, {'idle_start'}, [-1.1, 0]);
 
 %% reject noisy epochs
 
-[idle_erp, idle_noisy_epochs] = pop_autorej(idle_erp, 'nogui','on','eegplot','off');
-[pre_move_erp, pre_move_noisy_epochs] = pop_autorej(pre_move_erp, 'nogui','on','eegplot','off');
+thresh = max(abs(idle_erp.data(:))) / 2;
+idle_erp = pop_eegthresh(idle_erp,1,[1:size(idle_erp.data,1)],-thresh,thresh,idle_erp.xmin,idle_erp.xmax,0,0);
+idle_erp = pop_rejepoch(idle_erp, idle_erp.reject.rejthresh,0); % actually reject high prob epochs
+
+%[pre_move_erp, pre_move_noisy_epochs] = pop_autorej(pre_move_erp, 'nogui','on','eegplot','off');
+thresh = max(abs(pre_move_erp.data(:))) / 2;
+pre_move_erp = pop_eegthresh(pre_move_erp,1,[1:size(pre_move_erp.data,1)],-thresh,thresh,pre_move_erp.xmin,pre_move_erp.xmax,0,0);
+pre_move_erp = pop_rejepoch(pre_move_erp, pre_move_erp.reject.rejthresh,0); % actually reject high prob epochs
 
 %% select best channels
 
@@ -82,8 +90,10 @@ if ~exist(path, 'dir')
     mkdir(path);
 end
 
-idle = idle_erp.data;
-pre_move = pre_move_erp.data;
+% make same size
[…]

### 2023-03-07 input/2_introduction.tex

@@
 These technologies are becoming increasingly seamless to use.
-However, augmented user's frequently report dissociative experiences following augmented action.
+However, augmented user's frequently report dissociative experiences during or following augmented action.
 They do not experience agency.
@@
 A key question remains: How can we maintain agency during physical augmentation without training and over-fitting parameters?
-In other words, how can we create closed systems for users to experience natural augmentation?
+In other words, how can we create closed systems for users to experience \textit{natural} augmentation?
 
@@
 In this paper, we present a system that establishes a fast communication channel between the user's brain signals and a physical end effector, here EMS.
-The system controls the user's muscles through EMS at the time of the user's intent to interact, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
+The system controls the user's muscles through EMS at the time of the user's intent to act, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
 In our user study we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.

### 2023-03-09 signal_processing/pi_bemobil_config.m
export for erp per condition

@@ -2,8 +2,11 @@ clear bemobil_config
 
 %% General Setup
 
-path = 'P:\Lukas_Gehrke\2021-fastReach';
-%'/Users/lukasgehrke/Documents/publications/2021-fastReach';
+if strcmp(current_sys,'win')
+    path = 'P:\Lukas_Gehrke\2021-fastReach';
+else
+    path = '/Volumes/projects/Lukas_Gehrke/2021-fastReach';
+end
 
 bemobil_config.study_folder = [path filesep 'data' filesep]; %(NEEDS to have a filesep at the end, sorry!) 
 bemobil_config.filename_prefix = 'sub-';

### 2023-03-09 signal_processing/pi_extract_results.m

@@ -10,7 +10,7 @@ pi_bemobil_config;
 
 %% load data and parse events
 
-pIDs = 1:10;
+pIDs = 2; %1:10;
 conds = {'Baseline', 'EMS1', 'EMS2'};
 results = [];
 
@@ -24,40 +24,7 @@ for pID = pIDs
         [ALLEEG EEG CURRENTSET] = pop_newset(ALLEEG, EEG, 0,'gui','off');
         EEG = pi_parse_events(EEG);
         cz_ix = find(contains({EEG.chanlocs.labels},'Cz'));
-    
-        %% add erp pre-move class
-        if strcmp(cond{1}, 'Baseline') % consider delay from EMG onset to button press
-            
-            EMG = pop_eegfiltnew(EEG, 20, 100);
-            EEG.data(end,:) = EMG.data(end,:).^2;
-            
-            pre_move_event_ixs = find(contains({EEG.event.type}, 'reach_end'));
-            pre_move_events = EEG.event(pre_move_event_ixs);
-            pre_move_data = EEG;
-            pre_move_data.event = pre_move_events;
-            [pre_move_data.event.type] = deal('reach');
-            pre_move_erp = pop_epoch(pre_move_data, {'reach'}, [-1, 0]);
-            
-            pre_move_erp = mean(pre_move_erp.data(end,:,:),3);
-            pre_move_erp = movmean(pre_move_erp, 10);
-            
-            emg_onset_raw = min(find(pre_move_erp > prctile(pre_move_erp, 95)));
-            emg_onset_fine = max(find(diff(pre_move_erp(1:emg_onset_raw)) < 0));
-            
-            delay = (emg_onset_fine - EEG.srate) / EEG.srate;
 
-            pre_move_data = pop_select(pre_move_data, 'nochannel',{'EMG'});
-            pre_move_erp = pop_epoch(pre_move_data, {'reach'}, [-1.1 + delay, 0 + delay]);
-            
-            [pre_move_erp, pre_move_noisy_epochs] = pop_autorej(pre_move_erp, 'nogui','on','eegplot','off');
-                
-            erp = array2table(squeeze(pre_move_erp.data(cz_ix,26:end,:) - mean(pre_move_erp.data(cz_ix,1:25,:)))');
-        else
-            % TODO add the erp of second preceeding the ems trigger
-            % for trials where there was no ems trigger save an nan row.
-%             erp = 
-        end
-        
         %% behavior
         rt = {EEG.event.rt};
[…]

### 2023-03-10 signal_processing/pi_extract_bci_data.m
added C3, C4 to channels that are included for all participants

@@ -82,10 +82,15 @@ n_best_chans = 20;
 [best_chans_ixs, crit1, crit2] = rp_ERP_select_channels(pre_move_erp.data, idle_erp.data, EEG.srate/n_wins, 1); % extract informative channels
 sel_chans = best_chans_ixs(1:n_best_chans);
 
-cz_ix = find(contains({EEG.chanlocs.labels},'Cz'));
+chans_to_keep = {'C3', 'C4', 'Cz'};
 
-if sum(ismember(sel_chans, cz_ix)) < 1
-    sel_chans(end) = cz_ix;
+for chan = chans_to_keep
+    chan_ix = find(strcmp({EEG.chanlocs.labels}, chan{1}));
+
+    if sum(ismember(sel_chans, chan_ix)) < 1
+        sel_chans(2:end) = sel_chans(1:end-1);
+        sel_chans(1) = chan_ix;
+    end
 end
 
 %% save

### 2023-03-10 signal_processing/pi_extract_results.m

@@ -23,7 +23,7 @@ for pID = pIDs
             'streamtype', 'EEG', 'exclude_markerstreams', {});
         [ALLEEG EEG CURRENTSET] = pop_newset(ALLEEG, EEG, 0,'gui','off');
         EEG = pi_parse_events(EEG);
-        cz_ix = find(contains({EEG.chanlocs.labels},'Cz'));
+        cz_ix = find(strcmp({EEG.chanlocs.labels},'Cz'));
 
         %% behavior
         rt = {EEG.event.rt};

### 2023-03-10 signal_processing/restream_EEG.m
repo maintenance

@@ -1,14 +1,22 @@
 
-eeglab_ver("mac")
+pID = 2;
 
-% config
-pID = 13;
-study_folder = '/Users/lukasgehrke/Documents/publications/2021-fastReach/';
-subject_folder = fullfile(study_folder, 'data/study/0_source-data/', num2str(pID));
-fname = num2str(pID);
+%% config
+current_sys = "mac";
+% eeglab
+eeglab_ver(current_sys);
 
-% load data
-EEG = pop_loadxdf([fullfile(subject_folder, fname), '.xdf'] , 'streamname', 'BrainVision RDA', 'exclude_markerstreams', {});
+% addpath(genpath('D:\Lukas\signal-processing-motor-intent'));
+%addpath('/Users/lukasgehrke/Documents/publications/2021-fastReach/signal_processing');
+addpath('/Users/lukasgehrke/Documents/code.nosync/signal-processing-motor-intent');
+
+pi_bemobil_config;
+
+%% load data and parse events
+
+EEG = pop_loadxdf(fullfile(bemobil_config.study_folder, 'study', bemobil_config.source_data_folder, ...
+    ['sub-' sprintf('%03d', pID)], 'EMS1.xdf'), ...
+    'streamtype', 'EEG', 'exclude_markerstreams', {});
 
 % EEG = pop_eegfiltnew(EMG, 10);
 allEventsLats = [EEG.event.latency];

### 2023-03-10 signal_processing/pi_extract_bci_data.m
Topomaps
- export of channel names
- using mne python to create eeg topomaps

@@ -108,6 +108,7 @@ end
 idle = idle_erp.data;
 pre_move = pre_move_erp.data;
 
+writematrix(table({EEG.chanlocs.labels}'), fullfile(path, 'sel_chans_names.csv'));
 writematrix(sel_chans, fullfile(path, 'sel_chans.csv'));
 save(fullfile(path, 'pre_move'), 'pre_move');
 save(fullfile(path, 'idle'), 'idle');

### 2023-03-10 app/fastReach/bci_funcs.py
Added documentation

@@ -1,15 +1,33 @@
 import numpy as np
 
 def windowed_mean(data, windows = 10):
+    """Computes windowed mean of data
+
+    Args:
+        data (_type_): 2D array, usually chans x time
+        windows (int, optional): Number of windows data is split into . Defaults to 10.
+
+    Returns:
+        _type_: Windowed mean of data, of shape chans x windows
+    """
     stepsize = data.shape[1] // windows
     win_data = np.reshape(data, (data.shape[0], windows, stepsize))
     return np.mean(win_data, axis = 2)
 
 def base_correct(data):
+    """Subtracts baseline from data
+
+    Args:
+        data (_type_): 2D array, usually chans x time
+
+    Returns:
+        _type_: Baseline corrected 2D array, usually chans x time
+    """
     data_base_correct = data - data[:,0][:,None]
     return data_base_correct
 
 def drop_baseline(data, baseline_end_ix):
+    """Drops baseline from data"""
     data = data[:,baseline_end_ix:]
     return data
 

### 2023-03-10 app/fastReach/fastReach.py
Documentation, Bug Fixes and Smoothing BCI Output

@@ -10,7 +10,7 @@ from LSL import LSL
 import threading
 import ptext
 import random
-from bci_funcs import windowed_mean, base_correct, reshape_trial, drop_baseline
+from bci_funcs import windowed_mean, base_correct, drop_baseline, reshape_trial
 # import matplotlib.pyplot as plt
 
 EMS_RESET_TIME = time.time()
@@ -24,8 +24,11 @@ class fastReach:
 
     Args:
         pID (integer): participant ID used to save data
+        path (string): path to the data folder
         ems_on (boolean): whether to run with or without electrical muscle stimulation (EMS)
         arduino_port (string): port of connected arduino device
+        num_trials (integer): Number of trials to complete
+        debug (boolean): whether to print debug messages
     """
 
     def __init__(self, pID, path, ems_on, arduino_port, num_trials, debug) -> None:
@@ -72,16 +75,17 @@ class fastReach:
             model_path_eeg = data_path+'model_'+str(self.pID)+'_eeg.sav'
             chans = pickle.load(open(data_path+'chans_'+str(self.pID)+'_eeg.sav', 'rb'))
 
+            self.ems_training_delay = 2
             stim_delay = pd.read_csv(data_path+'delay.csv')
             self.ems2min = float(stim_delay.columns[0])
             self.ems2max = float(stim_delay.columns[1])
 
-            classifier_update_rate = 25
+            classifier_update_rate = 5
             data_srate = 250
             windows = 11
             baseline_ix = 1
             target_class = 1
-            threshold = .7
+            threshold = .8
 
             self.eeg = Classifier2('eeg_classifier', classifier_update_rate, data_srate, model_path_eeg, target_class, chans, threshold, windows, baseline_ix)
             self.eeg.start()
@@ -428,8 +432,12 @@ class fastReach:
         """Experiment logic. Runs different task components based on increasing time, then waits for button inputs to reset time and step through different experiment components.
 
         Args:
-            trial_type (string): Can be "training" or "experiment", "training" is always without EMS
-            num_trials (integer): Number of trials to complete
+            trial_type (string): Can be "training", "baseline", "ems1" or "ems2". 
+                - In 'training' mode, the ems is trigger 2 seconds after the fixation cross dissapears. 
+                This can be used to make participants comfortable with the EMS stimulation device in the trial logic.
+                - In 'baseline' mode, the ems is not triggered at all. This trial_type is used to obtain physiological training data.
[…]

### 2023-04-28 input/5_user_study_and_methods.tex
Updates from Overleaf

--- 2023-03-06 input/4_user_study_and_methods.tex
@@
 
-With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
+% With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
 To this end, we build a system that leverages the brain signals known to reflect the volitional intent to interact, i.e.
@@
 
-To assess if agency was preserved, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews.
-Our interviews followed a phenomenological structure \todo{add reference from AH review paper mueller}.
-We t
+Our idea to overcome this disruption in agency is to use EEG signals reflecting the intent to (inter)act and leverage them for the control of an end-effector, here EMS.
+We trained an EEG classifier to discriminate between two user states and applied it in real-time: either the user is predicted to be in an \textit{idle}-- or a \textit{pre-movement} state.
+Is the user predicted to be in a \textit{pre-movement} state, EMS is triggered, augmenting their action.
 
-\subsection{Participants}
+\input{input/4_prototype}
+
+To assess whether our prototype preserves agency during action augmentation, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews.
+Our interviews followed a phenomenological structure~\citep{Danry2022-xk}.
+
+\subsection{User study}
+
+\subsubsection{Participants}
 
@@
 
-\subsection{Apparatus}
+\subsubsection{Apparatus}
+\todo{add the components that were used to run the STUDY, touchpad, keyboard etc.}
 The experimental setup, depicted in Figure~\ref{}, comprised: (1) a wrist-mounted wearable HTC VIVE tracker for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
@@
 
+% TODO change this to the use of EMG for now
 \indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz.
@@
 
-\subsection{Task}
+\subsubsection{Task}
 \missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
@@
 
-\subsection{Training of Classifiers from Training Phase Data}
-% Brain-Computer Interface to Detect EEG Readiness Potentials }
-% Classifiying EEG Signals of the Intent to Interact
-
-Before the test phase of our experiment, the data recorded during the training phase was processed in three consecutive steps: (1) we extracted movement onsets in each trial to construct three event classes: an \textit{idle} class representing resting background activity, a \textit{movement} class and a \textit{pre-movement} class, representing the intent to interact.
-Next, (2) a motion classifier was trained on \textit{movement} and \textit{idle} data segments and (3) an EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels.
-To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually.
-The EEG classifier was then used in the test phase to detect RPs and trigger EMS.
-The motion classifier was used to make the interaction more robust.
-When it detected the participants hand to be moving, it deactivated the control loop from RP to EMS.
-
-\missingfigure{some picture from the setup and prototype}
-
-\indent\textbf{(1) Labelling Pre-movement, Movement and Idle Data Segments.}
-In order to obtain training data segments for the \textit{pre-movement} class, a detection algorithm was applied on the hand velocity time series.
-From the training data, the raw hand motion data was filtered with a 6Hz low-pass filter and re-sampled to match the EEG sample rate using BeMoBIL Pipeline functions \textit{xdf2bids} and \textit{bids2set}\footnote{https://github.com/BeMoBIL/bemobil-pipeline}.
-Subsequently the first derivative was computed and velocity was extracted.
-Then, for each trial a two step thresholding process was applied to extract the exact time of movement onset: (1) the first time point when the velocity exceeded 70\% of the maximum velocity in the trial was selected, then (2) the signal was flipped at that time point and the time point where the flipped signal first fell below 10\% of the maximum in the remaining data was selected as the movement onset.
-
-The three event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets, \textit{movement} from 0 to 1000 ms succeeding the movement onsets and \textit{idle}, a one second data segment that fell exactly between two succeeding trials, i.e.
-time-locked to the inter-trial-interval.
-Here, participants where looking at a fixation cross on the screen with their hand resting on the table.
-
-\indent\textbf{(2) Training of Hand Motion Classifier.}
-An LDA was trained using the mean velocity per trial and movement and idle time segment.
-Hence, 60 values for each class were subjected to train the motion classifier.
-Since in the idle time segments almost no movement occurred we added some artificial noise.
-To this end, the idle features were multiplied by a factor of 3, chosen by simple trial and error.
-In the real-time feedback, this rendered the classifier less susceptible to false positives, i.e.
-detecting a movement when there was none.
-
-\indent\textbf{(3) Channel Selection \& Training of EEG Classifier.}
-For the EEG classifier, we followed the approach outlined by Schultze-Kraft et al.\cite{} with slight modifications.
-First, the 20 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure.
-For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment.
-The resulting value thus indicated how much the signal had changed in the course of the 1s segments.
-In line with the literature there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment.
-Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments.
-The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment.
-The ranks were joined and the 20 best channels were selected.
-In any case, channels C3, C4 and Cz were kept for every participant as these are most frequently reported in studies on the RP.
-
-To train the LDA EEG classifier, we extracted spatiotemporal features from the EEG; 10 windowed means were extracted for each segment and channel.
-These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment.
-The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.
-
-\subsubsection{Real-time Feedback}
-
-During the test phase, both classifiers were applied to the real-time data.
-With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier.
-For the EEG data streaming at 250 Hz, the EEG data of the last 1000ms was buffered for the selected best channels and 10 windowed means per channel were computed.
-Identical to the training data, these features were baseline corrected by subtracting the mean from the last 100 ms preceding the last second, i.e.
--1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector.
-Hence, both classifiers yielded one output value at each sample point.
-
-The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability.
-Additionally, EMS was only triggered when the hand was \textit{not} moving.
-In each trial, as soon as the hand was in motion once after trial start, the EEG-EMS control was deactivated for the rest of the trial.
-This guaranteed that participants experienced EMS only during movement preparation.
-
-\subsection{Interview \& Questionnaire}
+\subsubsection{Measures of agency experience: Intentional Binding, Questionnaire \& Interview}
 
@@
 
-\subsection{Statistical Analyses}
+\subsubsection{Statistical Analyses}
 \todo{how where the results analyzed: sentiment analyses and statistics of IB data and questionnaire results}
@@
 
-\subsection{Reproducing Results and Data Availability}
+\subsubsection{Reproducing Results and Data Availability}
 Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}.

### 2023-04-28 input/7_discussion.tex
Updates from Overleaf

\section{Discussion}

% summary here! BLUF!

% Outlook Ideas
% - Strength of EMS could be set by the amplitude of the RP -> is there evidence for a correlation between amplitude and 'strength' of volitional action?
%[ - EEG Signal may be contaminated by eye movement, maybe a worthwhile thing to build this interaction relying solely on eye tracking information, gaze fixating on the target before moving]
%- not only physical integration but also other adpative interfaces, or control of mobile phone (see Cornelio2022 review); % say here somewhere what action augmentation is and how it differs from body and environment/outcome augmentation

### 2023-04-28 input/2_introduction.tex

--- 2022-09-14 input/2_introduction.tex
@@
 % Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
-Technological advances in hardware that allow physically moving a user's body have reignited the dreams of overcoming human limitations, recorvering lost abilities as well as simplifying skill acquisition.
-Wearable exoskeletons move a user's body by applying forces to the extremities, for example by pulling on the fingertips (see Dexmo Haptic Glove \footnote{https://www.dextarobotics.com/}) while electrical muscle stimulation (EMS) makes the user's extremities move by sending current into the muscle activating nerves.
-These technologies are becoming increasingly seamless to use.
-However, augmented user's frequently report dissociative experiences following augmented action.
+
+Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}.
+There are two main technologies to physically augment users' actions: First, through the use of mechanical actuators, exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips, and second, by making the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS).
+However, augmented user's frequently report dissociative experiences during or following augmented action~\citep{}.
 They do not experience agency.
+\todo{this needs a good citation}
 
-The sense of agency refers to the experience of being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
-This being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ubd} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
-The key challenge in human \textit{physical augmentation} is to create an experience of pre-reflective agency, so user's feel as though they are in the "driving seat" once again.
+The experience of agency means being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
+It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}.
+The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so user's feel as though they are in the ``driving seat'' once again.
 % if something happens for you in an interaction there is a cost to that:
-% - impacts sense of agency which in turn makes user's engage less
-% - impacts precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control
-
-% What have others done to address this and what do we propose "instead"
-% summarize Kasahara's work
-This challenge has driven researcher
+% - impacts sense of agency which in turn may makes user's engage less -> citation?
+% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?
 
@@
 % Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
-But how can we build a physical augmentation system that keeps user's engaged in the action, preserving sense of agency? In other words how can we design for users to experience natural augmentation?
+% What have others done to address this and what do we propose "instead"
+Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping targets on a touchscreen can be increased by augmenting the user through the use of EMS while maintaining their sense of agency.
+The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e.
+how much earlier the action of the user can be elicited.
+However, setting this gain factor may depend on each specific task and requires fitting the parameters over and over again.
+A key question then remains: How can we maintain agency during physical action augmentation without any training or running the risk of over-fitting task specific parameters?
+In other words, how can we design closed systems without the need for external tuning, to deliver an integrated, \textit{natural}, experience of user's augmented actions?
+% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.
+
+\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.
+<- better use this for the teaser image.
+Here better present the main results, i.e.
+sense of agency scores?}
+\input{input/1_teaser_figure}
 
@@
 % keep this short
-In this paper, we present an augmentation that maintains the user's sense of agency.
-Our system establishes a fast communication channel between the user's thoughts and a physical end effector, here EMS.
-The system controls the user's muscles through EMS at the time of the user's intent to interact, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
-In our user study we then investigated whether keeping the physical impact on the user's body in line with their intention to move, preserves their sense of agency.
+In this paper, we present a fast communication channel between the user's brain signals and a physical end effector, here EMS.
+Crucially, our system controls the user's muscles through EMS at the time of the user's \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
+In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
+% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.
 
-\subsection{Preserving Agency using Brain Signal of the Intent to Interact}
+% \subsection{Preserving Agency using Brain Signals of the Intent to (Inter)act}
 % 4. (long) Anwser with teaser image
 % Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
-% Now extend into a longer paragraph with a teaser image explaining the system in the following subsection
-
-% not only physical integration but also other adpative interfaces, or control of mobile phone
-
-\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.}
-\input{input/1_teaser_figure}

### 2023-04-28 input/3_related_work.tex

@@
 
-We present a bodily integration experience moving the finger of a user after the user intends to do so, but before the user directs its muscles to move.
+% In this paper, we present a physical action augmentation prototype and a user study to investigate the experience of agency during the use of the prototype.
+Hence, our research builds on previous qualitative and quantitative research on the sense of agency.
+% on actuated haptic systems, brain-computer interfacing as well as
+% moving the finger of a user when the user intends to do so, but before the user directs their muscles to move.
 
-\subsection{The Disruption in Sense of Agency in Augmented Humans}
+The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
 
-\subsection{Assessing the Sense of Agency when Integrating with Computers}
-% how can we measure our outcome measure sense of agency?
+\subsection{Measuring agency experience}
+% explicit measures - reflective (short paragraph)
 % - questionnaires
-% - intentional binding
 % - phenomenological interview
 
-\subsection{Brain-Computer Interfaces for Intent Prediction}
-% Single-trial Classification of the Intent to Interact
-Our idea to overcome the disruption in agency is to use EEG signals reflecting (inter)action intent.
-One such signal is the Readiness Potential (RP), or \textit{lateralized readiness potential}.
-It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
-It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand.
-Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
-However, evidence abounds for its role in action preparation.
-An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
-The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after.
-The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}.
-A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}.
-In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible.
-After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for.
-They found participants \textit{not} to be able to suppress their RP.
-This inability to suppress the RP renders it a reliable feature for classification.
-While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.
+% implicit measures - pre-reflective (longer paragraph)
+% - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here
+
+\textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.
+
+% what does that have to do with agency?
+\subsection{Sense of agency and actuated haptic systems}
+
+% close this paragraph with the problem that of the control mechanism of the

### 2023-04-28 input/4_prototype.tex

--- 2023-03-06 input/4_user_study_and_methods.tex
@@
-\section{Methods}
+\subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}
 
-With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
-To this end, we build a system that leverages the brain signals known to reflect the volitional intent to interact, i.e.
-the EEG readiness potential, to trigger electrical muscle stimulation to cue a reaching movement.
+To discriminate between the two \textit{action} states, we leveraged the Readiness Potential (RP), or \textit{lateralized readiness potential}.
+It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
+It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand.
+Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
+However, evidence abounds for its role in action preparation.
+An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
+The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after.
+The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}.
+A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}.
+In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible.
+After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for.
+They found participants unable to suppress their RP.
+This inability to suppress the RP renders it a reliable feature for classification.
 
-To assess if agency was preserved, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews.
-Our interviews followed a phenomenological structure \todo{add reference from AH review paper mueller}.
-We t
+% While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.
 
-\subsection{Participants}
+% \subsection{Action augmentation}
+% TODO shorten this down to only the 'prototype' part -> EEG controlling the arduino switch.
+make obvious that what is attached to the switch can be different -> here in this particular study we used EMS -> see User Study
 
-\todo{add actual number} Ten participants (\todo{add mean age and sd}) from our local institution participated in our study.
-All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
-Participants were compensated with 12 Euro per hour of study participation.
-Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form.
-
-\subsection{Apparatus}
-The experimental setup, depicted in Figure~\ref{}, comprised: (1) a wrist-mounted wearable HTC VIVE tracker for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
+% \subsubsection{Apparatus}
+% The experimental setup, depicted in Figure~\ref{}, comprised: (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
 To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
 
-\indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz.
-Therefore, we ran a simple Unity3D scene and streamed the data via the labstreaminglatyer (LSL)\footnote{https://github.com/sccn/labstreaminglayer}.
-
-\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
+% \indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
 Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
@@
 
-\missingfigure{show participant with measurement system and description next to each element}
+% \missingfigure{show participant with measurement system and description next to each element}
 
-\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
+% \indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
 We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles.
@@
 The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.
-
-\subsection{Task}
-\missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
-Participants performed a simple reaching task: resting their hand on the table, participants were instructed to press a specified key on a notebook in front of them.
-For each trial, a key was randomly selected from one of two groups of letters: (1) 'a', 's', 'd' or (2) 'j', 'k', 'l'.
-The selected group was then changed with every trial.
-This was done to keep the task engaging while maintaining a comparable reaching movement between trials.
-
-The interaction flow for one trial of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key.
-In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move".
-(4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone.
-(5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement.
-Following an inter-trial-interval the next trial started.
-
-\subsubsection{Experiment Design and Procedure}
-Participants completed three experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
-
-%The condition of the second and third block, see below, were alternated between participants.
-
-In a second block, participants completed 80 trials of the same task and trial order as in the first block.
-Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background.
-After the target letter was shown to the participants and they were preparing to reach out and press the corresponding key, the BCI was live.
-Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.
-
-Both experimental blocks two and three were followed by a mixed methods data collection of participants' perceived agency.
-
-In summary, the experiment consisted of four phases: (1) a setup phase; (2) a first experimental block: the training phase; (3) a second experiment block: the test phase; and (4) a test battery of questionnaire and interview to inquire about their experienced agency.
 
@@
 This guaranteed that participants experienced EMS only during movement preparation.
-
-\subsection{Interview \& Questionnaire}
-
-To assess if agency was preserved, three measures were recorded during and after the test phase.
-
-\indent\textbf{(1) Intentional Binding.}
-At the end of each trial participants were tasked to compare the duration of their arm movement to a tone and indicate whether the tone was longer or shorter.
-We set the initial duration at 500 ms for each participant.
-When participants indicated the tone to be longer or shorter by pressing either the up or down key we added, or respectively subtracted, 100 ms from the tone duration.
-This new duration was then used for the next trial and so on.
-With 60 trials in the training phase, we hypothesized participants would oscillate 50-50 around a target duration after about 2/3 of the training trials.
-The final duration after all training trials was then used again as the start duration during the test phase.
-In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS.
-A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.
-
-\indent\textbf{(2) Questionnaire.}
-In line with~\cite{Hornbaek} we prompted participants with three statements and a 7-point Likert scale centered at 0.
-The first one ``I felt like the device I was using was part of my body'' was in relation to body ownership.
-The following two statements ``It felt like I was in control of the movements during the task.'' and ``What is the degree of control you felt?'' prompted participants on their experienced agency.
-
-\indent\textbf{(3) Interview.}
-We interviewed user's about the experience with the EMS integration.
-We asked open how questions in a collaborative interview inquiring about the user's lived experience.
-After prompting user's to recall their experience we started the interview with: "how did it feel like?".
-If user's made reference themselves to the EMS device we followed up by inquiring: "why do you think it (the EMS trigger) happened?", "What do you think triggered it?".
-The goal was to learn about the source attribution user's made about the EMS trigger.
-Since we leveraged the RP, we hypothesized that user's would attribute, at least in part, the EMS trigger to themselves, i.e.
-their movement intention.
-Next, we pressed deeper into learning about the cooperative nature between user and system.
-We asked "Do you believe you did it alone?" and when they made reference to a specific situation where the EMS was triggered we followed up with "What did it feel like, did you had the feeling the system controlled you or did you cooperate?" and "What would have to be there for it to feel like a cooperation?".
-We closed the interview with two additional exploratory questions: "Did the system influence your performance?
-If yes, how and why?" as well as asking them about any application scenarios they have in mind.
-
-To not break the user's immersion by disturbing the flow of the interaction we decided against asking qualitative questions at the end of each and every trial.
-
-\subsection{Statistical Analyses}
-\todo{how where the results analyzed: sentiment analyses and statistics of IB data and questionnaire results}
-
-% ERP with significance mask for Cz? Assessing significance of classifier
-
-% IB Task: ttest of final ib time between with and without EMS -> # subjects input values
-% did they increase the IB task in more than 50% of the trials after EMS?
-
-% questionnaires: ttest against zero to report direction
-
-% interviews: anecdotal summary and sentiment analyses with two groups splitting questionnaires at 0
-
-% correlation with EEG classifier accuracy?
-
-% evaluation of EEG classifier: can report false negatives, how often was it missing to predict a movement onset?
-hand was moving without EMS being triggered occurred how often?
-Not possible to evlaute false positives as no ground truth available
-
-\subsection{Reproducing Results and Data Availability}
-Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}.
-BIDS formatted data is hosted on openneuro \cite{}.

### 2023-06-19 input/2_introduction.tex
Updates from Overleaf

@@
 % Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
+\todo{add something even more general here: e.g. increase human wellbeing by XY}
+Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}.
+There are two main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e.
+exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips.
+Another technology makes the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS).
+However, augmented users often report dissociative experiences, frequently disrupting their sense of agency~\citep{Gilbert2017-ze, Gilbert2019-uc}.
 
-Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}.
-There are two main technologies to physically augment users' actions: First, through the use of mechanical actuators, exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips, and second, by making the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS).
-However, augmented user's frequently report dissociative experiences during or following augmented action~\citep{}.
-They do not experience agency.
-\todo{this needs a good citation}
-
-The experience of agency means being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
+Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
 It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}.
-The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so user's feel as though they are in the ``driving seat'' once again.
+The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
 % if something happens for you in an interaction there is a cost to that:
@@
 % What have others done to address this and what do we propose "instead"
-Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping targets on a touchscreen can be increased by augmenting the user through the use of EMS while maintaining their sense of agency.
-The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e.
-how much earlier the action of the user can be elicited.
-However, setting this gain factor may depend on each specific task and requires fitting the parameters over and over again.
-A key question then remains: How can we maintain agency during physical action augmentation without any training or running the risk of over-fitting task specific parameters?
-In other words, how can we design closed systems without the need for external tuning, to deliver an integrated, \textit{natural}, experience of user's augmented actions?
+Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimuli on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency.
+The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited.
+However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy.
+A key question then remains: How can we maintain agency during physical action augmentation without controlling the environment?
+In other words, how can we design a closed-loop system to deliver an integrated, \textit{natural}, experience for users' augmented actions?
 % At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.
@@
 % keep this short
-In this paper, we present a fast communication channel between the user's brain signals and a physical end effector, here EMS.
-Crucially, our system controls the user's muscles through EMS at the time of the user's \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
+In this paper, we present a fast communication channel between a user's brain signals and a physical end effector, here EMS.
+Crucially, our brain-computer interface controls the user's muscles through EMS at the time of their \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
 In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.

### 2023-06-19 input/4_prototype.tex
Updates from Overleaf

@@
+\section{Agency Prototype}
+Here we present a prototype to overcome the disruption in experiencing agency during action augmentation.
+Our proposed system controls an end-effector, here EMS, by leveraging EEG signals reflecting the intent to (inter)act.
+To control the interaction in real-time, an EEG classifier is used to discriminate between two user states: either the user is predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or a \textit{pre-movement} state, indicating the presence of an intent to act.
+When the system predicts a \textit{pre-movement} state, EMS is triggered in real-time, augmenting the user's action even before their own voluntary motor command.
+
 \subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}
+% alternative headings:
+% Triggering EMS using EEG Signals
 
-To discriminate between the two \textit{action} states, we leveraged the Readiness Potential (RP), or \textit{lateralized readiness potential}.
-It is an amplitude fluctuation that has been frequently observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
-It is reliably observed in electrodes placed over the motor cortex contralateral to the acting hand.
+To discriminate between the two \textit{action} states, we leveraged the EEG Readiness Potential (RP), or \textit{lateralized readiness potential}.
+It is an amplitude fluctuation that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
+The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand.
 Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
 However, evidence abounds for its role in action preparation.
-An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
+A RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
 The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after.
@@
 A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}.
-In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform a voluntary movement so that the accompanying RP amplitude was as small as possible.
+In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform voluntary movements while keeping accompanying RP amplitudes as small as possible.
 After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for.
@@
 
-% While its role in movement preparation remains somewhat controversial, it is established as a signal occurring during movement planning and motivates a usage as a reliable predictor of movement.
+\subsection{Prototype components}
+Our system, depicted in Figure~\ref{}, comprised: (1) a 64-channel EEG system, (2) a 1 channel Electromyography (EMG) system, (3) a simple Arduino controlled switch, and (4) a medically-compliant EMS device connected via two electrodes worn on the arm.
+More information on the specific hardware specifications is given in the methods and user study sections below.
+To assist readers in replicating our prototype, we provide the necessary technical details and source code for data processing, classifier training, and real-time application\footnote{[anonymized]}.
+We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the EEG and EMG data streams available in the network and synchronize the recordings of EEG data, motion capture, and an experiment marker stream that marked sections of the study procedure.
 
-% \subsection{Action augmentation}
-% TODO shorten this down to only the 'prototype' part -> EEG controlling the arduino switch.
-make obvious that what is attached to the switch can be different -> here in this particular study we used EMS -> see User Study
+\missingfigure{show participant with measurement system and description next to each element}
 
-% \subsubsection{Apparatus}
-% The experimental setup, depicted in Figure~\ref{}, comprised: (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
-To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
-
-% \indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
-Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
-One electrode was placed under the left eye to provide additional information about eye movements (vEOG).
-After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes.
-EEG data was recorded with a sampling rate of 250 Hz.
-We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.
-
-% \missingfigure{show participant with measurement system and description next to each element}
-
-% \indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
-We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles.
-This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus).
-The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer.
-The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.
-
-\subsection{Training of Classifiers from Training Phase Data}
+\subsection{Brain-Computer Interface}
+% alternative headings:
 % Brain-Computer Interface to Detect EEG Readiness Potentials }
@@
 
-Before the test phase of our experiment, the data recorded during the training phase was processed in three consecutive steps: (1) we extracted movement onsets in each trial to construct three event classes: an \textit{idle} class representing resting background activity, a \textit{movement} class and a \textit{pre-movement} class, representing the intent to interact.
-Next, (2) a motion classifier was trained on \textit{movement} and \textit{idle} data segments and (3) an EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels.
-To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually.
+The presented classification scheme required individual training data per participant, as transfer learning remains challenging with EEG~\cite{Wan2021-zz}.
+Therefore, we initially obtained training data on a variant of our task that did not include any muscle stimulation.
+In our task, participants were instructed, following an initial waiting period, to tap a touchscreen at their own volition with their ring finger, see figure ~\ref{fig:teaser}.
+
+\subsubsection{Labelling movement classes using EMG}
+To obtain behavioral labels at a high temporal resolution, we leveraged EMG from the flexor digitorum profundus to detect finger movement onsets of the ring finger.
+Subsequently the data were labelled into two classes: an \textit{idle} class representing resting background activity and a \textit{pre-movement} class, representing the intent to (inter-)act.
+To this end, EMG amplitudes of the second preceding the screen tap were band-pass filtered from 20 to 100 Hz and subsequently squared.
+Next, noisy data segments were determined.
+A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters.
+To label the time of movement onset, the first sample where the amplitude exceeded the 95th percentile of the data epoch was selected.
+The two event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets and \textit{idle}, a one second data segment between trials where participants were looking at a fixation cross.
+% from isoutlier documentation (mean segment amplitude exceeding 3 scaled median absolute deviation)
+
+% \missingfigure{some picture from the setup and prototype}
+
+\subsubsection{Selecting discriminative EEG channels}
+In a first step to prepare the EEG data for classification, noisy data segments were rejected using the identical method as for rejection of EMG segments described above.
+Then, to decrease dimensionality of the EEG data, a selection of discriminative channels was conducted following an approach outlined by~\citep{Schultze-Kraft2021-cu} albeit with slight modifications.
+
+First, for each channel and both \textit{idle} and \textit{pre-movement} class, the mean slope of the amplitude was obtained by fitting a linear regression, indicating how much the signal had changed in the course of the 1s segments.
+In line with the literature on the RP, there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment.
+Then, all channels were sorted (1) in descending order by slope in the \textit{pre-movement} segments and (2) in ascending order by slope in the \textit{idle} segments.
+The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment.
+The ranks were joined and the 10 highest-ranking channels were selected.
+In any case, channels C3, C4, and Cz were included in the 10 selected channels for every participant even if they were not within the 10 highest ranks as these are most frequently reported in studies on the RP.
+
+\subsubsection{Training of EEG classifier}
+To classify EEG data a regularized linear discriminant analysis (LDA) was trained for each participant individually.
 The EEG classifier was then used in the test phase to detect RPs and trigger EMS.
-The motion classifier was used to make the interaction more robust.
-When it detected the participants hand to be moving, it deactivated the control loop from RP to EMS.
+The classifier was trained using the slopes of the \textit{idle} and \textit{pre-movement} segments of the 10 most discriminative channels as features, hence comprising a 10 dimensional feature vector.
+\todo{extend this after settling on a final feature set}
 
-\missingfigure{some picture from the setup and prototype}
+\subsubsection{Real-time application and EMS control}
+During real-time application, the EEG data was buffered for the last second for the selected channels.
+The slope features were then computed and applied to the model.
+An Arduino was used to flip a switch whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability, thereby triggering the attached actuating hardware.
+\todo{this may still change, adapt accordingly later}
+In our current prototype, we used a medically-compliant EMS device.
+We constrained the times when the switch could be flipped to specific moments in the experiment where participants were getting ready to perform an action, thereby limiting the impact of false-positive classification.
+% This guaranteed that participants experienced EMS only during movement preparation.
 
-\indent\textbf{(1) Labelling Pre-movement, Movement and Idle Data Segments.}
-In order to obtain training data segments for the \textit{pre-movement} class, a detection algorithm was applied on the hand velocity time series.
-From the training data, the raw hand motion data was filtered with a 6Hz low-pass filter and re-sampled to match the EEG sample rate using BeMoBIL Pipeline functions \textit{xdf2bids} and \textit{bids2set}\footnote{https://github.com/BeMoBIL/bemobil-pipeline}.
-Subsequently the first derivative was computed and velocity was extracted.
-Then, for each trial a two step thresholding process was applied to extract the exact time of movement onset: (1) the first time point when the velocity exceeded 70\% of the maximum velocity in the trial was selected, then (2) the signal was flipped at that time point and the time point where the flipped signal first fell below 10\% of the maximum in the remaining data was selected as the movement onset.
+%%%% Resources
 
-The three event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the movement onsets, \textit{movement} from 0 to 1000 ms succeeding the movement onsets and \textit{idle}, a one second data segment that fell exactly between two succeeding trials, i.e.
-time-locked to the inter-trial-interval.
-Here, participants where looking at a fixation cross on the screen with their hand resting on the table.
-
-\indent\textbf{(2) Training of Hand Motion Classifier.}
-An LDA was trained using the mean velocity per trial and movement and idle time segment.
-Hence, 60 values for each class were subjected to train the motion classifier.
-Since in the idle time segments almost no movement occurred we added some artificial noise.
-To this end, the idle features were multiplied by a factor of 3, chosen by simple trial and error.
-In the real-time feedback, this rendered the classifier less susceptible to false positives, i.e.
-detecting a movement when there was none.
-
-\indent\textbf{(3) Channel Selection \& Training of EEG Classifier.}
-For the EEG classifier, we followed the approach outlined by Schultze-Kraft et al.\cite{} with slight modifications.
-First, the 20 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure.
-For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment.
+% old version
+% First, the 10 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure: For both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment.
 The resulting value thus indicated how much the signal had changed in the course of the 1s segments.
@@
 The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment.
-The ranks were joined and the 20 best channels were selected.
+The ranks were joined and the 10 best channels were selected.
 In any case, channels C3, C4 and Cz were kept for every participant as these are most frequently reported in studies on the RP.
-
-To train the LDA EEG classifier, we extracted spatiotemporal features from the EEG; 10 windowed means were extracted for each segment and channel.
-These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment.
+% An EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels.
+To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually.
+% These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment.
 The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.
 
-\subsubsection{Real-time Feedback}
-
-During the test phase, both classifiers were applied to the real-time data.
+% During the test phase, both classifiers were applied to the real-time data.
 With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier.
@@
 Hence, both classifiers yielded one output value at each sample point.
-
-The EMS was triggered whenever the EEG classifier predicted the \textit{pre-movement} class with an above .7 probability.
-Additionally, EMS was only triggered when the hand was \textit{not} moving.
-In each trial, as soon as the hand was in motion once after trial start, the EEG-EMS control was deactivated for the rest of the trial.
-This guaranteed that participants experienced EMS only during movement preparation.

### 2023-06-19 input/3_related_work.tex

@@
 
-The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
+Our research builds on previous engineering work in brain-computer interfacing and on theories of the sense of agency as discussed in cognitive science and neuroscience.
+
+\subsection{The sense of agency in human-computer interaction}
+% LT to write first version
+\begin{comment}
+
+%%% theory on sense of agency 
+Agency is largely explained with a comparator model, describing internal comutational mechanisms of human action control 
+-- comparator model (Blakemore 2002, Frith 2000, Frith 2005)
+-- integrate predictive coding??
+
+\subsection{neural basis of sense of agency}
+
+\end{comment}
+
+The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low-level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
 
 \subsection{Measuring agency experience}
+% LT to write first version
+\begin{comment}
+
 % explicit measures - reflective (short paragraph)
 % - questionnaires
+
+%% from bergström 22
+- read 18, 37 from  Bergstöem 
+- 14,14,19.31,39,41,48
+11,20,33
+
+Sense of control is mostly measured with simple items with ratings like "It felt like I was in control of the hand I was looking at” rated on a 7-point Likert scale, +3 indicating strong agreement and −3 indicating strong disagreement (Longo and Haggard) or  "Indicate how much it felt like moving the joystick caused the object on the computer screen to move” as a measure of explicit sense of agency, also rated on a 7-point scale" (11)
+item asking participants to indicate the degree of control they have felt over the changes on the screen [1]. We used this item also as an agency measure (1)
+
+Individuals’ sense of agency can be estimated with explicit self-reported measures using Likert’s scales or percentages ( van der Wel et al., 2012; Chambon et al., 2013),“
+
 % - phenomenological interview
@@
 % implicit measures - pre-reflective (longer paragraph)
+-subjective measures address a high level-SoA  based on subjective judgments of feeling of control (Barlas and Kopp)
+- proxy for the low-level SoA as they do not require conscious reflection on one’s SoA (Synofzik et al., 2008a; Desantis et al., 2011; Moore and Obhi, 2012; Barlas and Obhi, 2014).“ ([Barlas und Kopp, 2018, p.
+2]
+- one such measure is intentional binding
+
 % - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here
+- „refers to the perceived temporal attraction between voluntary actions and their outcomes (Haggard et al., 2002).
+More clearly, the temporal interval between actions and outcomes is perceived as shorter when these outcomes are produced by voluntary actions compared to when they follow, for instance, involuntary movements or external causes (Haggard et al., 2002).“ ([Barlas und Kopp, 2018, p.
+2]
+- While details about the relationship between measures of intentional binding and SoA are far from clear, they are extensivly used
+
+- SoA is influeced by different processes, 
+    - one of the is the source of action selection (self selected vs. externally instructed) (Chambon et al. 2014, Harrad 2017)
+    - another: motor control processes (read Sidarus 2017a)
+    -
+
+-
+
+- Read More 2009 (interval estimations were shorter (i.e., stronger binding) in voluntary than involuntary movements) 
+- add figuure action binding and effect binding (e.g.  in WEN and Imamizu)
+-
 
 \textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.
+When a voluntary action is causally linked with a sensory outcome, the action and its consequent effect are perceived as being closer together in time.
+This effect is called intentional binding.“ (from Jo 2014)
 
 % what does that have to do with agency?
+
+- intentional binding effects are are sometimes descrepant from explicit judgment of agency (see 30-33 in Wen adn Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
+- Intentional binding is e.g. influenced by causal relations between events, attention, arousal (see Wen and Imamizu)
+
+- EEG patterns related to readiness potential and intentional  binding sometimes show disceprant patterns ( see WEN and Imamizu 38 und 39) (wittmann 2014, Goldberg 2017)
+
+- alternatives to intentional binding - sensory attenuation / visual attention
+
+% RESEARCH GAP
+
+- reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
+- maintain sense of agency in augmented interactions 
+- doing this by leveraging action intention in brain
+
+----
+involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
+it has been shown, that mere peripheral body movements, elicted by TMS simulations, produce a perceptual repulsion oppisite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
+
+There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome
+
+-
+
 \subsection{Sense of agency and actuated haptic systems}
 
-% close this paragraph with the problem that of the control mechanism of the
+\ subsection{Movement Intention}
+idea: when we now compare completely voluntary movements with movements triggerd by the brain signal of our intention to move - what happens?
+
+% close this paragraph with the problem that of the control mechanism of the actuator hardware and the disruotion in agency
+
+% PosssessedHand, Gilbert-I miss being me,
+
+\end{comment}

### 2023-06-19 input/5_user_study_and_methods.tex

@@
-\section{Methods}
-
-% With this study we wanted to find out whether agency can be preserved during \textit{physical actuation} of the user's body when their own brain signals serve as the control signal.
-To this end, we build a system that leverages the brain signals known to reflect the volitional intent to interact, i.e.
-the EEG readiness potential, to trigger electrical muscle stimulation to cue a reaching movement.
-
-Our idea to overcome this disruption in agency is to use EEG signals reflecting the intent to (inter)act and leverage them for the control of an end-effector, here EMS.
-We trained an EEG classifier to discriminate between two user states and applied it in real-time: either the user is predicted to be in an \textit{idle}-- or a \textit{pre-movement} state.
-Is the user predicted to be in a \textit{pre-movement} state, EMS is triggered, augmenting their action.
-
-\input{input/4_prototype}
-
-To assess whether our prototype preserves agency during action augmentation, we used a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as qualitative interviews.
+\section{User study \& Methods}
+With this study we wanted to find out whether agency can be preserved during \textit{physical action augmentation} when the user's own brain signals serve as the control signal.
+To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as a qualitative interview.
 Our interviews followed a phenomenological structure~\citep{Danry2022-xk}.
 
-\subsection{User study}
-
 \subsubsection{Participants}
-
 \todo{add actual number} Ten participants (\todo{add mean age and sd}) from our local institution participated in our study.
@@
 Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form.
+The experiment was approved by the ethics committee of [anonymized]
+%the Department of Psychology and Ergonomics at TU Berlin (tracking number: BPN\_GEH\_2\_230130220421).
 
+% TODO continue here 19.05.2023
 \subsubsection{Apparatus}

### 2023-06-19 input/3_related_work.tex
introduction and prototype sections, working on related work

@@
 
-There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome
+There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome (Buhner 2012)
 

### 2023-07-26 input/3_related_work.tex
Updates from Overleaf

--- 2023-06-19 input/3_related_work.tex
@@
 \section{Related Work}
+%Intro
+In a simplified world, intentional movements directly leading to external events provide a strong sense of responsibility and control over the action's outcome.
+For instance, when we press a piano key, it is unequivocally clear that we are the ones performing and causing the action.
+However, the introduction of technology can complicate this relationship.
+Imagine playing a computer game where a key press controls an avatar playing the piano.
+The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the action.
+Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of control.
+
+This lingering sense of control might be due to our intentions being reflected in the resulting action (sources needed).
+However, in augmented interactions, the dynamics change.
+When technology moves our fingers, our body performs the action, but we may experience a diminished feeling of causing the action [sources - augmented actions lead to less agency; check Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009, Haggard, Clark 2002].
+
+% Why is feeling of control important
+Because the feeling of control is related to (check for sources on learning effects etc.), understanding the reason for its loss in augmented interactions is relevant.
+
+% Theorie
+
+The concept of eliciting a sense of agency is commonly explained within the framework of motor control (see Frith and collegus 2000).
+When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (the efference copy).
+However, in situations where the movement is elicited by an external cause, as in augmented actions, the predicted sensory consequences may differ or be absent altogether.
+In such cases, the actual sensory feedback does not match the predicted sensory feedback.
+
+According to this account, the feeling of agency arises only when the predicted sensory feedback perfectly corresponds to the actual sensory feedback.
+In other words, when our actions and their outcomes align seamlessly, we experience a sense of agency over those actions.
+
+(see Carruthers 2012 for a summary of arguments against this model)
 
@@
 
-Our research builds on previous engineering work in brain-computer interfacing and on theories of the sense of agency as discussed in cognitive science and neuroscience.
+%Our research builds on previous engineering work in brain-computer interfacing and on theories of the sense of agency as discussed in cognitive science and neuroscience.
 
@@
 
-%%% theory on sense of agency 
-Agency is largely explained with a comparator model, describing internal comutational mechanisms of human action control 
+% Sense of agency in augmented interaction
+
+- intention / voluntary movement is important for sense of agency
+- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
+it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
+
+There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome (Buhner 2012)
+--> they suggest intentional action is not necessary for temporal binding, but that the binding results from a causal relation linking actions with consequences -- more general causal binding -- do we need to go there?
+
+One idea: - reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
+- maintain sense of agency in augmented interactions 
+- doing this by leveraging action intention in brain
+
+Experiencing control is an important factor in human-computer interaction (see Shneiderman and Nielsen).
+
+%%%%%%%%%% theory on sense of agency  / neural basis of sense of agency
+
+The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low-level pre-reflective sensory process, and the \textit{judgement} of agency, a higher level reflective cognitive process, ~\cite{Moore2016-ub, Danry2022-xk}.
+
+Agency is largely explained with a comparator model, describing internal computational mechanisms of human action control 
 -- comparator model (Blakemore 2002, Frith 2000, Frith 2005)
@@
 
-\subsection{neural basis of sense of agency}
+- In the words of Harrard: making an association between action and outcome (Intention - Volition-Movement-Agency -Effect) --> find the source; is from yt video
 
-\end{comment}
+- Looking at the sense of agency from the perspective of the comparator model
+    - if prediction and sensory feedback match = we perceive sense of agency - not only for the big loop with sensory feedback (beep), but also for the forward loop (smaller) --> I can still think that I turned the light on, even if it takes 10 sec after pushing the switch, cause movement still worked as predicted
 
-The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low-level pre-reflective sensory process, and a more higher level reflective cognitive process, the \textit{judgement} of agency~\cite{Moore2016-ub, Danry2022-xk}.
+- In Augmented movements different (explain)
+- Before deep diving in these results - excurse on how to measure agency
 
 \subsection{Measuring agency experience}
-% LT to write first version
-\begin{comment}
 
@@
 
+Researchers use implicit and explicit methods.
+Explicit methods involve directly asking participants to report their subjective experiences using items rated on Likert scales or percentages.
+These items, such as "It felt like I was in control of the hand I was looking at" (Longo & Haggard) and "Indicate how much it felt like moving the joystick caused the object on the computer screen to move" (Ebert & Wegner), may focus on either the action element or the outcome element of the sense of agency (see Moore).
+
+( van der Wel et al., 2012; Chambon et al., 2013),“ 
 Individuals’ sense of agency can be estimated with explicit self-reported measures using Likert’s scales or percentages ( van der Wel et al., 2012; Chambon et al., 2013),“
+
+Items such as ""It felt like I was in control of the hand I was looking at” (longo Haggard) and "Indicate how much it felt like moving the joystick caused the object on the computer screen to move” (Ebert und Wegner).
+Notably, such items differ between focusing either on the action element or the outcome element (See MOore)
 
@@
 - While details about the relationship between measures of intentional binding and SoA are far from clear, they are extensivly used 
+- alternatives to intentional binding - sensory attenuation / visual attention
 
-- SoA is influeced by different processes, 
-    - one of the is the source of action selection (self selected vs. externally instructed) (Chambon et al. 2014, Harrad 2017)
-    - another: motor control processes (read Sidarus 2017a)
-    -
+\textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.
+- comparisons between active and passive finger tab 
+    Kühn 2013 + Engbert 2007 : experimenter presses the finger down in a passive condition
 
--
+When a voluntary action is causally linked with a sensory outcome, the action and its consequent effect are perceived as being closer together in time.
+This effect is called intentional binding.“ (from Jo 2014)
 
@@
 - add figuure action binding and effect binding (e.g.  in WEN and Imamizu)
--
 
-\textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.
-When a voluntary action is causally linked with a sensory outcome, the action and its consequent effect are perceived as being closer together in time.
-This effect is called intentional binding.“ (from Jo 2014)
+!! new paper suggesting that temporal binding / intentional binding is not a measure of sense of agency !!(Gutzeit!)
 
-% what does that have to do with agency?
+% connection between implicit and explicit measures
 
-- intentional binding effects are are sometimes descrepant from explicit judgment of agency (see 30-33 in Wen adn Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
-- Intentional binding is e.g. influenced by causal relations between events, attention, arousal (see Wen and Imamizu)
+- intentional binding effects are sometimes discrepant from explicit judgment of agency (see 30-33 in Wen and Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
+- Intentional binding is e.g. influenced by causal relations between events, attention, and arousal (see Wen and Imamizu)
 
-- EEG patterns related to readiness potential and intentional  binding sometimes show disceprant patterns ( see WEN and Imamizu 38 und 39) (wittmann 2014, Goldberg 2017)
+% Sense of agency in augmented interaction
 
-- alternatives to intentional binding - sensory attenuation / visual attention
+- intention / voluntary movement is important for sense of agency
+- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
+it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
 
-% RESEARCH GAP
+There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome (Buhner 2012)
+--> they suggest intentional action is not necessary for temporal binding, but that the binding results from a causal relation linking actions with consequences -- more general causal binding -- do we need to go there?
 
-- reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
+One idea: - reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
 - maintain sense of agency in augmented interactions 
@@
 
-----
-involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
-it has been shown, that mere peripheral body movements, elicted by TMS simulations, produce a perceptual repulsion oppisite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
-
-There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome
-
--
+% RESEARCH GAP
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+Other
+- EEG patterns related to readiness potential and intentional  binding sometimes show disceprant patterns ( see WEN and Imamizu 38 und 39) (wittmann 2014, Goldberg 2017)
 \end{comment}

### 2023-07-26 input/2_introduction.tex

@@
 
-Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
+Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as if they randomly happen to us.
 It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}.
-The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
+The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' when using such an augemntation system.
 % if something happens for you in an interaction there is a cost to that:
@@
 % What have others done to address this and what do we propose "instead"
-Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimuli on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency.
+Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimulus on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency.
 The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited.
@@
 Crucially, our brain-computer interface controls the user's muscles through EMS at the time of their \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
-In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
+In our user study, we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
 % We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

### 2023-07-26 input/6_results.tex

--- 2022-09-04 input/5_results.tex
@@
 
-\subsection{Human-Computer Collaboration for Human Augmentation}
-\missingfigure{A: Boxplot of questionnaire scores, are they significantly better than 0?
-Are they significantly better then after the training?
--> frame the training the same as the experiment, B: Sentiment Word Clouds}
-
-% system usability score
+\subsection{Agency}
+\missingfigure{A: Boxplot of questionnaire scores of three Blocks (block 2 grey, because not intention given) - play around with visualization of words that were used to explain control rating.
+B.
+Boxplots intentional binding (maybe integrate A and B) C.
+more subjective evaluation of interview / maybe world cloud / maybe sentiment analysis) }
 
@@
 
-\subsection{The BCI}
+\subsection{BCI performance}
 

### 2023-07-26 input/3_related_work.tex
figure

@@
 \section{Related Work}
-%Intro
+%intro
 In a simplified world, intentional movements directly leading to external events provide a strong sense of responsibility and control over the action's outcome.
@@
 
-% Why is feeling of control important
+% why is feeling of control important
 Because the feeling of control is related to (check for sources on learning effects etc.), understanding the reason for its loss in augmented interactions is relevant.
 
-% Theorie
-
-The concept of eliciting a sense of agency is commonly explained within the framework of motor control (see Frith and collegus 2000).
-When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (the efference copy).
+%theorie 
+Sense of agency is simplictily explained within the framework of motor control (see Frith and collegus 2000).
+When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (i.e., efference copy).
 However, in situations where the movement is elicited by an external cause, as in augmented actions, the predicted sensory consequences may differ or be absent altogether.
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[]{figures}
+   \caption{}
+  \Description{
+}
+  \label{fig:com_model}
+\end{figure}
 (see Carruthers 2012 for a summary of arguments against this model)

### 2023-07-28 input/3_related_work.tex
Updates from Overleaf

@@
 \section{Related Work}
-%intro
+
+%Theories of sense of agency + example
+The most widely used theory on the sense of agency is the comparator model.
+It states that our brain when we perform a voluntary, intentional action, generates a sensory prediction, which is compared to the actual sensory feedback received from the sensory system.
+If the comparison reveals no sensorimotor incongruency, a sense of agency arises (sources).
+Imagine pressing the key of a piano to produce a tone.
+In such simple interaction, it is unequivocally clear that we are the ones performing the movement and causing the tone.
+We experience authorship.
+
+This straightforward relationship is complicated by the introduction of technology.
+Think about, playing a computer game where a key press controls an avatar playing the piano.
+Here, we do not really press the piano key, the avatar's movement, and the tone are now generated by a computer.
+Despite the blurring lines between the performer and the cause of the tone, we still feel a sense of authorship when the comparison does not reveal incongruency.
+This lingering feeling of control is thought to be elicited by our intentions being reflected in the resulting action (sources needed).
+
+In situations without intentional action, our brain does not generate predictions.
+Thus, in alignment with the cooperator model, researchers generally measured a lower sense of agency when actions were externally triggered (e.g., by brain simulation (Haggard and Clark, 2002), the body part being moved by another person (Kühn 2013).
+Accordingly, playing a tune on a piano with technology causes our figure movement, our body performs the action but we do not have the intention to play.
+We will most likely will lack a sense of agency.
+
+However, action augmentation technology using motor actuation aims to augment the user’s actions diminishing the SoA (see Cornelio 2022).
+
+%old
+- %intro
 In a simplified world, intentional movements directly leading to external events provide a strong sense of responsibility and control over the action's outcome.
-For instance, when we press a piano key, it is unequivocally clear that we are the ones performing and causing the action.
+For instance, when we press a piano key, it is unequivocally clear that we are the ones performing the movement and causing the tone.
 However, the introduction of technology can complicate this relationship.
 Imagine playing a computer game where a key press controls an avatar playing the piano.
-The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the action.
-Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of control.
+The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the tone.
+Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of authorship over the tone as a consequence of our action.
+This lingering sense of control might be due to our intentions being reflected in the resulting action (sources needed).
 
-This lingering sense of control might be due to our intentions being reflected in the resulting action (sources needed).
-However, in augmented interactions, the dynamics change.
+In augmented interactions, the dynamics change.
 When technology moves our fingers, our body performs the action, but we may experience a diminished feeling of causing the action [sources - augmented actions lead to less agency; check Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009, Haggard, Clark 2002].
 
-% why is feeling of control important
-Because the feeling of control is related to (check for sources on learning effects etc.), understanding the reason for its loss in augmented interactions is relevant.
+- no intentional binding when action is triggerd by brain stimulation  (Haggard, P., Clark, S. & Kalogeras, J. Voluntary action and conscious awareness. Nat. Neurosci. 5, 382–385 (2002).
+- 
+The feeling of agency seems to be
 
-%theorie 
-Sense of agency is simplictily explained within the framework of motor control (see Frith and collegus 2000).
+However, now  Imagine playing a computer game where a key press controls an avatar playing the piano.
+The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the tone.
+Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of authorship over the tone as a consequence of our action.
+
+%old 
+A sense of agency is the feeling of intentional actions and their external consequences (Haggard, 2017).
+It is implicitly explained within the framework of motor control (see Frith and collegus 2000).
 When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (i.e., efference copy).
@@
 According to this account, the feeling of agency arises only when the predicted sensory feedback perfectly corresponds to the actual sensory feedback.
-In other words, when our actions and their outcomes align seamlessly, we experience a sense of agency over those actions.
+In other words, when our actions and their outcomes align seamlessly, we experience a sense of agency over those actions (see figure \ref{fig:com_model}.
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-  \includegraphics[]{figures}
-   \caption{}
-  \Description{
-}
+    \includegraphics[width=\columnwidth]{figures/augmented_com_model_draft.png}
+    \caption{draft - visualization of comparator model of the sense of agency in the case of augmented interactions (inspired by Dewey 2019)}
     \label{fig:com_model}
@@
 (see Carruthers 2012 for a summary of arguments against this model)
+
+% relevance of sense of agency
+- motor learning ( check Wen, W. et al. Perception and control: individual difference in the sense of agency is associated with learnability in sensorimotor adaptation. Sci. Rep. 11 , 20542 (2021).)
 

### 2023-07-28 input/7_discussion.tex
Updates from Overleaf

@@
 %- not only physical integration but also other adpative interfaces, or control of mobile phone (see Cornelio2022 review); % say here somewhere what action augmentation is and how it differs from body and environment/outcome augmentation
+
+% playful experiences: EMS
+
+\begin{comment}
+% critics on the intentional binding as a measure of agency
+- Recently, the adequacy of the intentional binding paradigm as a measure of agency is questioned.
+- see Suzuki 2019, Gutzeit 2023, Hoerl 2020
+
+% snippets that might be useful
+- a sense of agency is not an infallible reproduction of objective reality / can be quite divorced from objective reality (see Moore 2016) // --> if people still report agency, in no-intent (only EMS) condition
+
+%future work / relevance of work
+- relevance of research for prostheses ( see https://link.springer.com/chapter/10.1007/978-3-030-38740-2_8)
+\end{comment}

### 2023-09-15 input/2_introduction.tex
pre-submission text

--- 2023-06-19 input/2_introduction.tex
@@
 % Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
+Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw}.
+However, augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
+
 % 2. What is the complication, challenge identified
 % Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
-\todo{add something even more general here: e.g. increase human wellbeing by XY}
-Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}.
-There are two main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e.
-exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips.
-Another technology makes the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS).
-However, augmented users often report dissociative experiences, frequently disrupting their sense of agency~\citep{Gilbert2017-ze, Gilbert2019-uc}.
-
-Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us.
-It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}.
-The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
-% if something happens for you in an interaction there is a cost to that:
-% - impacts sense of agency which in turn may makes user's engage less -> citation?
-% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?
+Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us.
+It has frequently been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}.
+A key challenge to drive the adoption of human action augmentation then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
 
@@
 % What have others done to address this and what do we propose "instead"
-Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimuli on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency.
-The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited.
+One goal of user augmentation is to increase reaction times, thereby making an interaction more efficient.
+Previous work on such augmentation scenarios has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk}.
+Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
+Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.
+
+In experimental settings, this is typically achieved by using strongly controlled `stimulus-response' paradigms.
+For example, reaction times when tapping on a touchscreen in response to a presented stimulus on screen, can be reduced by an augmentation device while maintaining SoA for the user~\cite{Kasahara2019-sk, Kasahara2021-gy}.
+In such controlled scenarios, the timing of the action augmentation can be tuned to be near optimal because the user behavior can be predicted with very high certainty to follow a previously presented stimulus.
+However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
+In other words, how can we design a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions?
+
+% [4. (short) Answer Teaser
+% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
+% keep this very short]
+
+\subsection{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
+% 4. (long) Anwser with teaser image
+% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
+In this paper, we present a bleeding edge prototype to challenge the disruption in experiencing agency during action augmentation.
+We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector.
+The  INTENTION system controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Nguyen2023-me, Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+
+% TODO merge this with the figure below
+% \missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.
+<- better use this for the teaser image.
+Here better present the main results, i.e.
+SoA scores?
+-> Klaus comment: yes, but they don't have to be exclusive.
+The general setup with the LRP could appear with the main results both in the teaser or here?
+But the teaser would be more a visualization of the principle, no?
+That does not necessarily require real results but could show a hyperplane and data dots for examplification.}
+% make this as an infographic with comic style drawing of augmented user and then the three pathways drawn
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\columnwidth]{figures/setup_conditions_draft_small.png}
+%     \caption{draft apperatus and conditions}
+%     \label{fig:setup}
+% \end{figure}
+% \begin{figure}[!h]
+%     \centering
+%     \includegraphics[width=\columnwidth]{figures/augmented_com_model_colour.png}
+%     \caption{draft - visualization of comparator model of the SoA in the case of augmented interactions (inspired by Dewey 2019)}
+%     \label{fig:com_model}
+% \end{figure}
+
+To control an interaction in real-time, the EEG-based RP classifier was used to discriminate between two user states: either the user was predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or the user was predicted to be in a \textit{pre-movement} state, indicating the presence of an intent to act.
+When the system predicted a \textit{pre-movement} state, the user's action was augmented, potentially even before their own voluntary motor command.
+In our user study, we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their SoA.
+
+%%% Resources
+% add results and general outcomes here?
+
+% for now moved to introduction section
+% We follow classical work in the neuroscience describing SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay and reproduce it as such.
+Critically, this temporal compression only occurs following endogenous movement intention.
+The action outcome is mentally \textit{binded} to the intention.
+To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay <REF>.
+
+% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.
+
+% By controlling the stimulus presentation, the experimenter can control the timing of the \textit{pre-emptive} gain of the user's action augmentation.
+
+% Resources
+% The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited.
 However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy.
-A key question then remains: How can we maintain agency during physical action augmentation without controlling the environment?
-In other words, how can we design a closed-loop system to deliver an integrated, \textit{natural}, experience for users' augmented actions?
+
 % At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.
 
-\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.
-<- better use this for the teaser image.
-Here better present the main results, i.e.
-sense of agency scores?}
-\input{input/1_teaser_figure}
-
-% 4. (short) Answer Teaser
-% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
-% keep this short
-In this paper, we present a fast communication channel between a user's brain signals and a physical end effector, here EMS.
-Crucially, our brain-computer interface controls the user's muscles through EMS at the time of their \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG).
-In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
-% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.
-
-% \subsection{Preserving Agency using Brain Signals of the Intent to (Inter)act}
-% 4. (long) Anwser with teaser image
-% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
+% if something happens for you in an interaction there is a cost to that:
+% - impacts SoA which in turn may makes user's engage less -> citation?
+% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

### 2023-09-15 input/3_related_work.tex
pre-submission text

--- 2023-07-26 input/3_related_work.tex
@@
 \section{Related Work}
-%Intro
-In a simplified world, intentional movements directly leading to external events provide a strong sense of responsibility and control over the action's outcome.
-For instance, when we press a piano key, it is unequivocally clear that we are the ones performing and causing the action.
-However, the introduction of technology can complicate this relationship.
-Imagine playing a computer game where a key press controls an avatar playing the piano.
-The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the action.
-Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of control.
+Our research draws inspiration from neuroscience, especially in assessing the SoA qualitatively and quantitatively, and from engineering work on BCIs as well as on physical user augmentation.
 
-This lingering sense of control might be due to our intentions being reflected in the resulting action (sources needed).
-However, in augmented interactions, the dynamics change.
-When technology moves our fingers, our body performs the action, but we may experience a diminished feeling of causing the action [sources - augmented actions lead to less agency; check Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009, Haggard, Clark 2002].
+\subsection{Theories of Sense of Agency}
+The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: when we intentionally perform an action, the brain generates sensory predictions about the action outcome.
+These predictions are constantly compared to the actual sensory data available through the sensory system during the execution of the action.
+These include continuous signals such as proprioceptive and visual monitoring of the ongoing movement as well as higher level predictions about the semantic outcome of the action~\cite{Clark2013-ah, Haggard2003-ff, Haggard2017-uv}.
+If no sensorimotor incongruency arises, a SoA manifests.
 
-% Why is feeling of control important
-Because the feeling of control is related to (check for sources on learning effects etc.), understanding the reason for its loss in augmented interactions is relevant.
+In the simple case of pressing a key on a piano, the finger movement is constantly compared to the predicted proprioceptive feedback.
+Subsequently, the tone generated by the key press is evaluated against auditory predictions.
+On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key~\cite{Pangratz2023-ew}.
+If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced (see figure~\ref{fig:task_design} BASELINE).
 
-% Theorie
+In human-computer interaction (HCI) research, these constructs are categorized in their own terms.
+Often, the term \textit{pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements.
+At higher levels of the cognitive hierarchy, \textit{reflective}, i.e.
+conscious, experience is used to refer to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.
 
-The concept of eliciting a sense of agency is commonly explained within the framework of motor control (see Frith and collegus 2000).
-When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (the efference copy).
-However, in situations where the movement is elicited by an external cause, as in augmented actions, the predicted sensory consequences may differ or be absent altogether.
-In such cases, the actual sensory feedback does not match the predicted sensory feedback.
+As opposed to intended actions using our own body, movement augmentation hardware allows moving a user's body without their intention.
+Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e.
+exoskeletons, a user's body can be moved by applying forces to the extremities.
+Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS).
+Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle activating nerves.
+In these scenarios, the user computes no predictions about the movement and its outcome.
+Thus, in reference to the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
+With something or somebody else moving our body to perform, e.g., a keypress on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
+Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz} (see figure~\ref{fig:task_design} NO INTENTION)
 
-According to this account, the feeling of agency arises only when the predicted sensory feedback perfectly corresponds to the actual sensory feedback.
-In other words, when our actions and their outcomes align seamlessly, we experience a sense of agency over those actions.
+\subsubsection{Measuring Sense of Agency}
+Both explicit and implicit methods have been developed to evaluate the sense of agency.
+These methods provide the basis to investigate the effects of action augmentation technology on agency experience.
+Explicit methods directly query participants to report their subjective experiences using questionnaires.
+Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}.
+However, in most cases such questionnaires aim at a higher-level, reflective, judgment of agency.
+
+On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously reflected~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
+Following seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}.
+Critically, this temporal compression only occurs following movement intention.
+The action outcome is mentally \textit{bound} to the intention.
+To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay~\cite{Barlas2018-bq}.
+Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.
+
+% alternative subsection heading: Brain Signals Reflecting the Intent to (Inter-)Act
+% this must make the link clear that we need intention for agency experience and intention is something we can set out to measure using EEG!!!
+\subsection{Controlling Actuated Haptic Experiences using Brain Signals Reflecting the Intent to (Inter-)act}
+Of the possible physiological signals that can be used to further investigate the experience of agency, the electroencephalogram (EEG) is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
+The Readiness Potential (RP), or \textit{lateralized readiness potential} is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
+The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand.
+Since its measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
+However, evidence abounds for its role in action preparation.
+An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement.
+The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after.
+The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}.
+
+A recent study has shown that the RP is ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}.
+In their study,~\citet{Schultze-Kraft2021-cu} asked participants to find a way to perform voluntary movements while keeping accompanying RP amplitudes as small as possible.
+After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for.
+They found participants unable to suppress their RP.
+This inability to suppress the RP renders it a reliable feature for classification.
+For example, the RP can be detected in real-time using a brain-computer interface (BCI).~\citet{Schultze-Kraft2016-bx} demonstrated a prototype that detects RPs in participants ongoing EEG data and adapts an interface accordingly.
+Participants were instructed to veto their self-initiated movement whenever a red dot occurred on the screen.
+The red dot's appearance was controlled by the BCI.
+Whenever an RP was detected, the red dot appeared.
+The authors found that participants were able to veto their self-initiated movement if the red dot appeared no later than 200ms preceding their movement onset.
+After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot appearance on screen.
+
+In the present study, we leveraged the RP in a similar way, i.e.
+to discriminate between two \textit{action} states as the basis to trigger an action augmentation device or not.
+The technology we used with our prototype, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.
+
+%%% resources
+
+\begin{comment}
 
 (see Carruthers 2012 for a summary of arguments against this model)
-
-% In this paper, we present a physical action augmentation prototype and a user study to investigate the experience of agency during the use of the prototype.
-Hence, our research builds on previous qualitative and quantitative research on the sense of agency.
-% on actuated haptic systems, brain-computer interfacing as well as
-% moving the finger of a user when the user intends to do so, but before the user directs their muscles to move.
-
-%Our research builds on previous engineering work in brain-computer interfacing and on theories of the sense of agency as discussed in cognitive science and neuroscience.
-
-\subsection{The sense of agency in human-computer interaction}
-% LT to write first version
-\begin{comment}
-
-% Sense of agency in augmented interaction
-
-- intention / voluntary movement is important for sense of agency
-- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
-it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
 
@@
 %%%%%%%%%% theory on sense of agency  / neural basis of sense of agency
-
-The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low-level pre-reflective sensory process, and the \textit{judgement} of agency, a higher level reflective cognitive process, ~\cite{Moore2016-ub, Danry2022-xk}.
-
-Agency is largely explained with a comparator model, describing internal computational mechanisms of human action control 
--- comparator model (Blakemore 2002, Frith 2000, Frith 2005)
--- integrate predictive coding??
-
-- In the words of Harrard: making an association between action and outcome (Intention - Volition-Movement-Agency -Effect) --> find the source; is from yt video
-
-- Looking at the sense of agency from the perspective of the comparator model
-    - if prediction and sensory feedback match = we perceive sense of agency - not only for the big loop with sensory feedback (beep), but also for the forward loop (smaller) --> I can still think that I turned the light on, even if it takes 10 sec after pushing the switch, cause movement still worked as predicted
-
-- In Augmented movements different (explain)
-- Before deep diving in these results - excurse on how to measure agency
-
-\subsection{Measuring agency experience}
 
@@
 These items, such as "It felt like I was in control of the hand I was looking at" (Longo & Haggard) and "Indicate how much it felt like moving the joystick caused the object on the computer screen to move" (Ebert & Wegner), may focus on either the action element or the outcome element of the sense of agency (see Moore).
-
-( van der Wel et al., 2012; Chambon et al., 2013),“ 
-Individuals’ sense of agency can be estimated with explicit self-reported measures using Likert’s scales or percentages ( van der Wel et al., 2012; Chambon et al., 2013),“
 
@@
 - alternatives to intentional binding - sensory attenuation / visual attention
+- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
+it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
 
@@
 
-% connection between implicit and explicit measures
-
+% connection between implicit and explicit measures:
 - intentional binding effects are sometimes discrepant from explicit judgment of agency (see 30-33 in Wen and Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
@@
 
-% Sense of agency in augmented interaction
+% Interesting HCI papers:
+% PosssessedHand, Gilbert-I miss being me,
 
-- intention / voluntary movement is important for sense of agency
-- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
-it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
-
-There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome (Buhner 2012)
---> they suggest intentional action is not necessary for temporal binding, but that the binding results from a causal relation linking actions with consequences -- more general causal binding -- do we need to go there?
-
-One idea: - reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
-- maintain sense of agency in augmented interactions 
-- doing this by leveraging action intention in brain
-
-% RESEARCH GAP
-
-\subsection{Sense of agency and actuated haptic systems}
-
-\ subsection{Movement Intention}
-idea: when we now compare completely voluntary movements with movements triggerd by the brain signal of our intention to move - what happens?
-
-% close this paragraph with the problem that of the control mechanism of the actuator hardware and the disruotion in agency
-
-% PosssessedHand, Gilbert-I miss being me,
+% Why is sense of agency important? Why would it be bad if we lost it in augmented humans?
+% However, for augmentation hardware to be readily accepted, maintaining a sense of agency is important for several reasons. (find reasons)
+% - motor learning ( check Wen, W. et al. Perception and control: individual difference in the sense of agency is associated with learnability in sensorimotor adaptation. Sci. Rep. 11 , 20542 (2021).
+% - motivation (find sources)
 
@@
 \end{comment}
+
+% This straightforward relationship is complicated by the introduction of technology.
+Think about, playing a computer game where a key press controls an avatar playing the piano.
+Here, we do not really press the piano key, the avatar's movement, and the tone are now generated by a computer.
+Despite the blurring lines between the performer and the cause of the tone, we still feel a sense of authorship when the comparison does not reveal incongruency.
+This lingering feeling of control is thought to be elicited by our intentions being reflected in the resulting action (sources needed).
+
+% alternative headings:
+% Controlling Actuated Haptic experiences using brain signals reflecting the intent to interact
+% Triggering EMS using EEG Signals

### 2023-09-15 input/5_user_study_and_methods.tex

@@
 \section{User study \& Methods}
-With this study we wanted to find out whether agency can be preserved during \textit{physical action augmentation} when the user's own brain signals serve as the control signal.
-To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, questionnaires as well as a qualitative interview.
-Our interviews followed a phenomenological structure~\citep{Danry2022-xk}.
+With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user's own brain signals serve as the control signal to a physical end effector.
+To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.
+% todo add some more framing here to make it easier to step through the methods, maybe: For this paper we designed a prototype that lies between the two edge cases -- no intention and control to full control -- and investigate user's sense of agency using both explicit and implicit measures.
 
-\subsubsection{Participants}
-\todo{add actual number} Ten participants (\todo{add mean age and sd}) from our local institution participated in our study.
-All participants were right handed, had normal or corrected to normal vision and had not experienced EMS before.
-Participants were compensated with 12 Euro per hour of study participation.
-Prior to their participation, participants were informed of the nature of the experiment, recording and anonymization procedures and signed a consent form.
-The experiment was approved by the ethics committee of [anonymized]
+\subsection{Participants}
+Nine participants (M = 30.00 years, SD = 3.81) were recruited from our local institution and through the institute's online participant pool.
+All participants were right-handed.
+Participants were compensated with course credit or 12 Euro per hour of study participation.
+Prior to their participation, they were informed of the nature of the experiment, recording, and anonymization procedures and signed a consent form.
+The experiment was approved by the ethics committee of [anonymized].
+One participant had to be excluded from further data analyses due to significant deviations from the instructions in the execution of the task.
 %the Department of Psychology and Ergonomics at TU Berlin (tracking number: BPN\_GEH\_2\_230130220421).
 
-% TODO continue here 19.05.2023
-\subsubsection{Apparatus}
-\todo{add the components that were used to run the STUDY, touchpad, keyboard etc.}
-The experimental setup, depicted in Figure~\ref{}, comprised: (1) a wrist-mounted wearable HTC VIVE tracker for motion capture, (2) a 64-channel EEG system, and (3) a medically-compliant EMS device connected via two electrodes worn on the forearm.
-To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code to the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
+\subsection{Experimental Task, Design and Procedure}
+Participants performed a simple tapping task in three conditions with 75 trials each: BASELINE, NO INTENTION (to move) and INTENTION (to move) EMS, see figure~\ref{fig:task_design}.
 
-% TODO change this to the use of EMG for now
-\indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz.
-Therefore, we ran a simple Unity3D scene and streamed the data via the labstreaminglatyer (LSL)\footnote{https://github.com/sccn/labstreaminglayer}.
+\indent\textbf{BASELINE.} The task went as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) then, they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen.
+In line with the literature on the origin of the RP generating process they were told ``to avoid pre-planning the movement, avoid any obvious rhythm over trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
+(4) After the screen was tapped, a tone was played at a pseudo-random delay of 200, 350, or 500ms.
+Participants were now asked to estimate the delay, typing in their answer on a number pad of an attached keyboard.
+After confirming their answer by hitting the return key, the next trial started.
+During BASELINE, participants were equipped with Electromyogram (EMG) sensors instead of EMS electrodes.
+BASELINE data was used as training data for the BCI, see section~\ref{BCI} below, as well as to select stereotypical reaction times in the NO INTENTION condition.
+
+\indent\textbf{NO INTENTION.} The task structure was identical to BASELINE, however, participants were now instructed to hold and wait for the muscle stimulation to move their finger thereby eliciting the screen tap.
+The timing for the EMS trigger was taken by randomly choosing a time between the 5th and 95th percentile of their actual individual reaction time in the BASELINE condition.
+
+\indent\textbf{INTENTION.} The task and instruction were identical to BASELINE with one additional instruction: ``you will now work \textit{with} the system''.
+During INTENTION, the muscle stimulation hardware was controlled by the BCI.
+The classifier was set to active after the fixation cross disappeared and until a screen tap was registered.
+Hence, the muscle was not stimulated at other times during a trial, so as not to interfere with participants typing in their time estimation response.
+
+The order of the conditions was not pseudo-randomized since training data obtained in BASELINE was required for both NO INTENTION and INTENTION.
+Furthermore, INTENTION was always the last condition, thereby allowing for a prolonged interview.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/task_design.png}
+    \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
+    \label{fig:task_design}
+\end{figure*}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/task_progression.png}
+    \caption{Visualization of task progression}
+    \label{fig:progression}
+\end{figure}
+
+\subsection{Apparatus}
+The experimental setup, depicted in Figure~\ref{}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected via two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
+To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
+% TODO add links after anonymization is lifted.
+
+\indent\textbf{(1) EMG Recording.} EMG data was recorded from 1 bipolar channel using the BrainAmp ExG amplifier (BrainProducts GmbH, Gilching, Germany).
+The two electrodes were placed above the flexor digitorum profundus with a reference electrode located on the wrist bone.
+EMG data was collected in synchrony with the EEG data through BrainProducts BrainVision Recorder.
 
 \indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
-Electrodes were placed according to the extended international 10–20 system \cite{Chatrian1985-ys}.
-One electrode was placed under the left eye to provide additional information about eye movements (vEOG).
-After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity and electrode impedance was brought below 10k$\Omega$ for all electrodes.
-EEG data was recorded with a sampling rate of 250 Hz.
-We used LSL to make the EEG data stream available in the network and synchronize the recordings of EEG data, motion capture and an experiment marker stream that marked sections of the study procedure.
+Electrodes were placed according to the extended international 10–20 system \cite{Jasper1983-uw}.
+One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10$\Omega$.
+EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
 
-\missingfigure{show participant with measurement system and description next to each element}
+We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the data streams available in the network and synchronize the recordings of EEG/EMG data and an experiment marker stream that marked sections of the study procedure.
 
-\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the index finger via electrical muscle stimulation (EMS), which was delivered via two electrodes attached to the participants' extensor digitorum muscle.
-We utilized the extensor digitorum since we found that we can robustly actuate it without inducing parasitical motion of neighboring muscles.
-This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (prorelax TENS/EMS Super Duo Plus).
-The EMS system's output was controlled by flipping a solid state relay connected via an Arduino Uno to the experiment computer.
-The EMS was pre-calibrated per participant to ensure a pain-free stimulation and robust actuation.
+\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
+We utilized the flexor digitorum profundus since we found that we can robustly actuate it without inducing unintended motion of neighboring fingers.
+This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (TENS/EMS Super Duo Plus, prorelax, Düren, Germany).
+The EMS system's output was controlled by flipping a solid state relay (silent) connected via an Arduino Uno (Arduino, Monza, Italy) to the experiment computer.
+The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
+To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure ~\ref{fig:task_design} c.
+% [todo] What were the settings of the stimulation?
 
-\subsubsection{Task}
-\missingfigure{add an infographic of the task progression with a time axis here, this should be covering two columns and be at the top of the page!}
-Participants performed a simple reaching task: resting their hand on the table, participants were instructed to press a specified key on a notebook in front of them.
-For each trial, a key was randomly selected from one of two groups of letters: (1) 'a', 's', 'd' or (2) 'j', 'k', 'l'.
-The selected group was then changed with every trial.
-This was done to keep the task engaging while maintaining a comparable reaching movement between trials.
+\indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
+In addition to the tablet, we used an external keyboard to allow response input of users indicating their timing judgements.
 
-The interaction flow for one trial of our task, depicted in Figure~\ref{task_flow}, was as follows: (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key.
-In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move".
-(4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone.
-(5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement.
-Following an inter-trial-interval the next trial started.
+% alternative headings:
+% Brain-Computer Interface to Detect EEG Readiness Potentials }
+% Classifiying EEG Signals of the Intent to Interact
+\subsection{Brain-Computer Interface}\label{BCI}
+The data obtained in BASELINE was used to train the BCI.
+We utilized the [anonymized] and the EEGLAB~\cite{Delorme2004-sn} toolbox inside the MATLAB (The MathWorks Inc.
+Natick, MA, USA) environment for preprocessing both EEG and EMG data.
+First, to generate behavioral labels at a high temporal resolution for the EEG-based classifier, we leveraged EMG data from the flexor digitorum profundus.
+EMG amplitudes were band-pass filtered from 20 to 100 Hz and subsequently squared.
+Next, in order to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
+From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset.
+% bemobil pipeline
 
-\subsubsection{Experiment Design and Procedure}
-Participants completed three experimental blocks. The first block consisted of 60 trials and was used to obtain training data for the single-trial classification system outlined below.
+% TODO Next, noisy data segments were determined. 
+% A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters.
+% from isoutlier documentation (mean segment amplitude exceeding 3 scaled median absolute deviation)
 
-%The condition of the second and third block, see below, were alternated between participants.
+Two event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the (EMG detected) movement onsets and \textit{idle}, a one-second data segment between trials where participants were looking at a fixation cross.
 
-In a second block, participants completed 80 trials of the same task and trial order as in the first block.
-Now, the brain-computer interface (BCI) detecting readiness potentials from real-time EEG data was continuously running in the background.
-After the target letter was shown to the participants and they were preparing to reach out and press the corresponding key, the BCI was live.
-Whenever the BCI detected EEG signals to reflect an interaction intent, EMS was activated causing the participants index finger to lift.
+\subsubsection{Preprocessing EEG and Selecting discriminative channels}
+The EEG data was band pass filtered from 0.1 to 15 Hz.
+In the first step to prepare the EEG data for classification, noisy data segments were rejected.
+To this end, the EEGLAB function `autorej' was used, keeping default parameters.
+A trial was excluded if \textit{either} data from the \textit{idle} or \textit{pre-movement} class was rejected.
 
-Both experimental blocks two and three were followed by a mixed methods data collection of participants' perceived agency.
+Then, to decrease the dimensionality of the EEG data, a selection of discriminative channels was conducted following a selection approach outlined by~\citep{Schultze-Kraft2021-cu}.
+First, for both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100 ms of the segment.
+The resulting value thus indicated how much the signal had changed in the course of the 1 s segments.
+In line with the literature, there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment.
+Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments.
+With respect to the ability of the system to discriminate between movement intention and the absence of movement intention, an ideal channel should show a strong change in signal over the course of a \textit{pre-movement} segment but no difference during an \textit{idle} segment.
+The ranks of the two criteria were joined by summation and the resulting order was saved for later use in training and real-time application of the classifier.
+Lastly, channels C3, C4 and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.
 
-In summary, the experiment consisted of four phases: (1) a setup phase; (2) a first experimental block: the training phase; (3) a second experiment block: the test phase; and (4) a test battery of questionnaire and interview to inquire about their experienced agency.
+\subsubsection{Training of EEG classifier}
+To classify EEG data a linear discriminant analysis (LDA) with shrinkage regularization (automatic shrinkage using the Ledoit-Wolf lemma~\cite{Ledoit2004-bi}) was trained for each participant individually.
+As single-trial features, the (linear) slope coefficient was obtained for both \textit{idle} and \textit{pre-movement} segments by fitting a linear regression.
+The classifier was then trained using the slope coefficient feature of the selection of the most discriminative channels as features, hence in the end, generating a feature vector in the dimension of channels that were kept for classification.
+We purposefully constrained the dimensionality of the feature vector to avoid over-fitting.
 
-\subsubsection{Measures of agency experience: Intentional Binding, Questionnaire \& Interview}
+Using scikit-learn~\cite{Pedregosa2012-sj}, an LDA with automatic shrinkage was cross-validated (using 5-folds) for a grid search from 6 to 20 channels with a step size of adding 2 channels.
+Following this grid search, the cross-validation with the highest accuracy determined the number of channels that were kept for training the model for real-time application.
+Ultimately, to determine the threshold at which, during real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
 
-To assess if agency was preserved, three measures were recorded during and after the test phase.
+\subsubsection{Real-time application and EMS control}
+During real-time application, the EEG data was buffered for the last second for the selected channels.
+The data was band-pass filtered analogously to the training data from 0.1 to 15 Hz.
+Next, the slope was computed for the discriminative channels selected in the classifier cross-validation grid search.
+This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier.
+To smooth the prediction output with the goal of reducing false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the last two predictions with a weighting (.3 and .5).
+Then, at 10Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5s.
 
-\indent\textbf{(1) Intentional Binding.}
-At the end of each trial participants were tasked to compare the duration of their arm movement to a tone and indicate whether the tone was longer or shorter.
-We set the initial duration at 500 ms for each participant.
-When participants indicated the tone to be longer or shorter by pressing either the up or down key we added, or respectively subtracted, 100 ms from the tone duration.
-This new duration was then used for the next trial and so on.
-With 60 trials in the training phase, we hypothesized participants would oscillate 50-50 around a target duration after about 2/3 of the training trials.
-The final duration after all training trials was then used again as the start duration during the test phase.
-In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS.
-A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.
+% To trigger the EMS, an Arduino was used to flip a switch whenever triggering the attached EMS.
+We constrained the times when the switch could be flipped to specific moments in the experiment where participants were getting ready to perform an action, thereby limiting the impact of distracting EMS pulses at resting phases during the experimental trials.
+%This guaranteed that participants experienced EMS only during movement preparation
 
-\indent\textbf{(2) Questionnaire.}
-In line with~\cite{Hornbaek} we prompted participants with three statements and a 7-point Likert scale centered at 0.
-The first one ``I felt like the device I was using was part of my body'' was in relation to body ownership.
-The following two statements ``It felt like I was in control of the movements during the task.'' and ``What is the degree of control you felt?'' prompted participants on their experienced agency.
+\subsection{Measures of Agency Experience: Intentional Binding, Question \& Interview}
+To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the movement INTENTION block were investigated.
+At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Hornbaek}.
+Following INTENTION we interviewed users about their experience working with the EMS BCI.
+After prompting users to recall their experience and summarize what their task had been we set the focus to the tapping movement and asked them to ignore the time estimation task for the following questions.
+We entered the open part of the interview by asking: ``What did the system do?'' followed up by ``What was the difference between the three conditions''.
+After some time, and depending on their answers, we reset the focus to the INTENTION condition and asked ``How often was the system active?'' followed by ``What do you think caused the actions of the system?''.
+This was then followed up by an `open' interview in which we frequently asked `how' and `why' questions to inquire about the user's experience.
 
-\indent\textbf{(3) Interview.}
-We interviewed user's about the experience with the EMS integration.
-We asked open how questions in a collaborative interview inquiring about the user's lived experience.
-After prompting user's to recall their experience we started the interview with: "how did it feel like?".
+% add back in after anonymization: All interviews were manually transcribed and then automatically translated using DeepL\footnote{https://www.deepl.com/translator}
+We analyzed the interviews that always followed the INTENTION condition by loosely following~\citet{Mayring2015-pp}.
+All interviews were manually transcribed and [anonymized].
+Prior to screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
+Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e.
+participants estimate of how well the system was aligned with their intention.
+For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into positive and negative valence sentiments.
+
+\subsection{Statistical Analyses}
+To confirm and demonstrate the discriminative power of the EEG features, we plotted the amplitude time course of electrode Cz between \textit{pre-move} and \textit{idle} epochs.
+Next, the slope coefficients were extracted and a paired t-test was conducted.
+Ultimately, to assess the classifier performance, we report the F1 score, i.e.
+the harmonic mean between of the precision and recall, and plot the ROC.
+
+\subsubsection{Hypotheses Testing}
+Prior to any statistical analyses of the intentional binding measure, outlier trials were rejected.
+We applied `extreme outlier removal' using Tukey's method~\cite{Tukey1949-sl}.
+Three time intervals were considered relevant to describe `regular' behavior across trials which would manifest by: (1) Tapping the screen in a reasonable interval after the fixation cross disappeared.
+An excessively short or long delay indicated that participants either tapped the screen prematurely by accident or they were checking in with the experimenter, respectively.
+(2) Providing a `reasonable' estimation in the intentional binding task.
+(3) The EMS stimulation leading to an \textit{immediate} screen tap.
+A long delay between the EMS trigger and the subsequent screen tap indicated that the stimulation was not strong enough in this trial to lead to muscle actuation resulting in a screen tap.
+Taken together, applying Tukey's methods to these criteria led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
+
+In line with the literature on intentional binding, we hypothesized that for both BASELINE and INTENTION conditions the time intervals should be underestimated.
+This should not be the case for the augmented NO INTENTION condition where users had no intention to move.
+Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
+To test this, we fit a linear mixed effects model with \textit{condition} (BASELINE, NO INTENTION, INTENTION) as a fixed effect.
+As a random effect, we considered the intercepts for participants.
+We obtained p-values of the coefficients by calculating likelihood-ratio tests~\cite{winterLinearModelsLinear2013}.
+All parameters were estimated by maximum likelihood estimation~\cite[see][]{Pinheiro2000}.
+Post-hoc, pairwise contrasts were computed using estimated marginal means with Tukey's methods for multiple comparisons corrections~\cite{Lenth2020-xk}.
+
+Next, we hypothesized that subjective ratings of control over the tap movement are comparable between BASELINE and INTENTION and lower in the NO INTENTION condition.
+Again, we fit a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
+Coefficients were assessed in the same way as for the intentional binding parameter above.
+
+In short, we tested two main hypotheses with regard to experiencing agency in our three experimental conditions: (1) participants underestimate the tone delay when acting intentionally.
+Adding EMS in line with participants' intention to move does not affect this underestimation.
+(2) The subjective feeling of control is comparable between BASELINE and INTENTION conditions.
+NO INTENTION should decrease the feeling of control significantly.
+Additionally, we report the clustered interviews anecdotally.
+
+%%% Resources
+% An additional cluster ``takeaways'' was used to cluster other, miscellaneous,  interesting comments.
+
+% "how did it feel like?".
 If user's made reference themselves to the EMS device we followed up by inquiring: "why do you think it (the EMS trigger) happened?", "What do you think triggered it?".
@@
 
-To not break the user's immersion by disturbing the flow of the interaction we decided against asking qualitative questions at the end of each and every trial.
+% \indent\textbf{(1) Intentional Binding.}
+% At the end of each trial participants were tasked to estimate the delay between the screen tap and a subsequent tone.
+The real delay was pseudo-randomized out of [200, 350 and 500] ms.
+With 75 trials per condition, each real delay occurred 25 times.
 
-\subsubsection{Statistical Analyses}
-\todo{how where the results analyzed: sentiment analyses and statistics of IB data and questionnaire results}
+% \indent\textbf{(2) Questionnaire.}
+
+% \subsection{Brain-Computer Interface}
+% The presented classification scheme required individual training data per participant, as transfer learning remains challenging with EEG~\cite{Wan2021-zz}.
+Therefore, we initially obtained training data on a variant of our task that did not include any muscle stimulation.
+In our task, participants were instructed, following an initial waiting period, to tap a touchscreen with their ring finger at their own volition, see figure ~\ref{fig:teaser}.
+% \subsubsection{Labelling movement classes using EMG}3
+
+% Subsequently the data were labelled into two classes: an \textit{idle} class representing resting background activity and a \textit{pre-movement} class, representing the intent to (inter-)act.
+
+% In summary, the experiment consisted of five phases: (1) a setup phase; (2) the BASELINE condition to obtain training data and labels; (3) the NO INTENTION condition: being passively moved by the EMS system; (4) the INTENTION EMS condition where the BCI was triggering the EMS system that in turn made participants tap the screen; and (5) an open interview to inquire about their experienced agency following the INTENTION EMS condition.
+
+% \subsubsection{Experiment Design and Procedure}
+% The first condition was used to obtain training data for the single-trial classification system.
+Here participants were wearing the EMG sensors instead of the EMS actuators and were instructed to tap the screen at their own volition (BASELINE, see instructions above).
+In the second condition, participants were instructed to hold and wait for the EMS system to move their finger thereby eliciting the screen tap (NO INTENTION).
+The timing for the EMS trigger was taken by randomly choosing a time between the 5th and 95th percentile of their actual reaction time in the BASELINE condition.
+In the third condition, the EMS trigger was controlled through the readiness potential classifier (INTENTION EMS).
+Participants were given the instruction: ``you will now work \textit{with} the system'' along the same instruction as in the BASELINE condition.
+The classifier was set to active after the fixation cross disappeared and until a screen tap was registered.
+Hence, the EMS was not triggered at other times during a trial, so as not to interfere with participants typing in their time estimation response.
+
+% To not break the user's immersion by disturbing the flow of the interaction we decided against asking qualitative questions at the end of each and every trial.
+
+% old:
+% \subsubsection{Reproducing Results and Data Availability}
+% Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}.
+BIDS formatted data is hosted on openneuro~\cite{}.
+
+% , and the interval between EMS stimulation and tap as fixed effects
+
+% Hypothese:
+% In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS.
+A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.
+
+% operationalising temporal binding = underestimation (here or somewhere else)
+
+% \indent\textbf{(1) Hand Tracking.} We used an HTC Vive Tracker, attached to the participant's wrist, to track their right hand at 90Hz.
+Therefore, we ran a simple Unity3D scene and streamed the data via the labstreaminglatyer (LSL)\footnote{https://github.com/sccn/labstreaminglayer}
+
+% At the end of each trial participants were tasked to compare the duration of their arm movement to a tone and indicate whether the tone was longer or shorter.
+We set the initial duration at 500 ms for each participant.
+When participants indicated the tone to be longer or shorter by pressing either the up or down key we added, or respectively subtracted, 100 ms from the tone duration.
+This new duration was then used for the next trial and so on.
+With 60 trials in the training phase, we hypothesized participants would oscillate 50-50 around a target duration after about 2/3 of the training trials.
+The final duration after all training trials was then used again as the start duration during the test phase.
+In line with the literature on intentional binding, we hypothesized this duration to remain stable in the case that agency was preserved during the test phase with EMS.
+A deviation of the tone duration from the training phase was hypothesized to reflect a disruption in experienced agency.
+
+% (1) participants waited for a new letter to appear on screen; (2) then, they were instructed to wait for a brief moment (~2s), before (3) reaching out and pressing the key.
+In line with the literature on the origin of the RP generating process they were told "to avoid preplanning the movement, avoid any obvious rhythm, and to press when they felt the spontaneous urge to move".
+(4) Two \todo{how many, but do keep this constant in order for them to focus on the duration of their action} seconds after they pressed the key, they were presented with a tone and asked to judge whether their movement, from the start of their reach to the button press, lasted longer or shorter than the tone.
+(5) When they perceived the tone to be longer, they had to press the up key or vice versa the down key for perceiving that the tone was shorter than their movement.
+Following an inter-trial-interval the next trial started.
+
+% For each trial, a key was randomly selected from one of two groups of letters: (1) 'a', 's', 'd' or (2) 'j', 'k', 'l'.
+The selected group was then changed with every trial.
+This was done to keep the task engaging while maintaining a comparable reaching movement between trials.
 
@@
 Not possible to evlaute false positives as no ground truth available
-
-\subsubsection{Reproducing Results and Data Availability}
-Data, experimental protocol, analyses code including scripts for a reproduction of the presented results are hosted at open science foundation (OSF)\footnote{[anonymized]}.
-BIDS formatted data is hosted on openneuro \cite{}.

### 2023-09-15 input/6_results.tex

--- 2022-09-04 input/5_results.tex
@@
 \section{Results}
+In the intentional binding task, i.e.
+the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (BASELINE M = 160.3, SD = 116; NO INTENTION M = 176.7, SD = 120.7, INTENTION M = 162.6, SD = 118.9).
+The condition influenced the estimation (${\chi^{2}_{(2)}} = 15.9, p < .001$).
+In NO INTENTION the estimation was higher compared to BASELINE (beta = 19.2, p < 0.001).
+Also in NO INTENTION, the estimation was higher compared to INTENTION (beta = 18.6, p < 0.001).
+No difference was observed between BASELINE and INTENTION.
 
-\subsection{Human-Computer Collaboration for Human Augmentation}
-\missingfigure{A: Boxplot of questionnaire scores, are they significantly better than 0?
-Are they significantly better then after the training?
--> frame the training the same as the experiment, B: Sentiment Word Clouds}
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/behavioral_results.png}
+    \caption{Box-Whisker plots visualizing (a) the temporal binding effect over conditions.
+With an average presented time interval of 350ms, lower values in the estimated time interval indicate an underestimation, i.e., temporal binding.
+(b) visualizes the subjective ratings of control across conditions.
+Significance reportings refer to post-hocs test that followed a significant likelihood ratio test of the full model with condition as a factor against the model without it.}
+    \label{fig:results}
+\end{figure}
 
-% system usability score
+The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 35.9, p < .001$).
+Post-hoc pairwise comparisons revealed that participants rated their motion control higher in BASELINE compared to NO INTENTION (beta = 4.9 , p < .0001) and BASELINE compared to INTENTION (beta = 3, p < .0001).
+Further, higher control was observed in INTENTION compared to NO INTENTION (beta = 1.9, p = .006).
 
-% sentiment analysis ttest the counts?
+\subsection{Classifier Performance}
+Visual inspection of the amplitudes at electrode Cz revealed an increase in the difference between \textit{pre-movement} and \textit{idle} data segments towards the onset of the finger movement, see figure~\ref{fig:EEG_results} a.
+The slope feature for the exemplary channel discriminates well between the two classes (${t_{(8)}} = 3.6, p = .003$), see figure ~\ref{fig:EEG_results} b bottom.
+The scalp maps in figure~\ref{fig:EEG_results} b top show the (color-coded) mean slope for each channel and for each class.
+Central channels on the contralateral side to the moving finger on the right hand show a negative slope for the pre-movement class and a neutral slope for the idle class.
+Furthermore, differences in slope can also be observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifests only for the idle class.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/EEG_results.pdf}
+    \caption{![todo]: (a) depicts the grand average ERP at Cz comparing idle vs.
+movement intention with scalp map of voltage across all channels in the last [-100 0] interval of both idle and pre-move; B: What did the windows mean features look like.
+Boxplot with idle vs.
+pre-move and single observations, ROC -> in discussion mention that we want to do that continuously and not only look at the times preceding hand movement but at all time points}
+    \label{fig:EEG_results}
+\end{figure*}
+
+The grid-search over channels resulted in the BCI leveraging on average 11.5 (SD = 3.7) channels.
+Classifier cross-validation resulted in a mean F1 score of .7 (SD = .03), see figure~\ref{fig:EEG_results} c.
+We set the detection threshold to 15\% false positive rate and at that rate, observed a mean threshold of 57\% (SD = .02).
+Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57\% probability.
+% TODO add percentage of trials the prototype was activated before touching the screen (SD = ?)
+
+\subsection{Subjective Reports}
+Six out of the eight participants made a reference to the source of the EMS in the INTENTION condition.
+In that subgroup, most participants wondered how the system worked, one remarked, ``I can't explain how it works technically.''.
+They found the stimulation to be sporadic, even during their own actions, leading to a belief that it was coincidental.
+As one participant put it, ``I have no idea what controlled the stimulation, I think it was coincidence when stimulation occurred.''.
+Furthermore, participants reported that they found the timing of the stimulation to be unpredictable, with one participant noting, ``Stimulation was random in time.''.
+Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly involuntary, but partly as if it was following my decision / choice.''.
+One participant briefly considered the involvement of their brain waves in controlling the stimulation, but expressing doubts about this possibility, stating, ``But I don't think that is the case.''.
+
+Some participants reported that the stimulation ``sometimes overlapped with the movement, but not often,'' or, it ``came only rarely.'' Another one noted, ``In a few cases, the stimulation came when I had already started the movement.'' For several participants, this overlap between their actions and the system's response occurred infrequently, with another user noting, ``Once, in the millisecond range between my planned movement and its execution.'' However, for some participants, there were moments of near-perfect synchronization, as described by one participant, ``In 3 cases it happened that my intention to press and the impulse of the device happened simultaneously.'' Another one noted, ``Sometimes it really happened that they overlapped.
+So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40\%'' while another said [we worked] ``15\% together.'' These accounts collectively highlight the varied and occasionally synchronous nature of the  INTENTION system's timing in relation to the users' intention.
+
+In terms of valence sentiments for the cases where the stimulation aligned with participants' intentions, some participants indicated that the experience was positive, for example, participants described the experience as ``weird but funny'', ``pleasant'' or ``helped me with the execution'' and ``it was more of a collective movement.'' One participant remarked ``then it was ok to experience the stimulation, but also not more than ok'' while another one noted ``It had a bit of thinking ahead to it''.
+Some noted that they experienced an increase in their physical strength, remarking, ``supported my strength'', ``made me type more firmly'' and ``my typing performance was increased.'' On the other hand, some participants had negative sentiments during these moments of aligned stimulation, remarking, ``felt like I was still in competition with the system'', ``it did not feel like a acting together'' and ``on a psychological level it was a loss of control.''
+
+%%% Resources and Ideas
+% and the agency condition by xx (CI: ). Subsequent non-parametric comparisons revealed the significance of differences only between  
+% xx (p = ).
+% x\
+
+%\missingfigure{A: Boxplot of questionnaire scores of three Blocks (block 2 grey, because not intention given) - play around with visualization of words that were used to explain control rating.
+B.
+Boxplots intentional binding (maybe integrate A and B) C.
+more subjective evaluation of interview / maybe world cloud / maybe sentiment analysis) }
 
 % Did they get faster? Calculate time between hand motion onset and button press. If yes, evidence that there was acceptance for the help/augmentation the system provided
-
-\subsection{The BCI}
-
-\missingfigure{!: grand average ERP at Cz comparing idle vs.
-movement intention with scalp map of voltage across all channels in the last [-100 0] interval of both idle and pre-move; B: What did the windows mean features look like.
-Boxplot with idle vs.
-pre-move and single observations}
-
-% ttest of amplitudes in last time window
-
-% classification accuracy
-
-\missingfigure{confusion matrix!, AUC, ROC -> in discussion mention that we want to do that continuously and not only look at the times preceding hand movement but at all time points}

### 2023-09-15 input/7_discussion.tex

--- 2023-04-28 input/7_discussion.tex
@@
 \section{Discussion}
+With this paper, we contributed a new way to design augmented experiences with the potential to maintain a user's SoA: a bleeding edge BCI to control EMS at moments of users' intention to interact.
+By leveraging an average of 11.5 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of .7.
+We demonstrated in eight participants that one effect of SoA, intentional binding, is maintained when using our INTENTION system.
+However, the subjectively rated level of control was lower when using the INTENTION system as compared to voluntary interaction without EMS.
+This was differentiated by participants' qualitative reports: when the stimulation aligned with participants' intentions, positive sentiments outweighed negative ones.
+When it did not, participants reported a negatively perceived loss of control.
 
-% summary here! BLUF!
+% Sandwich: 
+% Good -> [short] cool system, first of its kind setting out to augment user's action preceeding their movement in time
+Our prototype was built using off-the-shelf, affordable, equipment to physically augment users' actions.
+Taken together, all the tech to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 [anonymized].
+While we used a 64-channel research-grade EEG system, the channel selection procedure resulted in the system ultimately using a low-density coverage of about 12 channels on average for classification.
+Today, many low-density EEG devices are available on consumer markets at affordable price points, see~\citet{Niso2023-ce} for a recent summary of available wireless systems.
 
-% Outlook Ideas
-% - Strength of EMS could be set by the amplitude of the RP -> is there evidence for a correlation between amplitude and 'strength' of volitional action?
-%[ - EEG Signal may be contaminated by eye movement, maybe a worthwhile thing to build this interaction relying solely on eye tracking information, gaze fixating on the target before moving]
+Still, the system achieved what is considered to be a `good' F1 score of .7, and hence sometimes detected users' intention to interact.
+In our user study, this system performance affected users' performance in the time estimation task when comparing the three experimental conditions.
+We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency with respect to augmentation.
+We believe this to be a very promising finding considering the fact that our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+
+In line with the literature, we found that participants estimated the delay between tap and tone to be lower in both conditions where they were instructed to act on their own volition as compared to the passive condition~\cite{Moore2012-dk}.
+More recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
+However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the INTENTION system.
+Hence, the effect was maintained for both conditions of voluntary action.
+
+% Bad -> (1) situate the findings of the IB, (2) bad classification metrics not good enough to feel in control
+\subsection{Limitations}
+Two main procedural issues arose with respect to the time estimation task: First, contrasting INTENTION with BASELINE, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
+This may have introduced additional variance in the time estimation task since the exact moment of the tap that causes the tone is more obscure.
+Second, following pilot recordings, we chose to obtain participants' time estimates by asking them to type in their estimates on the keyboard.
+However, we observed that many participants, while perceiving a continuous distribution of the delays, did not answer at a continuous ms resolution but rather at steps of 50 or 100ms, skewing the distribution of their answers.
+While many different versions of the intentional binding paradigm exist~\cite{Moore2012-dk}, we would now choose a continuous slider, perhaps set at different initial positions between trials, to obtain participants' estimates.
+
+\subsubsection{Improving the Prediction of the Intent to (Inter-) act}
+While an F1 score of .7 may be considered `good' for many classification scenarios, when aiming to elicit a feeling of control this level of performance may likely not be good enough, see~\citet{Papenmeier2022-oi} for a review.
+Balancing false positive rate (FPR) and true positive rate (TPR) appropriately may prove crucial in order to elicit an experience of agency.
+
+For the current study, we must note the following limitations arising from the (potentially) insufficient performance of the classification system.
+To provide a full picture we first describe the impact from a signal detection point of view: False negative classifications meant that the system did not trigger an EMS pulse in line with participants intent to tap on the screen.
+On the other hand, false positive classification meant that an EMS pulse was sent in the absence of a true intention by the participant.
+Both cases, individually and jointly, had the potential to significantly impact the user experience and erode trust in the system.
+In the interviews, participants reported only a relatively low number of correctly detected intentions.
+This may have ultimately lead to a misalignment between users' perceptions and the observed, measurable, system behavior.
+
+As a consequence, the insights gained into the INTENTION system's potential to preserve a sense of agency are, regrettably, limited.
+The quantitative metrics employed in our study, specifically those measuring the sense of control (including temporal binding and item assessment), may not have captured the nuanced temporal alignment experiences associated with EMS augmentation and user movement intentions but rather a more holistic assessment of the entire experimental block, which included trials with missed stimulation, too early stimulation, and some trials where the stimulation was in alignment with the participants' intent.
+
+% Good -> This can be improved by using Gaze Dynamics, EMG and other things
+% If it would work, how can it be used? Implications -> playful experiences, adaptive interfaces
+Hence, a \textit{key} objective remains in improving the overall performance of the classification system.
+This can be accomplished by fusing several complementary models.
+For virtual reality (VR) platforms,~\citet{David-John2021-vg} have recently shown that gaze dynamics, especially gaze velocity, carry information with respect to the intent to interact.
+Furthermore,~\citet{Nguyen2023-me} presented further evidence that physiological signals originating from muscle activity (EMG) provide a very reliable label of movement onset~\cite{Nguyen2023-me}.
+Arguably, an augmentation system based on several models tuned to specific physiological features will reach sufficient classification performance in the very near future.
+
+% \subsubsection{Opportunities}
+\begin{comment}
+% Kasahara work, Tajima, neuromodulation? reference my affordance position paper -> copy text from there!! or during painting -> scott/gehrke -> reuse image?, affordance ++ as well, 
+- relevance of research for prostheses ( see https://link.springer.com/chapter/10.1007/978-3-030-38740-2_8)
+% At this point in time, the movement plan is already in place and so <[rework] there is slack in the perceptual system that can be exploited so user's action can be altered with them noticing>. 
+- EMS augmented interaction might make sense for bigger movements (e,g.
+back muscle to support lifting) -> might have less effect (negative) effect on the feeling of control, since it is a less specific movement
+- However, brain signal might not be the best indicator in such cases (simply motion tracking indicating the start of the movement)
+
+% playful experiences: EMS
+
+\missingfigure[]{Hand weg zieh spiel foto, beispielhafte playful experience in der es interessant ist mit dem agency konzept herumzuspielen und es spielerisch zu entdecken?}
+
+- a bit on HCINT?
+% What do I want to say here?
+Just in general describe the level of integrations that exist and then say where we would place our prototype?
+Or is this more interesting for a later discussion part?
+Maybe yes, just move this to a later stage, it does not really add here and better fits in the discussion, also then to discuss the results and what it means to be integrated in a certain way
+
 %- not only physical integration but also other adpative interfaces, or control of mobile phone (see Cornelio2022 review); % say here somewhere what action augmentation is and how it differs from body and environment/outcome augmentation
+
+% briefly summarize the papers on discriminating action augementation and other such things, reference cornelio 2022 paper
+% Previous research on augmented technology and ideas to maintain sense of of agency
+% The experience of being integrated with a technology, or of being one ``human-technology assemblage'' <REF where is this from, Danry2022 used it>
+
+% Action augmentation technology using motor actuation aims to augment the user’s actions diminishing the SoA (see Cornelio 2022). 
+% - add research
+% Here I want to say where our system lies within the design space of augmentation devices and <cornelio2022> gives a good nomenclature for this.
+Then transition over to how agency is important to consider when localizing our prototpye and then in the next paragraph talk about what agency from a neuroscientific perspective
+\end{comment}
+
+% Conclusion
+\section{Conclusions}
+In this paper, we demonstrated a system to maintain users' SoA during augmented experiences using brain signals reflecting the intent to (inter-) act.
+In our user study, we found that intentional binding effects were stronger when participants worked with the system to tap on a screen compared to when they were passively moved.
+Furthermore, the level of control working with the system was higher than when being passively moved, as indicated by subjective responses.
+
+We believe this to be an important next step towards augmented users with full integration of the technology~\cite{Mueller2020-dl}.
+Ultimately, our closed-loop system enables novel predictive interfaces that leverage the users' body directly yet still feel \textit{natural} as they align with the users' intention: For example, altering the affordance structure of the interaction in real-time~\cite{Gehrke2022-kz, Lopes2015-ze, Nataraj2020-wm}.
+
+With AI technologies becoming more and more ubiquitous in our everyday life, questions about how agency is shared between us humans and these AI technologies arise on a regular basis.
+With the likely future that these systems move directly onto our bodies, alignment with users' intentions will be the \textit{key} component driving their adoption.

### 2023-09-15 input/old/4_prototype.tex

\section{Agency Prototype}
% Here we present a prototype to overcome the disruption in experiencing agency during action augmentation. Our proposed system controls an end-effector, here EMS, by leveraging EEG signals reflecting the intent to (inter)act. To control the interaction in real-time, an EEG classifier is used to discriminate between two user states: either the user is predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or a \textit{pre-movement} state, indicating the presence of an intent to act. When the system predicts a \textit{pre-movement} state, EMS is triggered in real-time, augmenting the user's action even before their own voluntary motor command.

% \subsection{Prototype components}
Our system, depicted in Figure~\ref{}, comprised: (1) a 64-channel EEG system, (2) a 1 channel Electromyography (EMG) system, (3) a simple Arduino controlled switch, and (4) a medically-compliant EMS device connected via two electrodes worn on the arm. More information on the specific hardware specifications is given in the methods and user study sections below. To assist readers in replicating our prototype, we provide the necessary technical details and source code for data processing, classifier training, and real-time application\footnote{[anonymized]}. We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the EEG and EMG data streams available in the network and synchronize the recordings of EEG--, EMG data, and an experiment marker stream that marked sections of the study procedure.

\missingfigure{show participant with measurement system and description next to each element}

\subsection{Brain-Computer Interface}
% alternative headings:
% Brain-Computer Interface to Detect EEG Readiness Potentials }
% Classifiying EEG Signals of the Intent to Interact

The presented classification scheme required individual training data per participant, as transfer learning remains challenging with EEG~\cite{Wan2021-zz}. Therefore, we initially obtained training data on a variant of our task that did not include any muscle stimulation. In our task, participants were instructed, following an initial waiting period, to tap a touchscreen with their ring finger at their own volition, see figure ~\ref{fig:teaser}.

\subsubsection{Labelling movement classes using EMG}
To obtain behavioral labels at a high temporal resolution, we leveraged EMG from the flexor digitorum profundus to detect finger movement onsets of the ring finger. Subsequently the data were labelled into two classes: an \textit{idle} class representing resting background activity and a \textit{pre-movement} class, representing the intent to (inter-)act. To this end, EMG amplitudes were band-pass filtered from 20 to 100 Hz and subsequently squared.

% TODO Next, noisy data segments were determined. 
% A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters. % from isoutlier documentation (mean segment amplitude exceeding 3 scaled median absolute deviation)

To label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap. From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected. The two event classes were then defined as follows: \textit{pre-movement} from -1000 to 0 ms preceding the (EMG detected) movement onsets and \textit{idle}, a one second data segment between trials where participants were looking at a fixation cross.

% \missingfigure{some picture from the setup and prototype}

\subsubsection{Preprocessing EEG and Selecting discriminative channels}
The EEG training data was band pass filtered from 0.1 to 15 Hz. In a first step to prepare the EEG data for classification, noisy data segments were rejected. To this end, the EEGLAB function `autorej' was used, keeping default parameters. A trial was excluded if \textit{either} data from the \textit{idle} or \textit{pre-movement} class was rejected.

Then, to decrease dimensionality of the EEG data, a selection of discriminative channels was conducted following a selection approach outlined by~\citep{Schultze-Kraft2021-cu}. First, for both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100ms of the segment. The resulting value thus indicated how much the signal had changed in the course of the 1s segments. In line with the literature there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment. Then, all channels were sorted (1) in descending order by the signal difference in the \textit{pre-movement} segments and (2) in ascending order by the signal difference in the \textit{idle} segments. The idea is, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment. The ranks of the two criteria were joined by summation and the resulting order was saved for later use in training and real-time application of the classifier. In any case, channels C3, C4 and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.

\subsubsection{Training of EEG classifier}
To classify EEG data a regularized linear discriminant analysis (LDA) was trained for each participant individually. The EEG classifier was then used in the test phase to detect RPs and trigger EMS.

As features, the (linear) slope coefficient was obtained for both \textit{idle} and \textit{pre-movement} segments by fitting a linear regression. The classifier was then trained using the slope coefficient feature of a selection of the most discriminative channels as features, hence in the end, generating a feature vector in the dimension of channels that were kept for classification, see next step below. We purposefully constrained the dimensionality of the feature vector to counter over-fitting.

Using scikit-learn, an LDA with automatic shrinkage was cross-validated (using 5-folds) using 4 to 16 channels with a step size of adding 2 channels. Following this grid-search, the cross-validation with the highest accuracy determined the number of channels that were kept for training the model for real-time application. Ultimately, to determine the threshold at which, during real-time application, the classifier would trigger the EMS, we computed the receiver-operator characteristic and from this selected the threshold at 15\% false positive rate.

\subsubsection{Real-time application and EMS control}
During real-time application, the EEG data was buffered for the last second for the selected channels. Using MNE-python <REF> the data was filtered analogously to the training data from .1 to 15 Hz. Next, the slope was computed for the discriminative channels selected in the classifier cross-validation grid-search. This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier. To smooth the prediction output, with the goal to reduce false prediction due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the last two predictions with a weighting (.3 and .5 <?>). Then, at 10Hz update rate, this smoothed probability as well as the predicted class for the current frame where gating the EMS trigger: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, EMS was triggered.

To trigger the EMS, an Arduino was used to flip a switch whenever triggering the attached actuating hardware (in the current case EMS). In our current prototype, we used a medically-compliant EMS device. We constrained the times when the switch could be flipped to specific moments in the experiment where participants were getting ready to perform an action, thereby limiting the impact of distracting EMS pulses at resting phases during the experimental trials. %This guaranteed that participants experienced EMS only during movement preparation

%%%% Resources

% \subsection{Controlling actuated haptic experiences using brain signals reflecting the intent to interact}
% alternative headings:
% Triggering EMS using EEG Signals

% To discriminate between the two \textit{action} states, we leveraged the EEG Readiness Potential (RP), or \textit{lateralized readiness potential}. It is an amplitude fluctuation that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. A RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}. A recent study has shown that the RP is heavily ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\cite{Schultze-Kraft2021-cu} asked participants to find a way to perform voluntary movements while keeping accompanying RP amplitudes as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants unable to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification.

% old version
% The ranks were joined and the 10 best channels were selected. 
% the mean slope of the amplitude was obtained by fitting a linear regression, indicating how much the signal had changed in the course of the 1s segments. In line with the literature on the RP, there should be a change over time in the \textit{pre-movement} segment but not in the \textit{idle} segment. Then, all channels were sorted (1) in descending order by slope in the \textit{pre-movement} segments and (2) in ascending order by slope in the \textit{idle} segments. The idea was, that an ideal channel shows a strong change in signal over the course of a \textit{pre-movement} segment but shows no difference during a \textit{idle} segment. 
% First, the 10 channels with the strongest discrimination between \textit{pre-movement} and \textit{idle} segments were selected with the following procedure: 
% An EEG classifier was trained on \textit{pre-movement} and \textit{idle} data segments using 10 features from the 20 most informative EEG channels. To classify single-trial motion and EEG data we used a regularized linear discriminant analysis (LDA) that was trained for each participant individually. 
% These were baseline corrected by subtracting the mean of the last 100 ms preceding the segment. The features were concatenated across all 20 selected channels and 10 time windows to obtain a 200 dimensional spatiotemporal feature vector per \textit{pre-movement} and \textit{idle} segment.

% During the test phase, both classifiers were applied to the real-time data. With the motion data streaming at 90 Hz we subjected the current sample to the motion classifier. For the EEG data streaming at 250 Hz, the EEG data of the last 1000ms was buffered for the selected best channels and 10 windowed means per channel were computed. Identical to the training data, these features were baseline corrected by subtracting the mean from the last 100 ms preceding the last second, i.e. -1100 ms to -1000 ms, resulting in a 200 dimensional baseline corrected spatiotemporal feature vector. Hence, both classifiers yielded one output value at each sample point.

### 2023-09-15 input/1_teaser_figure.tex
Updates from Overleaf

@@
 \begin{teaserfigure}
-  \includegraphics[width=\textwidth]{tex_sample_files/sampleteaser.pdf}
-  \caption{Seattle Mariners at Spring Training, 2010.}
-  \Description{Enjoying the baseball game from the third-base
-  seats. Ichiro Suzuki preparing to bat.}
+  \includegraphics[width=\textwidth]{figures/task_design.png}
+  \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
+  \Description{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
   \label{fig:teaser}

### 2023-09-15 input/2_introduction.tex
Updates from Overleaf

@@
 % Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
-Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw}.
+Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw, Kunze2017-co}.
 However, augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
@@
 We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector.
-The  INTENTION system controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Nguyen2023-me, Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+The AUGMENTATION system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Nguyen2023-me, Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/bci_game.png}
+    \caption{Our Augmentation system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's EEG signals.
+A brain-computer interface (BCI) predicts the data to be in either of two classes: Idle or pre-movement.
+In the latter case, the user's hand is moved by trigger electrical muscle stimulation (EMS).
+Image presented with consent from participant.}
+\end{figure}
 

### 2023-09-15 input/3_related_work.tex
Updates from Overleaf

@@
 On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key~\cite{Pangratz2023-ew}.
-If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced (see figure~\ref{fig:task_design} BASELINE).
+If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced.
+% (see figure~\ref{fig:task_design} INTENTION).
 
@@
 Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS).
-Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle activating nerves.
+Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle-activating nerves.
 In these scenarios, the user computes no predictions about the movement and its outcome.
 Thus, in reference to the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
-With something or somebody else moving our body to perform, e.g., a keypress on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
-Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz} (see figure~\ref{fig:task_design} NO INTENTION)
+With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
+Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz}.
+% (see figure~\ref{fig:task_design} EXTERNAL CONTROL)
 
@@
 to discriminate between two \textit{action} states as the basis to trigger an action augmentation device or not.
-The technology we used with our prototype, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.
+The technology we used with our system, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.
 

### 2023-09-15 input/5_user_study_and_methods.tex
Updates from Overleaf

@@
 \subsection{Experimental Task, Design and Procedure}
-Participants performed a simple tapping task in three conditions with 75 trials each: BASELINE, NO INTENTION (to move) and INTENTION (to move) EMS, see figure~\ref{fig:task_design}.
+Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED, see figure~\ref{fig:task_design}.
 
-\indent\textbf{BASELINE.} The task went as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) then, they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen.
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/task_new.pdf}
+    \caption{Interaction flow depicting one trial in our touchscreen tapping task.}
+    \label{fig:progression}
+\end{figure}
+
+\indent\textbf{INTENTION.} The task went as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) then, they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen.
 In line with the literature on the origin of the RP generating process they were told ``to avoid pre-planning the movement, avoid any obvious rhythm over trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
@@
 After confirming their answer by hitting the return key, the next trial started.
-During BASELINE, participants were equipped with Electromyogram (EMG) sensors instead of EMS electrodes.
-BASELINE data was used as training data for the BCI, see section~\ref{BCI} below, as well as to select stereotypical reaction times in the NO INTENTION condition.
+During INTENTION, participants were equipped with Electromyogram (EMG) sensors instead of EMS electrodes.
+INTENTION data was used as training data for the BCI, see section~\ref{BCI} below, as well as to select stereotypical reaction times in the EXTERNAL condition.
 
-\indent\textbf{NO INTENTION.} The task structure was identical to BASELINE, however, participants were now instructed to hold and wait for the muscle stimulation to move their finger thereby eliciting the screen tap.
-The timing for the EMS trigger was taken by randomly choosing a time between the 5th and 95th percentile of their actual individual reaction time in the BASELINE condition.
+\indent\textbf{EXTERNAL.} The task structure was identical to INTENTION, however, participants were now instructed to hold and wait for the muscle stimulation to move their finger thereby eliciting the screen tap.
+The timing for the EMS trigger was taken by randomly choosing a time between the 5th and 95th percentile of their actual individual reaction time in the INTENTION condition.
 
-\indent\textbf{INTENTION.} The task and instruction were identical to BASELINE with one additional instruction: ``you will now work \textit{with} the system''.
-During INTENTION, the muscle stimulation hardware was controlled by the BCI.
+\indent\textbf{AUGMENTED.} The task and instruction were identical to INTENTION with one additional instruction: ``you will now work \textit{with} the system''.
+During AUGMENTED, the muscle stimulation hardware was controlled by the BCI.
 The classifier was set to active after the fixation cross disappeared and until a screen tap was registered.
@@
 
-The order of the conditions was not pseudo-randomized since training data obtained in BASELINE was required for both NO INTENTION and INTENTION.
-Furthermore, INTENTION was always the last condition, thereby allowing for a prolonged interview.
+The order of the conditions was not pseudo-randomized since training data obtained in INTENTION was required for both EXTERNAL and AUGMENTED.
+Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview.
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{figures/task_design.png}
-    \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
-    \label{fig:task_design}
-\end{figure*}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{figures/task_progression.png}
-    \caption{Visualization of task progression}
-    \label{fig:progression}
-\end{figure}
+% \begin{figure*}
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/task_design.png}
+%     \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
+%     \label{fig:task_design}
+% \end{figure*}
 
 \subsection{Apparatus}
-The experimental setup, depicted in Figure~\ref{}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected via two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
+The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected via two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
 To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
@@
 
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/setup.png}
+    \label{fig:setup}
+    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
+\end{figure}
+
 \indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
@@
 The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
-To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure ~\ref{fig:task_design} c.
+To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure~\ref{fig:setup}.
 % [todo] What were the settings of the stimulation?
@@
 \subsection{Brain-Computer Interface}\label{BCI}
-The data obtained in BASELINE was used to train the BCI.
+The data obtained in INTENTION was used to train the BCI.
 We utilized the [anonymized] and the EEGLAB~\cite{Delorme2004-sn} toolbox inside the MATLAB (The MathWorks Inc.
@@
 
-\subsubsection{Preprocessing EEG and Selecting discriminative channels}
+\subsubsection{Preprocessing EEG and Selecting discriminative channels}\label{eeg_methods}
 The EEG data was band pass filtered from 0.1 to 15 Hz.
@@
 \subsection{Measures of Agency Experience: Intentional Binding, Question \& Interview}
-To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the movement INTENTION block were investigated.
+To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated.
 At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Hornbaek}.
-Following INTENTION we interviewed users about their experience working with the EMS BCI.
+Following AUGMENTED we interviewed users about their experience working with the system.
 After prompting users to recall their experience and summarize what their task had been we set the focus to the tapping movement and asked them to ignore the time estimation task for the following questions.
 We entered the open part of the interview by asking: ``What did the system do?'' followed up by ``What was the difference between the three conditions''.
-After some time, and depending on their answers, we reset the focus to the INTENTION condition and asked ``How often was the system active?'' followed by ``What do you think caused the actions of the system?''.
+After some time, and depending on their answers, we reset the focus to the AUGMENTED condition and asked ``How often was the system active?'' followed by ``What do you think caused the actions of the system?''.
 This was then followed up by an `open' interview in which we frequently asked `how' and `why' questions to inquire about the user's experience.
@@
 % add back in after anonymization: All interviews were manually transcribed and then automatically translated using DeepL\footnote{https://www.deepl.com/translator}
-We analyzed the interviews that always followed the INTENTION condition by loosely following~\citet{Mayring2015-pp}.
+We analyzed the interviews that always followed the AUGMENTED condition by loosely following~\citet{Mayring2015-pp}.
 All interviews were manually transcribed and [anonymized].
 Prior to screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
-Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e.
-participants estimate of how well the system was aligned with their intention.
+Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e., participants estimate of how well the system was aligned with their intention.
 For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into positive and negative valence sentiments.
@@
 
-In line with the literature on intentional binding, we hypothesized that for both BASELINE and INTENTION conditions the time intervals should be underestimated.
-This should not be the case for the augmented NO INTENTION condition where users had no intention to move.
+In line with the literature on intentional binding, we hypothesized that the time intervals should be underestimated for the INTENTION and the AUGMENTED conditions.
+This should not be the case for the EXTERNAL condition where users had no intention to move.
 Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
-To test this, we fit a linear mixed effects model with \textit{condition} (BASELINE, NO INTENTION, INTENTION) as a fixed effect.
+To test this, we fit a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect.
 As a random effect, we considered the intercepts for participants.
@@
 
-Next, we hypothesized that subjective ratings of control over the tap movement are comparable between BASELINE and INTENTION and lower in the NO INTENTION condition.
+Next, we hypothesized that subjective ratings of control over the tap movement are comparable between INTENTION and AUGMENTED and lower in the EXTERNAL condition.
 Again, we fit a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
@@
 Adding EMS in line with participants' intention to move does not affect this underestimation.
-(2) The subjective feeling of control is comparable between BASELINE and INTENTION conditions.
-NO INTENTION should decrease the feeling of control significantly.
+(2) The subjective feeling of control is comparable between INTENTION and AUGMENTED conditions.
+EXTERNAL should decrease the feeling of control significantly.
 Additionally, we report the clustered interviews anecdotally.

### 2023-09-15 input/6_results.tex
Updates from Overleaf

@@
 In the intentional binding task, i.e.
-the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (BASELINE M = 160.3, SD = 116; NO INTENTION M = 176.7, SD = 120.7, INTENTION M = 162.6, SD = 118.9).
+the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (INTENTION M = 160.3, SD = 116; EXTERNAL M = 176.7, SD = 120.7, AUGMENTED M = 162.6, SD = 118.9).
 The condition influenced the estimation (${\chi^{2}_{(2)}} = 15.9, p < .001$).
-In NO INTENTION the estimation was higher compared to BASELINE (beta = 19.2, p < 0.001).
-Also in NO INTENTION, the estimation was higher compared to INTENTION (beta = 18.6, p < 0.001).
-No difference was observed between BASELINE and INTENTION.
+In EXTERNAL the estimation was higher compared to INTENTION (beta = 19.2, p < 0.001).
+Also in EXTERNAL, the estimation was higher compared to AUGMENTED (beta = 18.6, p < 0.001).
+No difference was observed between INTENTION and AUGMENTED.
 
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/behavioral_results.png}
-    \caption{Box-Whisker plots visualizing (a) the temporal binding effect over conditions.
+    \includegraphics[width=\columnwidth]{figures/behavioral_results.pdf}
+    \caption{(a) Temporal binding effect over conditions.
 With an average presented time interval of 350ms, lower values in the estimated time interval indicate an underestimation, i.e., temporal binding.
-(b) visualizes the subjective ratings of control across conditions.
-Significance reportings refer to post-hocs test that followed a significant likelihood ratio test of the full model with condition as a factor against the model without it.}
+(b) Subjective ratings of control across conditions.
+Significance labels obtained from post-hoc tests on estimated marginal means.}
     \label{fig:results}
@@
 The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 35.9, p < .001$).
-Post-hoc pairwise comparisons revealed that participants rated their motion control higher in BASELINE compared to NO INTENTION (beta = 4.9 , p < .0001) and BASELINE compared to INTENTION (beta = 3, p < .0001).
-Further, higher control was observed in INTENTION compared to NO INTENTION (beta = 1.9, p = .006).
+Post-hoc pairwise comparisons revealed that participants rated their motion control higher in INTENTION compared to EXTERNAL (beta = 4.9, p < .0001), and higher in INTENTION compared to AUGMENTED (beta = 3, p < .0001).
+Further, higher control was observed in AUGMENTED compared to EXTERNAL (beta = 1.9, p = .006).
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/eeg_results.pdf}
+    \caption{(a) Grand average event-related potential (ERP) at electrode Cz of pre-movement EEG data epochs preceding the last second before a movement, idle epochs of the same trial are plotted alongside but originate from a different time window, see section~\ref{eeg_methods}; (b) Bottom: Slope features for both idle and pre-movement classes at electrode Cz.
+Top: Scalp maps of slope values for both classes and all channels.
+(c): Mean and per participant ROC curves.
+Dotted line at 15 \% false positive rate indicate the selected threshold for the real-time application.}
+    \label{fig:EEG_results}
+\end{figure*}
 
@@
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{figures/EEG_results.pdf}
-    \caption{![todo]: (a) depicts the grand average ERP at Cz comparing idle vs.
-movement intention with scalp map of voltage across all channels in the last [-100 0] interval of both idle and pre-move; B: What did the windows mean features look like.
-Boxplot with idle vs.
-pre-move and single observations, ROC -> in discussion mention that we want to do that continuously and not only look at the times preceding hand movement but at all time points}
-    \label{fig:EEG_results}
-\end{figure*}
-
 The grid-search over channels resulted in the BCI leveraging on average 11.5 (SD = 3.7) channels.
@@
 \subsection{Subjective Reports}
-Six out of the eight participants made a reference to the source of the EMS in the INTENTION condition.
+Six out of the eight participants made a reference to the source of the EMS in the AUGMENTED condition.
 In that subgroup, most participants wondered how the system worked, one remarked, ``I can't explain how it works technically.''.
@@
 Some participants reported that the stimulation ``sometimes overlapped with the movement, but not often,'' or, it ``came only rarely.'' Another one noted, ``In a few cases, the stimulation came when I had already started the movement.'' For several participants, this overlap between their actions and the system's response occurred infrequently, with another user noting, ``Once, in the millisecond range between my planned movement and its execution.'' However, for some participants, there were moments of near-perfect synchronization, as described by one participant, ``In 3 cases it happened that my intention to press and the impulse of the device happened simultaneously.'' Another one noted, ``Sometimes it really happened that they overlapped.
-So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40\%'' while another said [we worked] ``15\% together.'' These accounts collectively highlight the varied and occasionally synchronous nature of the  INTENTION system's timing in relation to the users' intention.
+So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40\%'' while another said [we worked] ``15\% together.'' These accounts collectively highlight the varied and occasionally synchronous nature of the system's timing in relation to the users' intentions.
 

### 2023-09-15 input/7_discussion.tex
Updates from Overleaf

@@
 By leveraging an average of 11.5 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of .7.
-We demonstrated in eight participants that one effect of SoA, intentional binding, is maintained when using our INTENTION system.
-However, the subjectively rated level of control was lower when using the INTENTION system as compared to voluntary interaction without EMS.
+We demonstrated in eight participants that one effect of SoA, intentional binding, is maintained when using our system.
+However, the subjectively rated level of control was lower when using the system as compared to voluntary interaction without EMS.
 This was differentiated by participants' qualitative reports: when the stimulation aligned with participants' intentions, positive sentiments outweighed negative ones.
@@
 % Good -> [short] cool system, first of its kind setting out to augment user's action preceeding their movement in time
-Our prototype was built using off-the-shelf, affordable, equipment to physically augment users' actions.
+Our system was built using off-the-shelf, affordable, equipment to physically augment users' actions.
 Taken together, all the tech to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 [anonymized].
@@
 In our user study, this system performance affected users' performance in the time estimation task when comparing the three experimental conditions.
-We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency with respect to augmentation.
+We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency with respect to intention-based augmentation.
 We believe this to be a very promising finding considering the fact that our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
@@
 More recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
-However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the INTENTION system.
+However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the system.
 Hence, the effect was maintained for both conditions of voluntary action.
@@
 \subsection{Limitations}
-Two main procedural issues arose with respect to the time estimation task: First, contrasting INTENTION with BASELINE, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
+Two main procedural issues arose with respect to the time estimation task: First, contrasting AUGMENTED with INTENTION, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
 This may have introduced additional variance in the time estimation task since the exact moment of the tap that causes the tone is more obscure.
 Second, following pilot recordings, we chose to obtain participants' time estimates by asking them to type in their estimates on the keyboard.
-However, we observed that many participants, while perceiving a continuous distribution of the delays, did not answer at a continuous ms resolution but rather at steps of 50 or 100ms, skewing the distribution of their answers.
+However, we observed that many participants, while perceiving a continuous distribution of the delays, did not answer at a continuous ms resolution but rather at steps of 50 or 100 ms, skewing the distribution of their answers.
 While many different versions of the intentional binding paradigm exist~\cite{Moore2012-dk}, we would now choose a continuous slider, perhaps set at different initial positions between trials, to obtain participants' estimates.
@@
 In the interviews, participants reported only a relatively low number of correctly detected intentions.
-This may have ultimately lead to a misalignment between users' perceptions and the observed, measurable, system behavior.
+This may have ultimately led to a misalignment between users' perceptions and the observed, measurable, system behavior.
 
-As a consequence, the insights gained into the INTENTION system's potential to preserve a sense of agency are, regrettably, limited.
+As a consequence, the insights gained into the system's potential to preserve a sense of agency are, regrettably, limited.
 The quantitative metrics employed in our study, specifically those measuring the sense of control (including temporal binding and item assessment), may not have captured the nuanced temporal alignment experiences associated with EMS augmentation and user movement intentions but rather a more holistic assessment of the entire experimental block, which included trials with missed stimulation, too early stimulation, and some trials where the stimulation was in alignment with the participants' intent.

### 2023-09-15 input/2_introduction.tex
Updates from Overleaf

@@
     \includegraphics[width=\columnwidth]{figures/bci_game.png}
-    \caption{Our Augmentation system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's EEG signals.
+    \caption{Our AUGMENTATION system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's EEG signals.
 A brain-computer interface (BCI) predicts the data to be in either of two classes: Idle or pre-movement.

### 2023-09-15 input/5_user_study_and_methods.tex
Updates from Overleaf

@@
 \subsection{Experimental Task, Design and Procedure}
-Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED, see figure~\ref{fig:task_design}.
+Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED.
+% , see figure~\ref{fig:task_design}.
 
@@
     \includegraphics[width=\columnwidth]{figures/setup.png}
+    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
     \label{fig:setup}
-    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
 \end{figure}
@@
 To smooth the prediction output with the goal of reducing false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the last two predictions with a weighting (.3 and .5).
-Then, at 10Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5s.
+Then, at 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s.
 
@@
 To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated.
-At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Hornbaek}.
+At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
 Following AUGMENTED we interviewed users about their experience working with the system.
@@
 As a random effect, we considered the intercepts for participants.
-We obtained p-values of the coefficients by calculating likelihood-ratio tests~\cite{winterLinearModelsLinear2013}.
-All parameters were estimated by maximum likelihood estimation~\cite[see][]{Pinheiro2000}.
+We obtained p-values of the coefficients by calculating likelihood-ratio tests.
+All parameters were estimated by maximum likelihood estimation~\cite[see][]{Pinheiro2006-bk}.
 Post-hoc, pairwise contrasts were computed using estimated marginal means with Tukey's methods for multiple comparisons corrections~\cite{Lenth2020-xk}.
+% ~\cite{winterLinearModelsLinear2013}
 

### 2023-09-15 signal_processing/pi_extract_rp_for_plot.m
submission CHI processing and misc files

@@ -1,0 +1,113 @@
+
+%% config
+current_sys = "mac";
+eeglab_ver(current_sys);
+
+pi_bemobil_config;
+
+%% load data and parse events
+
+pIDs = [12, 14:17, 19:21] ; %1:10; % remove also 18 1
+cond = {'Baseline'};
+cz_ix = 24;
+design_erp = [];
+design_slope = [];
+
+for pID = pIDs
+
+    %% load
+    % load(fullfile(bemobil_config.study_folder, 'eeglab2python', ['sub-' sprintf('%03d', pID)], 'idle_Baseline.mat'));
+    % load(fullfile(bemobil_config.study_folder, 'eeglab2python', ['sub-' sprintf('%03d', pID)], 'pre_move_Baseline.mat'));
+
+    load(fullfile(bemobil_config.study_folder, 'eeglab2python', ['sub-' sprintf('%03d', pID)], 'idle_processed.mat'));
+    load(fullfile(bemobil_config.study_folder, 'eeglab2python', ['sub-' sprintf('%03d', pID)], 'pre_move_processed.mat'));
+
+    %% get ERP at Cz
+
+    baseline_cond_trials = 70;
+    idle = idle(:,:,1:baseline_cond_trials);
+    pre_move = pre_move(:,:,1:baseline_cond_trials);
+
+    idle_means = mean(idle,3);
+    pre_move_means = mean(pre_move,3);
+
+    % base correct
+    idle_means = idle_means - mean(idle_means(:,1:25),2);
+    pre_move_means = pre_move_means - mean(pre_move_means(:,1:25),2);
+
+    idle_cz = idle_means(cz_ix,:)';
+    pre_move_cz = pre_move_means(cz_ix,:)';
+    cz = [idle_cz; pre_move_cz];
+
+    idle_last_win(pID,:) = mean(idle_means(:,226:end),2);
+    pre_move_last_win(pID,:) = mean(pre_move_means(:,226:end),2);
+
+    for t = 1:size(idle,3)
+        for i = 1:63
+            p = polyfit(1:250, idle(i,:,t), 1);
+            slope_idle(pID,i,t) = p(1);
+        end
+    end
+    
+    for t = 1:size(pre_move,3)
+        for i = 1:63
+            p = polyfit(1:250, pre_move(i,:,t), 1);
+            slope_pre_move(pID,i,t) = p(1);
+        end
+    end
+
+    slope_idle = mean(slope_idle,3);
[…]

### 2023-09-15 signal_processing/pi_import.m

@@ -1,6 +1,6 @@
 function pi_import(cfg, subject, training)
 
-    sessionNames                        = {'Baseline', 'EMS1', 'EMS2'};
+    sessionNames                        = {'Baseline', 'passive', 'agency1'};
     
     % general metadata shared across all modalities
     %--------------------------------------------------------------------------
@@ -67,7 +67,7 @@ function pi_import(cfg, subject, training)
         'PO4', 'PO8', 'P6', 'P2', 'CPz', 'CP4', 'TP8', 'C6', 'C2', 'FC4', ...
         'FT8', 'F6', 'AF8', 'AF4', 'F2', 'VEOG' , 'EMG'};
     config.eeg.ref_channel        = 'FCz'; % optional, relevant only if you want to re-use the ref channel after re-referencing
-    config.other_data_types        = {'physio'};
+    % config.other_data_types        = {'physio'};
     
     %%
     
@@ -75,9 +75,9 @@ function pi_import(cfg, subject, training)
         config.session                = session{1};              % required if there are more than 1 session
         config.filename               = [cfg.study_folder filesep '0_source-data' filesep sub_path filesep [session{1}, '.xdf']];
 
-        if ~strcmp(config.session, 'Baseline')
-            config.phys.streams{1}.stream_name          = 'eeg_classifier';            % optional
-        end
+        % if ~strcmp(config.session, 'Baseline')
+        %     config.phys.streams{1}.stream_name          = 'eeg_classifier';            % optional
+        % end
 
         bemobil_xdf2bids(config, ...
             'general_metadata', generalInfo,...
@@ -89,7 +89,7 @@ function pi_import(cfg, subject, training)
     %----------------------------------------------------------------------
     config.study_folder             = cfg.study_folder;
     config.session_names            = sessionNames;
-    config.set_folder               = cfg.study_folder; %fullfile(config.study_folder, '2_raw-EEGLAB');
+    config.set_folder               = fullfile(config.study_folder, '2_raw-EEGLAB');
     bemobil_bids2set(config);
     
 end

### 2023-09-15 signal_processing/pi_parse_events.m

@@ -5,12 +5,16 @@ function [eeglab_struct] = pi_parse_events(eeglab_struct)
     for i = 1:numel(eeglab_struct.event)    
         current_event = cellstr(strsplit(eeglab_struct.event(i).type, ';'));
 
-        for j=1:length(current_event)
-            key_val = cellstr(strsplit(current_event{j}, ':'));
-            eeglab_struct.event(i).(key_val{1}) = key_val{2};
-            if j==1
-                eeglab_struct.event(i).value = strcat(key_val{1}, ':', key_val{2});
+        if length(current_event) > 1
+            for j=1:length(current_event)
+                key_val = cellstr(strsplit(current_event{j}, ':'));
+                eeglab_struct.event(i).(key_val{1}) = key_val{2};
+                if j==1
+                    eeglab_struct.event(i).value = strcat(key_val{1}, ':', key_val{2});
+                end
             end
+        % else
+            % pass
         end
     end
 end

### 2023-09-15 signal_processing/pi_pipeline.m

@@ -14,10 +14,10 @@ pi_bemobil_config;
 
 % set to 1 if all files should be recomputed and overwritten
 force_recompute = 1;
-subjects = 2:3;
+subjects = [12, 14:17, 19:21] ;
 
 for subject = subjects
-%     pi_import(bemobil_config, subject, 0);
+    pi_import(bemobil_config, subject, 0);
     pi_preprocess_EEG;
 %     pi_preprocess_Physio;
 end

### 2023-09-15 signal_processing/pi_plot_topomaps.m

@@ -1,0 +1,32 @@
+%% plot topomap
+
+current_sys = "mac";
+pi_bemobil_config;
+
+load('/Volumes/Lukas_Gehrke/fastReach/data/eeglab2python/locs.mat');
+
+% load csv scores and concatenate to one matrix
+pIDs = [12, 14:17, 19:21] ; %1:10; % remove also 18 1
+
+% what is all_chans_score_order and what does it mean???
+for pID = pIDs
+    tmp = readtable(fullfile(bemobil_config.study_folder, 'eeglab2python', ['sub-' sprintf('%03d', pID)], 'all_chans_score_order.csv'));
+    all_chans_score_order(pID,:) = tmp.Var1;
+end
+all_chans_score_order = all_chans_score_order(pIDs,:);
+
+
+slope_pre_move = slope_pre_move(pIDs,:,:);
+slope_pre_move(slope_pre_move == 0) = NaN;
+
+slope = nanmean(slope_pre_move, 3);
+slope = mean(slope, 1);
+
+% slope = mean(slope_pre_move(pIDs,:),1);
+% slope = mean(slope_idle(pIDs,:),1);
+% slope(end) = 0;
+
+slope = zscore(slope);
+
+% topoplot(slope, EEG.chanlocs)
+figure; topoplot(slope,locs,'style','map','electrodes','labelpoint','chaninfo',EEG.chaninfo);

### 2023-09-15 signal_processing/pi_preprocess_EEG.m

@@ -35,13 +35,13 @@ EEG = pop_loadset('filename',[ bemobil_config.filename_prefix num2str(subject) '
 
 %% extract ECG and GSR channel and add to physio data
 
-Physio = pop_loadset('filename',[ bemobil_config.filename_prefix num2str(subject) '_' bemobil_config.merged_physio_filename],'filepath',input_filepath);
-Physio.data(end+1,:) = EEG.data(end,:);
-Physio.nbchan = Physio.nbchan+1;
-Physio.chanlocs(end+1).labels = EEG.chanlocs(end).labels;
-Physio = eeg_checkset(Physio);
-
-pop_saveset(Physio, 'filename',[ bemobil_config.filename_prefix num2str(subject) '_' bemobil_config.merged_physio_filename],'filepath',input_filepath);
+% Physio = pop_loadset('filename',[ bemobil_config.filename_prefix num2str(subject) '_' bemobil_config.merged_physio_filename],'filepath',input_filepath);
+% Physio.data(end+1,:) = EEG.data(end,:);
+% Physio.nbchan = Physio.nbchan+1;
+% Physio.chanlocs(end+1).labels = EEG.chanlocs(end).labels;
+% Physio = eeg_checkset(Physio);
+% 
+% pop_saveset(Physio, 'filename',[ bemobil_config.filename_prefix num2str(subject) '_' bemobil_config.merged_physio_filename],'filepath',input_filepath);
 
 EEG.chanlocs(end) = [];
 EEG.data(end,:) = [];

### 2023-12-13 input/6_results.tex
revision v1

@@
 The condition influenced the estimation (${\chi^{2}_{(2)}} = 15.9, p < .001$).
-In EXTERNAL the estimation was higher compared to INTENTION (beta = 19.2, p < 0.001).
-Also in EXTERNAL, the estimation was higher compared to AUGMENTED (beta = 18.6, p < 0.001).
+In EXTERNAL the underestimation was less pronounced compared to INTENTION ($beta = 19.2, p < 0.001$).
+This was also the case in EXTERNAL, with less pronounced underestimations of the time interval between tap and tone compared to AUGMENTED ($beta = 18.6, p < 0.001$).
 No difference was observed between INTENTION and AUGMENTED.
@@
 The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 35.9, p < .001$).
-Post-hoc pairwise comparisons revealed that participants rated their motion control higher in INTENTION compared to EXTERNAL (beta = 4.9, p < .0001), and higher in INTENTION compared to AUGMENTED (beta = 3, p < .0001).
-Further, higher control was observed in AUGMENTED compared to EXTERNAL (beta = 1.9, p = .006).
+Post-hoc pairwise comparisons revealed that participants rated their motion control higher in INTENTION compared to EXTERNAL ($beta = 4.9, p < .0001$), and higher in INTENTION compared to AUGMENTED ($beta = 3, p < .0001$).
+Further, higher control was observed in AUGMENTED compared to EXTERNAL ($beta = 1.9, p = .006$).
 
@@
 The slope feature for the exemplary channel discriminates well between the two classes (${t_{(8)}} = 3.6, p = .003$), see figure ~\ref{fig:EEG_results} b bottom.
-The scalp maps in figure~\ref{fig:EEG_results} b top show the (color-coded) mean slope for each channel and for each class.
+The scalp maps in figure~\ref{fig:EEG_results} b top show the (color-coded) mean slope for each channel and each class.
 Central channels on the contralateral side to the moving finger on the right hand show a negative slope for the pre-movement class and a neutral slope for the idle class.
-Furthermore, differences in slope can also be observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifests only for the idle class.
+Furthermore, differences in slope were also observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifested only for the idle class.
 
 The grid-search over channels resulted in the BCI leveraging on average 11.5 (SD = 3.7) channels.
-Classifier cross-validation resulted in a mean F1 score of .7 (SD = .03), see figure~\ref{fig:EEG_results} c.
-We set the detection threshold to 15\% false positive rate and at that rate, observed a mean threshold of 57\% (SD = .02).
-Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57\% probability.
+\textcolor{green}{Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9, AF3, F5, F7, F2, FT8, and AF7.} The classifier cross-validation resulted in a mean F1 score of .7 (SD = .03), see figure~\ref{fig:EEG_results} c.
+We set the detection threshold to 15 \% false positive rate and at that rate, observed a mean threshold of 57 \% (SD = .02).
+Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57 \% probability.
 % TODO add percentage of trials the prototype was activated before touching the screen (SD = ?)
@@
 Some participants reported that the stimulation ``sometimes overlapped with the movement, but not often,'' or, it ``came only rarely.'' Another one noted, ``In a few cases, the stimulation came when I had already started the movement.'' For several participants, this overlap between their actions and the system's response occurred infrequently, with another user noting, ``Once, in the millisecond range between my planned movement and its execution.'' However, for some participants, there were moments of near-perfect synchronization, as described by one participant, ``In 3 cases it happened that my intention to press and the impulse of the device happened simultaneously.'' Another one noted, ``Sometimes it really happened that they overlapped.
-So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40\%'' while another said [we worked] ``15\% together.'' These accounts collectively highlight the varied and occasionally synchronous nature of the system's timing in relation to the users' intentions.
+So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40 \%'' while another said [we worked] ``15 \% together.'' These accounts collectively highlight the varied and occasionally synchronous nature of the system's timing in relation to the users' intentions.
 

### 2023-12-13 input/rebuttal/1AC.tex
revision v1

This work presents a BCI-based augmentation that is able to stimulate users' muscles at their intent to interact. The system measured readiness potential (RP) from EEG signals to calculate action intention. The authors conducted a user study to measure intentional binding while interacting. The results show a higher level of control working with the system compared to when being passively moved (via EMS).

I find the goal of this work is well-motivated and would fit into CHI. All reviewers appraise the work for its ambitious goal (2AC), has a sensible approach, overall aim interesting (R1), and introduce an original system that can be an important step towards action augmentation using neural input (R2).

However, the paper is well-written and has good-quality research. However, there are some issues that should be addressed in the revision:

** 2AC: 
It is unclear if both EMS triggering and EMG measurements were performed in the AUGMENTED condition. If so, how that was done technically should be explained.

The authors did a comparison between three conditions. However, how those conditions are compared to other techniques (as a baseline) in previous studies should be discussed.

Both 2AC and R1 are concerned with a justification of the classification rate (10Hz) and the maximum interaction time reduction that this technique can achieve.

** R1: 
I agree with the R1's concern regarding the low number of participants (8 people). As the trials were repeatedly measured and analysed (independently?), would the means of those trials (i.e., time estimation and sense of control rating) should also be used?

The used timing in the EXTERNAL condition was not particularly extreme and randomly picked from within 5-95\% of that user's previous timings. So they weren't faster here, and it's unclear whether they were faster in the augmented condition either.

Justification about the electrode location (Cz): why the main reference throughout is Cz. The paper mentions this electrode earlier together with C3 and C4 as commonly used, but how does Cz stand out from the three?

** R2 has some suggestions about clarifying the definition of action augmentation and providing a brief explanation of how the present study achieves that much earlier in the paper can improve the readability and clarity. A good suggestion also is the inclusion of references to different states (idle and pre-movement state.

I urge the authors to read the reviewers' comments in detail to improve their work in the revision. I believe these can be fixed within the CHI timeline for R&R.

--

### 2023-12-13 input/2_introduction.tex

@@
 Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw, Kunze2017-co}.
-However, augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
+\textcolor{green}{These technological advances include the miniaturization of the actuating hardware to wearable form factors and the direct sensing and stimulation capabilities of neural interfaces.
+Especially due to these characteristics, recent perspectives promote a change of the computing era from human-computer \textit{interaction} to \textit{integration}~\cite{Mueller2020-dl}.
+One key change in perspective is, that \textit{integrated} user's share agency with the computing machinery to execute tasks.
+This is a critical distinction as \textit{integration} technologies are designed to directly influence people’s bodies, their actions, and the resulting action outcomes.}
+
+\textcolor{green}{Besides body and outcome augmentation, action augmentation has been defined as the case where a ``system assists the user’s action to produce the intended outcome.''~\cite{Cornelio2022-aq}.
+While such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer, when it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of ones body.
+Unfortunately,} such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
 
@@
 Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us.
-It has frequently been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}.
-A key challenge to drive the adoption of human action augmentation then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
+It has been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}.
+Hence, a key challenge to drive the adoption of human action augmentation then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
 
@@
 % What have others done to address this and what do we propose "instead"
-One goal of user augmentation is to increase reaction times, thereby making an interaction more efficient.
+\textcolor{red}{One goal of green `on-body' action augmentation is to increase reaction times, thereby making an interaction more efficient.
 Previous work on such augmentation scenarios has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk}.
 Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
-Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.
+Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.}
 
-In experimental settings, this is typically achieved by using strongly controlled `stimulus-response' paradigms.
+\textcolor{red}{In experimental settings, this is typically achieved by using strongly controlled `stimulus-response' paradigms.
 For example, reaction times when tapping on a touchscreen in response to a presented stimulus on screen, can be reduced by an augmentation device while maintaining SoA for the user~\cite{Kasahara2019-sk, Kasahara2021-gy}.
@@
 However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
-In other words, how can we design a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions?
+In other words, how can we design a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions?}
 
@@
 % Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
-In this paper, we present a bleeding edge prototype to challenge the disruption in experiencing agency during action augmentation.
+In this paper, we present a prototype to overcome the disruption in experiencing agency during action augmentation.
 We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector.
-The AUGMENTATION system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Nguyen2023-me, Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+The augmentation system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
 
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/bci_game.png}
-    \caption{Our AUGMENTATION system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's EEG signals.
+    \includegraphics[width=\columnwidth]{figures/teaser_new.png}%{figures/bci_game.png}
+    \caption{Our augmentation system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals.
 A brain-computer interface (BCI) predicts the data to be in either of two classes: Idle or pre-movement.
-In the latter case, the user's hand is moved by trigger electrical muscle stimulation (EMS).
-Image presented with consent from participant.}
+In the latter case, electrical muscle stimulation (EMS) is triggered and the user's hand is moved.
+Image taken with consent from participant.}
 \end{figure}
 
+To control the interaction in real-time, an RP-based classifier distinguished between two user states: \textit{idle}, reflecting the absence of an intent to act, and \textit{pre-movement}, indicating the presence of an intent to act.\textcolor{red}{When the system predicted a \textit{pre-movement} state, the user's action was augmented, potentially even before their own voluntary motor command.
+In our user study, we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their SoA.} \textcolor{green}{During \textit{idle}, participants were passively looking at a fixation cross.
+Instead, during \textit{pre-movement}, participants were instructed to voluntarily initiate a tap on a touchscreen whenever they felt the urge to do so.
+Previous work has indicated that the RP emerges during formation of conscious intention and is specific to voluntary action~\cite{Schultze-Kraft2020-rm, Travers2020-hf, Pares-Pujolras2019-ll}.}
+
+\textcolor{green}{Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially preceding their voluntary motor command.
+This augmentation moved the ring finger in accordance with the user's intention to act.
+We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle.
+In our user study, a mixed-methods research approach was employed to explore whether aligning the physical impact on the user's body with their intention to move preserved their Sense of Agency (SoA).}
+
+%%% Resources
+\begin{comment}
 % TODO merge this with the figure below
@@
 
-To control an interaction in real-time, the EEG-based RP classifier was used to discriminate between two user states: either the user was predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or the user was predicted to be in a \textit{pre-movement} state, indicating the presence of an intent to act.
-When the system predicted a \textit{pre-movement} state, the user's action was augmented, potentially even before their own voluntary motor command.
-In our user study, we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their SoA.
-
-%%% Resources
 % add results and general outcomes here?
@@
 % - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?
+
+\end{comment}

### 2023-12-13 input/3_related_work.tex

@@
 If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced.
-% (see figure~\ref{fig:task_design} INTENTION).
+%, see figure~\ref{fig:task_design} INTENTION).
 
-In human-computer interaction (HCI) research, these constructs are categorized in their own terms.
+In human-computer interaction (HCI) research, these constructs are categorized using a different terminology.
 Often, the term \textit{pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements.
@@
 In these scenarios, the user computes no predictions about the movement and its outcome.
-Thus, in reference to the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
+Thus, concerning the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
 With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
-Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz}.
-% (see figure~\ref{fig:task_design} EXTERNAL CONTROL)
+~\textcolor{red}{Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz}.}
+%, see figure~\ref{fig:task_design} EXTERNAL CONTROL)}
 
@@
 
-On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously reflected~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
-Following seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}.
-Critically, this temporal compression only occurs following movement intention.
+On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
+Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}.
+Critically, this temporal compression only occurs following movements that were intended.
 The action outcome is mentally \textit{bound} to the intention.
@@
 
-% alternative subsection heading: Brain Signals Reflecting the Intent to (Inter-)Act
-% this must make the link clear that we need intention for agency experience and intention is something we can set out to measure using EEG!!!
-\subsection{Controlling Actuated Haptic Experiences using Brain Signals Reflecting the Intent to (Inter-)act}
-Of the possible physiological signals that can be used to further investigate the experience of agency, the electroencephalogram (EEG) is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
-The Readiness Potential (RP), or \textit{lateralized readiness potential} is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
-The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand.
-Since its measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
+\subsection{\textcolor{green}{Controlling Actuated Haptic Experiences}}
+
+\textcolor{green}{Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA frequently use highly controlled `stimulus-response' paradigms.
+For example, scenarios where participants are instructed to tap on a touchscreen in response to a presented stimulus on the screen.
+Here, participants' behavior can be predicted with very high certainty to follow the presented stimulus, and estimating their reaction time is very accurate.
+In such controlled scenarios, the timing of an action augmentation device can be tuned to be near optimal.
+Hence, \textit{pre-empting} the user's motion can be designed to fall in line with their intention to move, thereby maintaining SoA.
+Previous work on such an augmentation scenario has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.}
+
+\textcolor{green}{Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
+Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.
+However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
+In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?}
+
+\textcolor{red}{\subsection{Controlling Actuated Haptic Experiences using Brain Signals Reflecting the Intent to (Inter-)act}}
+
+\subsubsection{\textcolor{green}{Using Brain Signals Reflecting the Intent to (Inter-)act for Action Augmentation}}
+\textcolor{green}{One possible design solution is to leverage physiological signals for action augmentation}.
+Of the possible physiological signals that can be leveraged\textcolor{red}{used to further investigate the experience of agency}, the EEG is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
+The RP, or \textit{lateralized readiness potential}, is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
+The RP is reliably observed at electrodes placed over the \textcolor{green}{sensori}motor cortex contralateral to the acting hand.~\textcolor{green}{In the 10-20 system for EEG electrode placement~\cite{Jasper1983-uw}, these are electrode C3 located over the sensorimotor cortex of the left hemisphere, and C4 vice versa.
+However, activity observed at electrode Cz is reported most frequently as it reflects neural activity originating from the sensorimotor cortex without lateralization bias.} Since the RPs' measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
 However, evidence abounds for its role in action preparation.
@@
 The authors found that participants were able to veto their self-initiated movement if the red dot appeared no later than 200ms preceding their movement onset.
-After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot appearance on screen.
+After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot's appearance on screen.
 
-In the present study, we leveraged the RP in a similar way, i.e.
+\textcolor{red}{In the present study, we leveraged the RP in a similar way, i.e.
 to discriminate between two \textit{action} states as the basis to trigger an action augmentation device or not.
-The technology we used with our system, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.
+The technology we used with our system, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.}
 
 %%% resources
-
 \begin{comment}
@@
 - EEG patterns related to readiness potential and intentional  binding sometimes show disceprant patterns ( see WEN and Imamizu 38 und 39) (wittmann 2014, Goldberg 2017)
-\end{comment}
 
@@
 % Triggering EMS using EEG Signals
+
+\end{comment}

### 2023-12-13 input_marked_up/rebuttal/rebuttal_letter.tex
create duplicate marked up version

\section{comparison with baseline conditions}
% [1AC] The authors did a comparison between three conditions. However, how those conditions are compared to other techniques (as a baseline) in previous studies should be discussed.

% [2AC] The current baseline (EXTERNAL) condition is too weak. It is a trivial finding that the SoA of the AUGMENTED condition is higher than that of the EXTERNAL condition. Why has no comparison been made with a simpler technique where EMS is triggered by heuristic rules rather than based on EEG signals? Is there a reason the authors did not use as a baseline any technique suggested in previous studies?
% So it seems we should consider the INTENTION condition as a more important baseline for this study. And the fact that the AUGMENTED condition shows a lower SoA than the INTENTION condition (Figure 4) is a common problem with EMS-based techniques found in previous studies and is also a key motivation for this study. In other words, it appears that important problems have not been solved even though EEG signals have been utilized.
% How does the difference in SoA between the INTENTION condition and the AUGMENTED condition in previous studies compare to the difference in this study?

- [1AC, 2AC, R1]: comparison with other baseline conditions
The major motivation for the current work was to present a closed-loop augmentation system using neural input that is entirely decoupled from any outside constraints. Ultimately, we aim at designing a system that a user can wear and use in a variety of settings and applications as no knowledge about the user's environment should b enecessary for it to work. With this goal in mind, we did not find a useful baseline condition from previous studies to compare our interface to. We believe the simple case of random activation, for example 'send 10 EMS pulses within every 5 minutes' to not be a useful heuristic since we have shown that our classifier performs above chance level (F1 score of .7). Hence, as we have demonstrated, it outperforms activating EMS at chance level. Any other method to predict the user's intent to interact would be reliant on measuring some form of behavior or physiology, or constraining the interaction to a stimulus-response pattern.

% - Instead we chose to position the agency experience when using the prototype between the two edge cases of full control (INTENTION) vs. being passively moved (EXTERNAL).
% - because it would distract from the goal that this work was set out to tackle.
% - we want to have free behavior therefore there is no heuristic, or just a very basic one like activate 10 times every minute.
% 1. we focused on investigating sense of agency, not rt
% [ ] we adjusted the framing of the paper in the introduction accordingly and better situate it now within the action augmentation literature
% [ ] compare it in reference to earlier work and add a future outlook on subsequent validation steps against existing approaches and the potential fusion of these.

\section{on reduction in reaction time}

% [1AC] Both 2AC and R1 are concerned with a justification of the classification rate (10Hz) and the maximum interaction time reduction that this technique can achieve.

% [1AC] The used timing in the EXTERNAL condition was not particularly extreme and randomly picked from within 5-95\% of that user's previous timings. So they weren't faster here, and it's unclear whether they were faster in the augmented condition either.

% [2AC] I agree with the authors' idea that reaction time can be reduced through EMS if we know in advance when the user will move. However, according to the technique proposed by the authors, the maximum amount of reaction time we can reduce is the time interval between detection of the readiness potential (RP) and execution of the movement. Isn't this time interval usually very short (i.e., less than 200 ms)? Furthermore, for more robust classification, wouldn't EEG data sometimes need to be observed for a longer period of time, and as a result, the margin for reducing reaction time could be further reduced? We even have to consider the latency of the EMS itself.

% [2AC] Considering all these factors, I am curious to what maximum reaction time reduction the authors' proposed technique can actually achieve.

% [2AC] The time interval between the detection of RP and the actual movement of the participant's finger is very short (in Figure 5a), so isn't the classification rate of 10 Hz too low?

% [R1] But I'd like to start with something I found a bit odd. The paper positions itself within the human augmentation space, and I think that makes sense. The intro and method sections also discuss this. But did the augmented condition here actually improve reaction times? That does not actually become clear. As stated on p. 6, in the external condition, the used timing was not particularly extreme and randomly picked from within 5-95\% of that users previous timings. So they weren't actually faster here, and it's not clear whether they were faster in the augmented condition either. That might well not matter, given that this paper is not about reaction times per se. However, it is a curious aspect of the experimental design, reporting, and framing and thus would be good to discuss a bit more. Given that there probably would be a higher experienced loss of control with out of the ordinary reaction times, this might well have had an influence on the results, though.

- [1AC, 2AC, R1]: on the potential reduction in reaction time

The focus of the revised manuscript is now on SoA and is not deferred as much anymore by a secondary aspect of the system, i.e., a reduction in reaction times. In the revised manuscript we tightened the focus and framing on SoA and moved the part on previous studies reducing reaction times from the introduction section to related work. This better fits with the direct contrast of using predictive models (based on physiological signals) instead of controlled environments for action augmentation that falls in line with a user's intentions. Furthermore, we clarified reviewers' comments concerning the reduction in RT our system can achieve, i.e. prediction smoothing.

With the current implementation of the prototype, we saw an end-to-end latency from the measured physiological signals to the gating of the EMS switch of around 40 ms. Since this was implemented in relatively slow python code the time could be brought down significantly with more native code. While we could have applied the classifier at 20 Hz (one prediction every 50 ms) we wanted to be certain not to run into problems due to buffer overflow which is why we kept the real-time classification at 10 Hz. As 2AC points out, the RP precedes the movement onset by 200-300 ms. If the classifier works as intended, it should be able to pre-empt the movement, sending the first EMS pulse at 150 ms (200 - 50 ms) before the movement.

However, it is difficult to assess the reduction in reaction time at a high resolution as no label exists for the 'true' intention onset. This is the key contribution (and distinction) as compared to previous works in controlled environments. We are adding labels based on implicit classification of EEG signals in freely behaving users. Hence, we cannot assess the reduction based on any other metric other than the classifier, but if it works as designed, the reduction is there. However, we tried to explore the subjective experience of a potential reduction in RT through interviews. Another way would be to determine this on a single trial level by asking questions (explicit) or by inferring from neural signals (implicit). We added a paragraph to the discussion.

\section{technical details about the classifier}

By including a random effect for participants in our LME analyses, we accounted for the dependency of repeated measures within participants. The analysis is correct. We clarified [R2]'s comment about multiple comparison corrections and added details about why electrode Cz was used to report exemplary EEG results. Furthermore, we now report electrodes that were leveraged for classification in several participants and rephrased how we smoothed the prediction output [R1].

EMG and EMS were NOT recorded at the same time, as EMG was only recorded in the first condition to obtain labels for training data for the classifier. Following [2AC]'s suggestion, we added a new figure to reflect this and show the processing steps in detail. Additionally, we changed the order of the apparatus and procedure section, so readers have a better understanding of the setup before going into the task.

To improve readability and clarity [R2, R1] we now explain how the augmentation was achieved much earlier in our paper (both in the abstract and introduction). Furthermore, we better situate our augmentation system in the human augmentation/integration literature, thereby making our paper more accessible to a broader HCI audience. Lastly, we now include references to the distinction between pre-movement and idle EEG states.

% [R1] A big part of the work here is the classifier that determines whether a participant is idling or just about to move. I'm not an expert here, but I was left with a few questions regarding this classifier. For starters, it's not clear to me why the main reference throughout is Cz. The paper mentions this electrode earlier together with C3 and C4 as commonly used, but how does Cz stand out from the three? All the training is then in relation to the time of movement as detected by the EMG. So not when the brain started to plan the motion, but when the muscle actually is activating. As the whole second up to that moment is training data, it thus contains not just the planning either (but of course _also_ the planning). As Section 3.4.1 notes, 1s signal windows are condensed to slope values, based on the first and last 100ms. Per Section 3.4.3, this window is updated 10 times a second, so essentially shifted by 100ms. However, the predictions are smoothed with two previous ones (btw. I don't get the weighting: 0.3 & 0.5 for the previous predictions, and then presumably only 0.2 as weight for the current one? Why such a low value for the most up-to-date value?). However, doesn't that also mean that prediction of pre-movement is quite laggy? For a slope to be noticeable, the window must likely overlap quite a bit with the increase and then another 200ms of delay come to the averaging. With the EMS then active for 500ms, I'm pretty much wondering whether this did preempt the motion and whether the system then remained active longer than needed. As the paper itself states, the performance of the classifier is mixed with several participants reporting issues. So this is probably an aspect to provide a bit more details and explanations for.

% [ ] also why wasnt a link function used? assumed linear relationship between time estimation / questionnaires and study design

% [2AC] Most of the movements we perform in daily life are continuous rather than discrete, such as tapping. For example, human continuous movement (e.g., pointing) is known to be based on intermittent motor control [1]. In such cases, RP is expected to occur intermittently and repeatedly. Even in such cases, can the technique suggested by the authors be applied?
% -> unfold reference, % - unfold overlapping ERPs, for example with unfold toolbox

### 2023-12-13 input_marked_up/5_user_study_and_methods.tex
create duplicate marked up version

--- 2023-09-15 input/5_user_study_and_methods.tex
@@
 \section{User study \& Methods}
-With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user's own brain signals serve as the control signal to a physical end effector.
-To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.
-% todo add some more framing here to make it easier to step through the methods, maybe: For this paper we designed a prototype that lies between the two edge cases -- no intention and control to full control -- and investigate user's sense of agency using both explicit and implicit measures.
+\textcolor{green}{With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user's own brain signals serve as the control signal to a physical end effector.
+Therefore, we conducted a user study comparing three conditions of agency experience.
+With the first two conditions, INTENTION and EXTERNAL we queried users at the two edges of agency experience.
+In EXTERNAL participants had no intention to move and no control over their movement.
+On the other hand, in INTENTION, participants were acting as they would in their day-to-day lives, fully in control.
+We then investigated in a third condition where our AUGMENTED prototype sits on the spectrum between INTENTION and EXTERNAL, in other words, whether it preserved agency and to what degree.
+To answer our question, we employed a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.}
+
+\textcolor{red}{With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user’s own brain signals serve as the control signal to a physical end effector.
+To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.}
 
@@
 %the Department of Psychology and Ergonomics at TU Berlin (tracking number: BPN\_GEH\_2\_230130220421).
+
+% \begin{figure*}
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/task_design.png}
+%     \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
+%     \label{fig:task_design}
+% \end{figure*}
+
+\subsection{Apparatus}
+The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected to two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
+To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
+% TODO add links after anonymization is lifted.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/setup_fixed.png}
+    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
+    \label{fig:setup}
+\end{figure}
+
+\indent\textbf{(1) EMG Recording.} EMG data was recorded from 1 bipolar channel using the BrainAmp ExG amplifier (BrainProducts GmbH, Gilching, Germany).
+The two electrodes were placed above the flexor digitorum profundus with a reference electrode located on the wrist bone.
+EMG data was collected in synchrony with the EEG data through BrainProducts' BrainVision Recorder.
+\textcolor{green}{EMG data was only recorded for the first experimental condition to obtain labels for classifier training.
+After that, the EMG electrodes were changed for EMS electrodes.}
+
+\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes referenced to \textcolor{green}{electrode FCz} in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
+Electrodes were placed according to the extended international 10–20 system \cite{Jasper1983-uw}.
+One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10k$\Omega$.
+EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
+
+We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the data streams available in the network and synchronize the recordings of EEG/EMG data and an experiment marker stream that marked sections of the study procedure.
+
+\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
+We utilized the flexor digitorum profundus since we found that we can robustly actuate it without inducing unintended motion of neighboring fingers.
+This finger actuation was achieved via a medically compliant battery-powered muscle stimulator (TENS/EMS Super Duo Plus, prorelax, Düren, Germany).
+The EMS system's output was controlled by flipping a solid state relay (silent) connected via an Arduino Uno (Arduino, Monza, Italy) to the experiment computer.
+The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
+To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure~\ref{fig:setup}.
+
+\indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
+In addition to the tablet, we used an external keyboard to allow \textcolor{red}{response input of users indicating}~\textcolor{green}{users to input} their timing judgements.
 
@@
 
-\indent\textbf{INTENTION.} The task went as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) then, they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen.
-In line with the literature on the origin of the RP generating process they were told ``to avoid pre-planning the movement, avoid any obvious rhythm over trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
+\indent\textbf{INTENTION.} The task was as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen, see figure~\ref{fig:progression}.
+In line with the literature on the origin of the RP generating process, they were told ``to avoid pre-planning the movement, avoid any obvious rhythm across trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
 (4) After the screen was tapped, a tone was played at a pseudo-random delay of 200, 350, or 500ms.
-Participants were now asked to estimate the delay, typing in their answer on a number pad of an attached keyboard.
+Participants were now asked to estimate the delay, typing in their answers on a number pad of an attached keyboard.
 After confirming their answer by hitting the return key, the next trial started.
-During INTENTION, participants were equipped with Electromyogram (EMG) sensors instead of EMS electrodes.
-INTENTION data was used as training data for the BCI, see section~\ref{BCI} below, as well as to select stereotypical reaction times in the EXTERNAL condition.
+
+During INTENTION, participants were equipped with \textcolor{red}{Electromyogram (EMG)}\textcolor{green}{EMG} sensors instead of EMS electrodes.
+\textcolor{red}{The EEG data from INTENTION was used as training data for the BCI, see section~\ref{BCI} below}.
+Participants' reaction times in INTENTION were used to select stereotypical reaction times for the EXTERNAL condition.
+At the end of INTENTION, EMG electrodes were exchanged for EMS electrodes at the identical location on the forearm.
+
+% \textcolor{green}{The EMG data was used to detect muscle activity onsets, i.e.
+movement onsets.
+The detected onsets were then used to label the data of the \textit{pre-movement} class, see section~\ref{BCI} below for more details.}
 
@@
 The order of the conditions was not pseudo-randomized since training data obtained in INTENTION was required for both EXTERNAL and AUGMENTED.
-Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview.
-
-% \begin{figure*}
-%     \centering
-%     \includegraphics[width=\textwidth]{figures/task_design.png}
-%     \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
-%     \label{fig:task_design}
-% \end{figure*}
-
-\subsection{Apparatus}
-The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected via two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
-To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
-% TODO add links after anonymization is lifted.
-
-\indent\textbf{(1) EMG Recording.} EMG data was recorded from 1 bipolar channel using the BrainAmp ExG amplifier (BrainProducts GmbH, Gilching, Germany).
-The two electrodes were placed above the flexor digitorum profundus with a reference electrode located on the wrist bone.
-EMG data was collected in synchrony with the EEG data through BrainProducts BrainVision Recorder.
-
-\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
-Electrodes were placed according to the extended international 10–20 system \cite{Jasper1983-uw}.
-One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
-After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10$\Omega$.
-EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
-
-We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the data streams available in the network and synchronize the recordings of EEG/EMG data and an experiment marker stream that marked sections of the study procedure.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\columnwidth]{figures/setup.png}
-    \label{fig:setup}
-    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
-\end{figure}
-
-\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
-We utilized the flexor digitorum profundus since we found that we can robustly actuate it without inducing unintended motion of neighboring fingers.
-This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (TENS/EMS Super Duo Plus, prorelax, Düren, Germany).
-The EMS system's output was controlled by flipping a solid state relay (silent) connected via an Arduino Uno (Arduino, Monza, Italy) to the experiment computer.
-The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
-To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure~\ref{fig:setup}.
-% [todo] What were the settings of the stimulation?
-
-\indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
-In addition to the tablet, we used an external keyboard to allow response input of users indicating their timing judgements.
+Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview~\textcolor{green}{immediately after the experience with the prototype.}
 
@@
 EMG amplitudes were band-pass filtered from 20 to 100 Hz and subsequently squared.
-Next, in order to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
+Next, to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
 From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset.
-% bemobil pipeline
+% TODO remove anonymization bemobil pipeline
 
-% TODO Next, noisy data segments were determined. 
+% TODO/Ignore Next, noisy data segments were determined. 
 % A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters.
@@
 The EEG data was band pass filtered from 0.1 to 15 Hz.
-In the first step to prepare the EEG data for classification, noisy data segments were rejected.
+In the first step to prepare the EEG data for \textcolor{red}{classification}~\textcolor{green}{classifier training}, noisy data segments were rejected.
 To this end, the EEGLAB function `autorej' was used, keeping default parameters.
@@
 
-Then, to decrease the dimensionality of the EEG data, a selection of discriminative channels was conducted following a selection approach outlined by~\citep{Schultze-Kraft2021-cu}.
+Then, to decrease the dimensionality of the EEG data, we selected discriminative channels following an approach outlined by~\citep{Schultze-Kraft2021-cu}.
 First, for both \textit{pre-movement} and \textit{idle} the mean signal in the last 100 ms of each segment was subtracted from the mean signal in the first 100 ms of the segment.
@@
 The ranks of the two criteria were joined by summation and the resulting order was saved for later use in training and real-time application of the classifier.
-Lastly, channels C3, C4 and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.
+Lastly, channels C3, C4, and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.
 
 \subsubsection{Training of EEG classifier}
-To classify EEG data a linear discriminant analysis (LDA) with shrinkage regularization (automatic shrinkage using the Ledoit-Wolf lemma~\cite{Ledoit2004-bi}) was trained for each participant individually.
+To classify EEG data, a linear discriminant analysis (LDA) with shrinkage regularization (automatic shrinkage using the Ledoit-Wolf lemma~\cite{Ledoit2004-bi}) was trained for each participant individually.
 As single-trial features, the (linear) slope coefficient was obtained for both \textit{idle} and \textit{pre-movement} segments by fitting a linear regression.
-The classifier was then trained using the slope coefficient feature of the selection of the most discriminative channels as features, hence in the end, generating a feature vector in the dimension of channels that were kept for classification.
-We purposefully constrained the dimensionality of the feature vector to avoid over-fitting.
+The classifier was then trained using the slope coefficient of the selection of the most discriminative channels as features, generating a feature vector in the dimension of channels that were kept for classification.
+We purposefully constrained the dimensionality of the feature vector to avoid over-fitting~\textcolor{green}{and decrease the computational load.}
 
-Using scikit-learn~\cite{Pedregosa2012-sj}, an LDA with automatic shrinkage was cross-validated (using 5-folds) for a grid search from 6 to 20 channels with a step size of adding 2 channels.
+Using scikit-learn~\cite{Pedregosa2012-sj}, the LDA \textcolor{red}{with automatic shrinkage} was cross-validated (using 5-folds) for a grid search from 6 to 20 channels with a step size of adding 2 channels.
 Following this grid search, the cross-validation with the highest accuracy determined the number of channels that were kept for training the model for real-time application.
-Ultimately, to determine the threshold at which, during real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
+Ultimately, to determine the threshold at which, during the real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
 
 \subsubsection{Real-time application and EMS control}
-During real-time application, the EEG data was buffered for the last second for the selected channels.
+During real-time application, the EEG data was buffered for the last second for the selected \textcolor{green}{discriminative channels}.
 The data was band-pass filtered analogously to the training data from 0.1 to 15 Hz.
-Next, the slope was computed for the discriminative channels selected in the classifier cross-validation grid search.
-This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier.
-To smooth the prediction output with the goal of reducing false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the last two predictions with a weighting (.3 and .5).
-Then, at 10Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5s.
+Next, the slope was computed.
+\textcolor{red}{for the discriminative channels selected in the classifier cross-validation grid search.} This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier.
+To smooth the prediction output to reduce false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the \textcolor{red}{last two} \textcolor{green}{current and the preceding prediction with a weighting (weight of .3 for the preceding, and .5 for the current prediction).
+The weights were obtained through trial-and-error during piloting}.
+Then, with a 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s.
 
@@
 To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated.
-At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Hornbaek}.
+After each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
 Following AUGMENTED we interviewed users about their experience working with the system.
-After prompting users to recall their experience and summarize what their task had been we set the focus to the tapping movement and asked them to ignore the time estimation task for the following questions.
+After prompting users to recall their experience and summarize what their task had been, we set the focus to the tapping movement and asked them to ignore the time estimation task in the questions that followed.
 We entered the open part of the interview by asking: ``What did the system do?'' followed up by ``What was the difference between the three conditions''.
@@
 
-% add back in after anonymization: All interviews were manually transcribed and then automatically translated using DeepL\footnote{https://www.deepl.com/translator}
-We analyzed the interviews that always followed the AUGMENTED condition by loosely following~\citet{Mayring2015-pp}.
-All interviews were manually transcribed and [anonymized].
-Prior to screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
-Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e., participants estimate of how well the system was aligned with their intention.
-For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into positive and negative valence sentiments.
+% add back in after anonymization: All interviews were manually transcribed and then automatically translated using }
+We analyzed the interviews \textcolor{red}{that always followed the AUGMENTED condition} by loosely following~\citet{Mayring2015-pp}.
+All interviews were manually transcribed and translated to English using DeepL\footnote{https://www.deepl.com/translator} (DeepL SE, Cologne, Germany).
+Before screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
+Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e., participants' estimate of how well the system was aligned with their intention.
+For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into sentiments with positive and negative valence.
 
@@
 To confirm and demonstrate the discriminative power of the EEG features, we plotted the amplitude time course of electrode Cz between \textit{pre-move} and \textit{idle} epochs.
-Next, the slope coefficients were extracted and a paired t-test was conducted.
-Ultimately, to assess the classifier performance, we report the F1 score, i.e.
-the harmonic mean between of the precision and recall, and plot the ROC.
+\textcolor{green}{Electrode Cz was chosen for exemplary presentation, as it is frequently reported in studies on the RP, since it captures neural activity originating from the sensorimotor cortex.} Next, the slope coefficients were extracted and a paired t-test was conducted.
+Ultimately, to assess the classifier performance, we calculated the F1 score, i.e.
+the harmonic mean of precision and recall, and plotted the ROC\textcolor{green}{, see figure~\ref{fig:EEG_results} a to c.}
 
@@
 A long delay between the EMS trigger and the subsequent screen tap indicated that the stimulation was not strong enough in this trial to lead to muscle actuation resulting in a screen tap.
-Taken together, applying Tukey's methods to these criteria led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
+Taken together, applying Tukey's method \textcolor{green}{to each of these time windows and fusing the rejected trials} \textcolor{red}{criteria} led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
 
@@
 Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
-To test this, we fit a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect.
-As a random effect, we considered the intercepts for participants.
-We obtained p-values of the coefficients by calculating likelihood-ratio tests~\cite{winterLinearModelsLinear2013}.
-All parameters were estimated by maximum likelihood estimation~\cite[see][]{Pinheiro2000}.
-Post-hoc, pairwise contrasts were computed using estimated marginal means with Tukey's methods for multiple comparisons corrections~\cite{Lenth2020-xk}.
+To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect and as a random effect, we considered the intercepts for participants \textcolor{green}{($time interval ~ condition + 1|participantID$)}.
+We obtained p-values of the coefficients by calculating likelihood-ratio tests.
+All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
+Post hoc, \textcolor{green}{we computed pairwise contrasts with the Tukey HSD (Honest Significant Differences), accounting for multiple companions with alpha correction~\cite{Lenth2020-xk}.}
+% ~\cite{winterLinearModelsLinear2013}
 
 Next, we hypothesized that subjective ratings of control over the tap movement are comparable between INTENTION and AUGMENTED and lower in the EXTERNAL condition.
-Again, we fit a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
+Again, we fitted a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
 Coefficients were assessed in the same way as for the intentional binding parameter above.
 
-In short, we tested two main hypotheses with regard to experiencing agency in our three experimental conditions: (1) participants underestimate the tone delay when acting intentionally.
+In short, we tested two main hypotheses with regard to the agency experience in our three experimental conditions: (1) participants underestimate the tone delay when acting intentionally.
 Adding EMS in line with participants' intention to move does not affect this underestimation.
@@
 %%% Resources
+\begin{comment}
+
 % An additional cluster ``takeaways'' was used to cluster other, miscellaneous,  interesting comments.
@@
 Not possible to evlaute false positives as no ground truth available
+
+\end{comment}

### 2023-12-13 input_marked_up/7_discussion.tex
create duplicate marked up version

--- 2023-09-15 input/7_discussion.tex
@@
 \section{Discussion}
-With this paper, we contributed a new way to design augmented experiences with the potential to maintain a user's SoA: a bleeding edge BCI to control EMS at moments of users' intention to interact.
+With this paper, we contributed a new way to design \textcolor{green}{the experience of action augmentation}\textcolor{red}{augmented experiences} with the potential to maintain a user's SoA: a BCI to control EMS at moments of users' intention to interact.
 By leveraging an average of 11.5 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of .7.
@@
 Our system was built using off-the-shelf, affordable, equipment to physically augment users' actions.
-Taken together, all the tech to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 [anonymized].
+Taken together, all technical devices to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 Euros.
 While we used a 64-channel research-grade EEG system, the channel selection procedure resulted in the system ultimately using a low-density coverage of about 12 channels on average for classification.
@@
 Still, the system achieved what is considered to be a `good' F1 score of .7, and hence sometimes detected users' intention to interact.
-In our user study, this system performance affected users' performance in the time estimation task when comparing the three experimental conditions.
-We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency with respect to intention-based augmentation.
-We believe this to be a very promising finding considering the fact that our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+In our user study, this system performance affected users' \textcolor{green}{tendency for temporal binding}\textcolor{red}{performance in the time estimation task} when comparing the three experimental conditions.
+We conclude that, taking users' sentiments and ratings of control into account, this indicates a general modulation of agency for our intention-based augmentation.
+We believe this to be a very promising finding because our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+
+\textcolor{green}{Our system ran at a 10 Hz update rate.
+This was chosen in order to avoid buffer overflow due to the fact that the end-to-end computing latency from measuring the physiological signal to outputting the predicted class and probability took about 40 ms.
+The RP is reported to occur several hundred milliseconds prior to the physical movement~\cite{Schurger2021-vp}.
+Therefore, our augmentation should have resulted in a motion preemption, considering subtracting the computing latency and a delay that the EMS stimulation takes to affect the muscle from the RP-based intention detection.
+However, it is difficult to assess whether participants movements were in fact pre-empted as no label exists for the `true' intention onset to compare the classifier to.
+That participants were able to act on their own volition, and hence no labelling of their movement intention could be derived, is the key distinction to related works using controlled stimulus-response paradigms.}
+
+\textcolor{green}{However, we learned from these paradigms that increasing participants reaction times by about 80 ms is optimal with regards to maintaining SoA~\cite{Kasahara2019-sk}.
+On one hand our system is capable of delivering pre-emption within this `SoA-optimal' time range.
+On the other hand however, the performance of the sub-optimal performance of the classifier means a lot of variation due to false positive detections.
+While in some cases, as we observed in the users' comments, the stimulation may have pre-empted a user's motion, false positive stimulation may have had a significant impact on the overall impression of the system, see section~\ref{improving_classifier} below.}
 
 In line with the literature, we found that participants estimated the delay between tap and tone to be lower in both conditions where they were instructed to act on their own volition as compared to the passive condition~\cite{Moore2012-dk}.
-More recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
+We point out that recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
 However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the system.
@@
 \subsection{Limitations}
-Two main procedural issues arose with respect to the time estimation task: First, contrasting AUGMENTED with INTENTION, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
+Two main procedural issues arose concerning the time estimation task: First, contrasting AUGMENTED with INTENTION, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
 This may have introduced additional variance in the time estimation task since the exact moment of the tap that causes the tone is more obscure.
@@
 However, we observed that many participants, while perceiving a continuous distribution of the delays, did not answer at a continuous ms resolution but rather at steps of 50 or 100 ms, skewing the distribution of their answers.
-While many different versions of the intentional binding paradigm exist~\cite{Moore2012-dk}, we would now choose a continuous slider, perhaps set at different initial positions between trials, to obtain participants' estimates.
+While many different versions of the intentional binding paradigm exist~\cite{Moore2012-dk}, we would choose a continuous slider for future experiments with different initial positions over trials to reduce the bias in participants' estimates.
 
-\subsubsection{Improving the Prediction of the Intent to (Inter-) act}
+\subsubsection{Improving the Prediction of the Intent to (Inter-) act}\label{improving_classifier}
 While an F1 score of .7 may be considered `good' for many classification scenarios, when aiming to elicit a feeling of control this level of performance may likely not be good enough, see~\citet{Papenmeier2022-oi} for a review.
-Balancing false positive rate (FPR) and true positive rate (TPR) appropriately may prove crucial in order to elicit an experience of agency.
+Balancing false positive rate (FPR) and true positive rate (TPR) appropriately may prove crucial to elicit an experience of agency.
 
@@
 
-As a consequence, the insights gained into the system's potential to preserve a sense of agency are, regrettably, limited.
+As a consequence, the insights gained into the system's potential to preserve a sense of agency are limited.
 The quantitative metrics employed in our study, specifically those measuring the sense of control (including temporal binding and item assessment), may not have captured the nuanced temporal alignment experiences associated with EMS augmentation and user movement intentions but rather a more holistic assessment of the entire experimental block, which included trials with missed stimulation, too early stimulation, and some trials where the stimulation was in alignment with the participants' intent.
@@
 
-% \subsubsection{Opportunities}
+% Conclusion
+\section{Conclusions}
+In this paper, we demonstrated a system to maintain users' SoA during augmented experiences using brain signals reflecting the intent to (inter-) act.
+In our user study, we found that intentional binding effects were stronger when participants worked with the system to tap on a screen compared to when they were passively moved.
+Furthermore, the level of control working with the system was higher than when being passively moved, as indicated by subjective responses.
+
+We believe this to be an important next step towards augmented users with full integration of the technology~\cite{Mueller2020-dl}.
+Ultimately, our closed-loop system enables novel predictive interfaces that leverage the users' body directly yet still feel \textit{natural} as they align with the users' intention: For example, altering the affordance structure of the interaction in real-time~\cite{Gehrke2022-kz, Lopes2015-ze, Nataraj2020-wm}.
+
+With AI technologies becoming more and more ubiquitous in our everyday lives, questions about how agency is shared between us humans and AI technologies arise regularly.
+With the likely future that these systems move directly onto our bodies, alignment with users' intentions will be the \textit{key} component driving their adoption.
+
 \begin{comment}
@@
 \end{comment}
-
-% Conclusion
-\section{Conclusions}
-In this paper, we demonstrated a system to maintain users' SoA during augmented experiences using brain signals reflecting the intent to (inter-) act.
-In our user study, we found that intentional binding effects were stronger when participants worked with the system to tap on a screen compared to when they were passively moved.
-Furthermore, the level of control working with the system was higher than when being passively moved, as indicated by subjective responses.
-
-We believe this to be an important next step towards augmented users with full integration of the technology~\cite{Mueller2020-dl}.
-Ultimately, our closed-loop system enables novel predictive interfaces that leverage the users' body directly yet still feel \textit{natural} as they align with the users' intention: For example, altering the affordance structure of the interaction in real-time~\cite{Gehrke2022-kz, Lopes2015-ze, Nataraj2020-wm}.
-
-With AI technologies becoming more and more ubiquitous in our everyday life, questions about how agency is shared between us humans and these AI technologies arise on a regular basis.
-With the likely future that these systems move directly onto our bodies, alignment with users' intentions will be the \textit{key} component driving their adoption.

### 2023-12-13 input/rebuttal/rebuttal_letter.tex
revision

--- 2023-12-13 input_marked_up/rebuttal/rebuttal_letter.tex
@@
+We want to thank the reviewers for their critical, and helpful, comments.
+We thoroughly revised the paper and believe that the contribution of our BCI based muscle augmentation is a very valuable contribution to the CHI community.
+
 \section{comparison with baseline conditions}
@@
 The major motivation for the current work was to present a closed-loop augmentation system using neural input that is entirely decoupled from any outside constraints.
-Ultimately, we aim at designing a system that a user can wear and use in a variety of settings and applications as no knowledge about the user's environment should b enecessary for it to work.
-With this goal in mind, we did not find a useful baseline condition from previous studies to compare our interface to.
+Ultimately, we aim at designing a system that a user can wear and use in a variety of settings and applications as no knowledge about the user's environment should be necessary for it to work.
+With this goal in mind, we did not find a useful heuristic to compare our interface to.
 We believe the simple case of random activation, for example 'send 10 EMS pulses within every 5 minutes' to not be a useful heuristic since we have shown that our classifier performs above chance level (F1 score of .7).
@@
 The focus of the revised manuscript is now on SoA and is not deferred as much anymore by a secondary aspect of the system, i.e., a reduction in reaction times.
-In the revised manuscript we tightened the focus and framing on SoA and moved the part on previous studies reducing reaction times from the introduction section to related work.
 This better fits with the direct contrast of using predictive models (based on physiological signals) instead of controlled environments for action augmentation that falls in line with a user's intentions.

### 2023-12-13 input/2_introduction.tex
revision

@@
 \section{Introduction}
-% 1. Describe the Current Situtation
-% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
+
 Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw, Kunze2017-co}.
-\textcolor{green}{These technological advances include the miniaturization of the actuating hardware to wearable form factors and the direct sensing and stimulation capabilities of neural interfaces.
+These technological advances include the miniaturization of the actuating hardware to wearable form factors and the direct sensing and stimulation capabilities of neural interfaces.
 Especially due to these characteristics, recent perspectives promote a change of the computing era from human-computer \textit{interaction} to \textit{integration}~\cite{Mueller2020-dl}.
-One key change in perspective is, that \textit{integrated} user's share agency with the computing machinery to execute tasks.
-This is a critical distinction as \textit{integration} technologies are designed to directly influence people’s bodies, their actions, and the resulting action outcomes.}
+One key change in perspective is, that \textit{integrated} users share agency with the computing machinery to execute tasks.
+This is a critical distinction as \textit{integration} technologies are designed to directly influence people’s bodies, their actions, and the resulting action outcomes.
 
-\textcolor{green}{Besides body and outcome augmentation, action augmentation has been defined as the case where a ``system assists the user’s action to produce the intended outcome.''~\cite{Cornelio2022-aq}.
-While such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer, when it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of ones body.
-Unfortunately,} such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
+Besides body and outcome augmentation, action augmentation has been defined as the case where a ``system assists the user’s action to produce the intended outcome.''~\cite{Cornelio2022-aq}.
+Such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer.
+When it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of one's body.
+Unfortunately, such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
 
-% 2. What is the complication, challenge identified
-% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
 Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us.
 It has been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}.
-Hence, a key challenge to drive the adoption of human action augmentation then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
-
-% 3. Question
-% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
-% What have others done to address this and what do we propose "instead"
-\textcolor{red}{One goal of green `on-body' action augmentation is to increase reaction times, thereby making an interaction more efficient.
-Previous work on such augmentation scenarios has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk}.
-Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
-Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.}
-
-\textcolor{red}{In experimental settings, this is typically achieved by using strongly controlled `stimulus-response' paradigms.
-For example, reaction times when tapping on a touchscreen in response to a presented stimulus on screen, can be reduced by an augmentation device while maintaining SoA for the user~\cite{Kasahara2019-sk, Kasahara2021-gy}.
-In such controlled scenarios, the timing of the action augmentation can be tuned to be near optimal because the user behavior can be predicted with very high certainty to follow a previously presented stimulus.
-However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
-In other words, how can we design a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions?}
-
-% [4. (short) Answer Teaser
-% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
-% keep this very short]
+Hence, a key challenge to drive the adoption of human action augmentation is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
 
 \subsection{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
-% 4. (long) Anwser with teaser image
-% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
 In this paper, we present a prototype to overcome the disruption in experiencing agency during action augmentation.
@@
 
-To control the interaction in real-time, an RP-based classifier distinguished between two user states: \textit{idle}, reflecting the absence of an intent to act, and \textit{pre-movement}, indicating the presence of an intent to act.\textcolor{red}{When the system predicted a \textit{pre-movement} state, the user's action was augmented, potentially even before their own voluntary motor command.
-In our user study, we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their SoA.} \textcolor{green}{During \textit{idle}, participants were passively looking at a fixation cross.
+To control the interaction in real-time, an RP-based classifier distinguished between two user states: \textit{idle}, reflecting the absence of an intent to act, and \textit{pre-movement}, indicating the presence of an intent to act.
+During \textit{idle}, participants were passively looking at a fixation cross.
 Instead, during \textit{pre-movement}, participants were instructed to voluntarily initiate a tap on a touchscreen whenever they felt the urge to do so.
-Previous work has indicated that the RP emerges during formation of conscious intention and is specific to voluntary action~\cite{Schultze-Kraft2020-rm, Travers2020-hf, Pares-Pujolras2019-ll}.}
+Previous work has indicated that the RP emerges during formation of conscious intention and is specific to voluntary action~\cite{Schultze-Kraft2020-rm, Travers2020-hf, Pares-Pujolras2019-ll}.
 
-\textcolor{green}{Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially preceding their voluntary motor command.
+Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially preceding their voluntary motor command.
 This augmentation moved the ring finger in accordance with the user's intention to act.
 We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle.
-In our user study, a mixed-methods research approach was employed to explore whether aligning the physical impact on the user's body with their intention to move preserved their Sense of Agency (SoA).}
-
-%%% Resources
-\begin{comment}
-% TODO merge this with the figure below
-% \missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm.
-<- better use this for the teaser image.
-Here better present the main results, i.e.
-SoA scores?
--> Klaus comment: yes, but they don't have to be exclusive.
-The general setup with the LRP could appear with the main results both in the teaser or here?
-But the teaser would be more a visualization of the principle, no?
-That does not necessarily require real results but could show a hyperplane and data dots for examplification.}
-% make this as an infographic with comic style drawing of augmented user and then the three pathways drawn
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\columnwidth]{figures/setup_conditions_draft_small.png}
-%     \caption{draft apperatus and conditions}
-%     \label{fig:setup}
-% \end{figure}
-% \begin{figure}[!h]
-%     \centering
-%     \includegraphics[width=\columnwidth]{figures/augmented_com_model_colour.png}
-%     \caption{draft - visualization of comparator model of the SoA in the case of augmented interactions (inspired by Dewey 2019)}
-%     \label{fig:com_model}
-% \end{figure}
-
-% add results and general outcomes here?
-
-% for now moved to introduction section
-% We follow classical work in the neuroscience describing SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay and reproduce it as such.
-Critically, this temporal compression only occurs following endogenous movement intention.
-The action outcome is mentally \textit{binded} to the intention.
-To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay <REF>.
-
-% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.
-
-% By controlling the stimulus presentation, the experimenter can control the timing of the \textit{pre-emptive} gain of the user's action augmentation.
-
-% Resources
-% The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited.
-However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy.
-
-% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.
-
-% if something happens for you in an interaction there is a cost to that:
-% - impacts SoA which in turn may makes user's engage less -> citation?
-% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?
-
-\end{comment}
+In our user study, a mixed-methods research approach was employed to explore whether aligning the physical impact on the user's body with their intention to move preserved their SoA.

### 2023-12-13 input/3_related_work.tex
revision

@@
 \subsection{Theories of Sense of Agency}
-The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: when we intentionally perform an action, the brain generates sensory predictions about the action outcome.
+The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: When we intentionally perform an action, the brain generates sensory predictions about the action outcome.
 These predictions are constantly compared to the actual sensory data available through the sensory system during the execution of the action.
@@
 As opposed to intended actions using our own body, movement augmentation hardware allows moving a user's body without their intention.
-Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e.
-exoskeletons, a user's body can be moved by applying forces to the extremities.
+Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e., exoskeletons, a user's body can be moved by applying forces to the extremities.
 Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS).
@@
 With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
-~\textcolor{red}{Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz}.}
-%, see figure~\ref{fig:task_design} EXTERNAL CONTROL)}
 
@@
 
-\subsection{\textcolor{green}{Controlling Actuated Haptic Experiences}}
+\subsection{Controlling Actuated Haptic Experiences}
 
-\textcolor{green}{Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA frequently use highly controlled `stimulus-response' paradigms.
+Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA frequently use highly controlled `stimulus-response' paradigms.
 For example, scenarios where participants are instructed to tap on a touchscreen in response to a presented stimulus on the screen.
@@
 Hence, \textit{pre-empting} the user's motion can be designed to fall in line with their intention to move, thereby maintaining SoA.
-Previous work on such an augmentation scenario has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.}
+Previous work on such an augmentation scenario has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.
 
-\textcolor{green}{Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
+Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
 Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.
 However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
-In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?}
+In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?
 
-\textcolor{red}{\subsection{Controlling Actuated Haptic Experiences using Brain Signals Reflecting the Intent to (Inter-)act}}
-
-\subsubsection{\textcolor{green}{Using Brain Signals Reflecting the Intent to (Inter-)act for Action Augmentation}}
-\textcolor{green}{One possible design solution is to leverage physiological signals for action augmentation}.
-Of the possible physiological signals that can be leveraged\textcolor{red}{used to further investigate the experience of agency}, the EEG is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
+\subsubsection{Using Brain Signals Reflecting the Intent to (Inter-)act for Action Augmentation}
+One possible design solution is to leverage physiological signals for action augmentation.
+Of the possible physiological signals that can be leveraged, the EEG is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
 The RP, or \textit{lateralized readiness potential}, is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
-The RP is reliably observed at electrodes placed over the \textcolor{green}{sensori}motor cortex contralateral to the acting hand.~\textcolor{green}{In the 10-20 system for EEG electrode placement~\cite{Jasper1983-uw}, these are electrode C3 located over the sensorimotor cortex of the left hemisphere, and C4 vice versa.
-However, activity observed at electrode Cz is reported most frequently as it reflects neural activity originating from the sensorimotor cortex without lateralization bias.} Since the RPs' measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
+The RP is reliably observed at electrodes placed over the sensorimotor cortex contralateral to the acting hand.In the 10-20 system for EEG electrode placement~\cite{Jasper1983-uw}, these are electrode C3 located over the sensorimotor cortex of the left hemisphere, and C4 vice versa.
+However, activity observed at electrode Cz is reported most frequently as it reflects neural activity originating from the sensorimotor cortex without lateralization bias.
+Since the RPs' measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective.
 However, evidence abounds for its role in action preparation.
@@
 After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot's appearance on screen.
-
-\textcolor{red}{In the present study, we leveraged the RP in a similar way, i.e.
-to discriminate between two \textit{action} states as the basis to trigger an action augmentation device or not.
-The technology we used with our system, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.}
-
-%%% resources
-\begin{comment}
-
-(see Carruthers 2012 for a summary of arguments against this model)
-
-There is an ongoing debate, whether the temporal attraction is specific to intentional movements, or is more generally related to the perception between action and outcome (Buhner 2012)
---> they suggest intentional action is not necessary for temporal binding, but that the binding results from a causal relation linking actions with consequences -- more general causal binding -- do we need to go there?
-
-One idea: - reach "we-mode" in human-machine joint control ---> decode human intention (Zander 2011, Felke 2019, Shiskin 2016)
-- maintain sense of agency in augmented interactions 
-- doing this by leveraging action intention in brain
-
-Experiencing control is an important factor in human-computer interaction (see Shneiderman and Nielsen).
-
-%%%%%%%%%% theory on sense of agency  / neural basis of sense of agency
-
-% explicit measures - reflective (short paragraph)
-% - questionnaires
-
-%% from bergström 22
-- read 18, 37 from  Bergstöem 
-- 14,14,19.31,39,41,48
-11,20,33
-
-Sense of control is mostly measured with simple items with ratings like "It felt like I was in control of the hand I was looking at” rated on a 7-point Likert scale, +3 indicating strong agreement and −3 indicating strong disagreement (Longo and Haggard) or  "Indicate how much it felt like moving the joystick caused the object on the computer screen to move” as a measure of explicit sense of agency, also rated on a 7-point scale" (11)
-item asking participants to indicate the degree of control they have felt over the changes on the screen [1]. We used this item also as an agency measure (1)
-
-Researchers use implicit and explicit methods.
-Explicit methods involve directly asking participants to report their subjective experiences using items rated on Likert scales or percentages.
-These items, such as "It felt like I was in control of the hand I was looking at" (Longo & Haggard) and "Indicate how much it felt like moving the joystick caused the object on the computer screen to move" (Ebert & Wegner), may focus on either the action element or the outcome element of the sense of agency (see Moore).
-
-Items such as ""It felt like I was in control of the hand I was looking at” (longo Haggard) and "Indicate how much it felt like moving the joystick caused the object on the computer screen to move” (Ebert und Wegner).
-Notably, such items differ between focusing either on the action element or the outcome element (See MOore)
-
-% - phenomenological interview
-
-% implicit measures - pre-reflective (longer paragraph)
--subjective measures address a high level-SoA  based on subjective judgments of feeling of control (Barlas and Kopp)
-- proxy for the low-level SoA as they do not require conscious reflection on one’s SoA (Synofzik et al., 2008a; Desantis et al., 2011; Moore and Obhi, 2012; Barlas and Obhi, 2014).“ ([Barlas und Kopp, 2018, p.
-2]
-- one such measure is intentional binding
-
-% - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here
-- „refers to the perceived temporal attraction between voluntary actions and their outcomes (Haggard et al., 2002).
-More clearly, the temporal interval between actions and outcomes is perceived as shorter when these outcomes are produced by voluntary actions compared to when they follow, for instance, involuntary movements or external causes (Haggard et al., 2002).“ ([Barlas und Kopp, 2018, p.
-2]
-- While details about the relationship between measures of intentional binding and SoA are far from clear, they are extensivly used 
-- alternatives to intentional binding - sensory attenuation / visual attention
-- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
-it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)
-
-\textit{Intentional binding} is the phenomenon of subjectively compressing the time interval between a voluntary action and the sensory consequences of that action~\cite{Moore2012-ic}.
-- comparisons between active and passive finger tab 
-    Kühn 2013 + Engbert 2007 : experimenter presses the finger down in a passive condition
-
-When a voluntary action is causally linked with a sensory outcome, the action and its consequent effect are perceived as being closer together in time.
-This effect is called intentional binding.“ (from Jo 2014)
-
-- Read More 2009 (interval estimations were shorter (i.e., stronger binding) in voluntary than involuntary movements) 
-- add figuure action binding and effect binding (e.g.  in WEN and Imamizu)
-
-!! new paper suggesting that temporal binding / intentional binding is not a measure of sense of agency !!(Gutzeit!)
-
-% connection between implicit and explicit measures:
-- intentional binding effects are sometimes discrepant from explicit judgment of agency (see 30-33 in Wen and Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
-- Intentional binding is e.g. influenced by causal relations between events, attention, and arousal (see Wen and Imamizu)
-
-% Interesting HCI papers:
-% PosssessedHand, Gilbert-I miss being me,
-
-% Why is sense of agency important? Why would it be bad if we lost it in augmented humans?
-% However, for augmentation hardware to be readily accepted, maintaining a sense of agency is important for several reasons. (find reasons)
-% - motor learning ( check Wen, W. et al. Perception and control: individual difference in the sense of agency is associated with learnability in sensorimotor adaptation. Sci. Rep. 11 , 20542 (2021).
-% - motivation (find sources)
-
-%%%%%%%%%%%%%%%%%%%%%%
-Other
-- EEG patterns related to readiness potential and intentional  binding sometimes show disceprant patterns ( see WEN and Imamizu 38 und 39) (wittmann 2014, Goldberg 2017)
-
-% This straightforward relationship is complicated by the introduction of technology.
-Think about, playing a computer game where a key press controls an avatar playing the piano.
-Here, we do not really press the piano key, the avatar's movement, and the tone are now generated by a computer.
-Despite the blurring lines between the performer and the cause of the tone, we still feel a sense of authorship when the comparison does not reveal incongruency.
-This lingering feeling of control is thought to be elicited by our intentions being reflected in the resulting action (sources needed).
-
-% alternative headings:
-% Controlling Actuated Haptic experiences using brain signals reflecting the intent to interact
-% Triggering EMS using EEG Signals
-
-\end{comment}

### 2023-12-13 input/5_user_study_and_methods.tex
revision

@@
 With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user's own brain signals serve as the control signal to a physical end effector.
-To this end, we assessed whether our prototype preserved agency using a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.
-% todo add some more framing here to make it easier to step through the methods, maybe: For this paper we designed a prototype that lies between the two edge cases -- no intention and control to full control -- and investigate user's sense of agency using both explicit and implicit measures.
+Therefore, we conducted a user study comparing three conditions of agency experience.
+With the first two conditions, INTENTION and EXTERNAL we queried users at the two edges of agency experience.
+In EXTERNAL participants had no intention to move and no control over their movement.
+On the other hand, in INTENTION, participants were acting as they would in their day-to-day lives, fully in control.
+We then investigated in a third condition where our AUGMENTED prototype sits on the spectrum between INTENTION and EXTERNAL, in other words, whether it preserved agency and to what degree.
+To answer our question, we employed a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.
 
@@
 One participant had to be excluded from further data analyses due to significant deviations from the instructions in the execution of the task.
-%the Department of Psychology and Ergonomics at TU Berlin (tracking number: BPN\_GEH\_2\_230130220421).
+Precisely, they did not initiate their movements at their own volition but rather immediately at the onset of each trial, thereby violating the task instruction of waiting 2-3 s after trial onset, see task description below for more detail.
+
+\subsection{Apparatus}
+The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected to two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
+To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
+% TODO add links after anonymization is lifted.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/setup_fixed.png}
+    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
+    \label{fig:setup}
+\end{figure}
+
+\indent\textbf{(1) EMG Recording.} EMG data was recorded from 1 bipolar channel using the BrainAmp ExG amplifier (BrainProducts GmbH, Gilching, Germany).
+The two electrodes were placed above the flexor digitorum profundus with a reference electrode located on the wrist bone.
+EMG data was collected in synchrony with the EEG data through BrainProducts' BrainVision Recorder.
+EMG data was only recorded for the first experimental condition to obtain labels for classifier training.
+After that, the EMG electrodes were changed for EMS electrodes.
+
+\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes referenced to electrode FCz in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
+Electrodes were placed according to the extended international 10–20 system \cite{Jasper1983-uw}.
+One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10k$\Omega$.
+EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
+
+We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the data streams available in the network and synchronize the recordings of EEG/EMG data and an experiment marker stream that marked sections of the study procedure.
+
+\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
+We utilized the flexor digitorum profundus since we found that we can robustly actuate it without inducing unintended motion of neighboring fingers.
+This finger actuation was achieved via a medically compliant battery-powered muscle stimulator (TENS/EMS Super Duo Plus, prorelax, Düren, Germany).
+The EMS system's output was controlled by flipping a solid state relay (silent) connected via an Arduino Uno (Arduino, Monza, Italy) to the experiment computer.
+The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
+To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure~\ref{fig:setup}.
+
+\indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
+In addition to the tablet, we used an external keyboard to allow users to input their timing judgments.
 
@@
 Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED.
-% , see figure~\ref{fig:task_design}.
 
@@
 
-\indent\textbf{INTENTION.} The task went as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) then, they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen.
-In line with the literature on the origin of the RP generating process they were told ``to avoid pre-planning the movement, avoid any obvious rhythm over trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
+\indent\textbf{INTENTION.} The task was as follows: (1) a fixation cross appeared on a tablet screen and participants were instructed to rest and wait until it disappeared; (2) they were instructed to wait for a brief moment (2 to 3s), before (3) initiating their movement and tap the screen, see figure~\ref{fig:progression}.
+In line with the literature on the origin of the RP generating process, they were told ``to avoid pre-planning the movement, avoid any obvious rhythm across trials, and to press when they felt the spontaneous urge to move''~\cite{Schultze-Kraft2021-cu}.
 (4) After the screen was tapped, a tone was played at a pseudo-random delay of 200, 350, or 500ms.
-Participants were now asked to estimate the delay, typing in their answer on a number pad of an attached keyboard.
+Participants were now asked to estimate the delay, typing in their answers on a number pad of an attached keyboard.
 After confirming their answer by hitting the return key, the next trial started.
-During INTENTION, participants were equipped with Electromyogram (EMG) sensors instead of EMS electrodes.
-INTENTION data was used as training data for the BCI, see section~\ref{BCI} below, as well as to select stereotypical reaction times in the EXTERNAL condition.
+
+During INTENTION, participants were equipped with EMG sensors instead of EMS electrodes.
+Participants' reaction times in INTENTION were used to select stereotypical reaction times for the EXTERNAL condition.
+At the end of INTENTION, EMG electrodes were exchanged for EMS electrodes at the identical location on the forearm.
 
@@
 The order of the conditions was not pseudo-randomized since training data obtained in INTENTION was required for both EXTERNAL and AUGMENTED.
-Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview.
+Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview immediately after the experience with the prototype.
 
-% \begin{figure*}
-%     \centering
-%     \includegraphics[width=\textwidth]{figures/task_design.png}
-%     \caption{(a) description of the experimental conditions, (b) their hypothesized effects on sense of agency from the perspective of a comparator model, is inspired by~\citet{Dewey2019-dg} (c) depicts the flow of movement generation in the participant and the system between conditions.}
-%     \label{fig:task_design}
-% \end{figure*}
-
-\subsection{Apparatus}
-The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected via two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
-To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
-% TODO add links after anonymization is lifted.
-
-\indent\textbf{(1) EMG Recording.} EMG data was recorded from 1 bipolar channel using the BrainAmp ExG amplifier (BrainProducts GmbH, Gilching, Germany).
-The two electrodes were placed above the flexor digitorum profundus with a reference electrode located on the wrist bone.
-EMG data was collected in synchrony with the EEG data through BrainProducts BrainVision Recorder.
-
-\indent\textbf{(2) EEG Recording.} EEG data was recorded from 64 actively amplified Ag/AgCl electrodes in an actiCap Snap cap using BrainAmp DC amplifiers from BrainProducts.
-Electrodes were placed according to the extended international 10–20 system \cite{Jasper1983-uw}.
-One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
-After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10$\Omega$.
-EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
-
-We used LSL\footnote{https://github.com/sccn/labstreaminglayer} to make the data streams available in the network and synchronize the recordings of EEG/EMG data and an experiment marker stream that marked sections of the study procedure.
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\columnwidth]{figures/setup.png}
-    \caption{Experimental setup of measurement-- and input devices (image with consent from participant).}
-    \label{fig:setup}
-\end{figure}
-
-\indent\textbf{(3) Electrical Muscle Stimulation.} We actuated the ring finger via EMS, which was delivered with two electrodes attached to the participants' flexor digitorum profundus muscle.
-We utilized the flexor digitorum profundus since we found that we can robustly actuate it without inducing unintended motion of neighboring fingers.
-This finger actuation was achieved via a medically-compliant battery powered muscle stimulator (TENS/EMS Super Duo Plus, prorelax, Düren, Germany).
-The EMS system's output was controlled by flipping a solid state relay (silent) connected via an Arduino Uno (Arduino, Monza, Italy) to the experiment computer.
-The EMS was pre-calibrated by the participant to ensure a pain-free but effective stimulation and robust actuation leading to an `immediate' tap on the touchscreen after actuation.
-To ensure a comfortable experimental experience so that participants relaxed their arm musculature as much as possible, a custom built hand rest (support device) was placed on top of the touchscreen for participants, see figure~\ref{fig:setup}.
-% [todo] What were the settings of the stimulation?
-
-\indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
-In addition to the tablet, we used an external keyboard to allow response input of users indicating their timing judgements.
-
-% alternative headings:
-% Brain-Computer Interface to Detect EEG Readiness Potentials }
-% Classifiying EEG Signals of the Intent to Interact
 \subsection{Brain-Computer Interface}\label{BCI}
@@
 EMG amplitudes were band-pass filtered from 20 to 100 Hz and subsequently squared.
-Next, in order to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
-From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset.
-% bemobil pipeline
+Next, to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
+From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset, see~\ref{fig:method} a.
+% TODO remove anonymization bemobil pipeline
 
-% TODO Next, noisy data segments were determined. 
+% TODO/Ignore Next, noisy data segments were determined. 
 % A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters.
@@
 The EEG data was band pass filtered from 0.1 to 15 Hz.
-In the first step to prepare the EEG data for classification, noisy data segments were rejected.
+In the first step to prepare the EEG data for classifier training, noisy data segments were rejected.
 To this end, the EEGLAB function `autorej' was used, keeping default parameters.
@@
 The ranks of the two criteria were joined by summation and the resulting order was saved for later use in training and real-time application of the classifier.
-Lastly, channels C3, C4 and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.
+Lastly, channels C3, C4, and Cz were moved to the top of the order for every participant as these are most frequently reported in studies on the RP.
+
+\begin{figure*}[!h]
+    \centering
+    \includegraphics[width=\textwidth]{figures/method.png}
+    \caption{Diagram summarizing: (a) signal processing and event labeling from muscle activity, (b) extraction of linear slope features from labeled brain signals and (c) real-time application of the BCI gating the muscle stimulation hardware.}
+    \label{fig:method}
+\end{figure*}
 
 \subsubsection{Training of EEG classifier}
-To classify EEG data a linear discriminant analysis (LDA) with shrinkage regularization (automatic shrinkage using the Ledoit-Wolf lemma~\cite{Ledoit2004-bi}) was trained for each participant individually.
+To classify EEG data, a linear discriminant analysis (LDA) with shrinkage regularization (automatic shrinkage using the Ledoit-Wolf lemma~\cite{Ledoit2004-bi}) was trained for each participant individually.
 As single-trial features, the (linear) slope coefficient was obtained for both \textit{idle} and \textit{pre-movement} segments by fitting a linear regression.
-The classifier was then trained using the slope coefficient feature of the selection of the most discriminative channels as features, hence in the end, generating a feature vector in the dimension of channels that were kept for classification.
-We purposefully constrained the dimensionality of the feature vector to avoid over-fitting.
+The classifier was then trained using the slope coefficient of the selection of the most discriminative channels as features, generating a feature vector in the dimension of channels that were kept for classification, see~\ref{fig:method} b.
+We purposefully constrained the dimensionality of the feature vector to avoid over-fitting and decrease the computational load.
 
-Using scikit-learn~\cite{Pedregosa2012-sj}, an LDA with automatic shrinkage was cross-validated (using 5-folds) for a grid search from 6 to 20 channels with a step size of adding 2 channels.
+Using scikit-learn~\cite{Pedregosa2012-sj}, the LDA was cross-validated (using 5-folds) for a grid search from 6 to 20 channels with a step size of adding 2 channels.
 Following this grid search, the cross-validation with the highest accuracy determined the number of channels that were kept for training the model for real-time application.
-Ultimately, to determine the threshold at which, during real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
+Ultimately, to determine the threshold at which, during the real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
 
 \subsubsection{Real-time application and EMS control}
-During real-time application, the EEG data was buffered for the last second for the selected channels.
+During real-time application, the EEG data was buffered for the last second for the selected discriminative channels.
 The data was band-pass filtered analogously to the training data from 0.1 to 15 Hz.
-Next, the slope was computed for the discriminative channels selected in the classifier cross-validation grid search.
+Next, the slope was computed.
 This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier.
-To smooth the prediction output with the goal of reducing false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the last two predictions with a weighting (.3 and .5).
-Then, at 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s.
+To smooth the prediction output to reduce false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging current and the preceding prediction with a weighting (weight of .3 for the preceding, and .5 for the current prediction).
+The weights were obtained through trial-and-error during piloting.
+Then, with a 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s, see~\ref{fig:method} c.
 
@@
 To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated.
-At the conclusion of each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
+After each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
 Following AUGMENTED we interviewed users about their experience working with the system.
-After prompting users to recall their experience and summarize what their task had been we set the focus to the tapping movement and asked them to ignore the time estimation task for the following questions.
+After prompting users to recall their experience and summarize what their task had been, we set the focus to the tapping movement and asked them to ignore the time estimation task in the questions that followed.
 We entered the open part of the interview by asking: ``What did the system do?'' followed up by ``What was the difference between the three conditions''.
@@
 
-% add back in after anonymization: All interviews were manually transcribed and then automatically translated using DeepL\footnote{https://www.deepl.com/translator}
-We analyzed the interviews that always followed the AUGMENTED condition by loosely following~\citet{Mayring2015-pp}.
-All interviews were manually transcribed and [anonymized].
-Prior to screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
-Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e., participants estimate of how well the system was aligned with their intention.
-For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into positive and negative valence sentiments.
+% add back in after anonymization: All interviews were manually transcribed and then automatically translated using }
+We analyzed the interviews by loosely following~\citet{Mayring2015-pp}.
+All interviews were manually transcribed and translated to English using DeepL\footnote{https://www.deepl.com/translator} (DeepL SE, Cologne, Germany).
+Before screening the texts, two experts clustered the responses into 3 clusters: First, ``Functionality'' referred to what participants attributed the source of the stimulation.
+Next, we clustered responses according to ``Guessed percentage'' of correct interaction, i.e., participants' estimate of how well the system was aligned with their intention.
+For the cluster ``Correct Interaction'', we specifically queried participants to recall the moments where the stimulation felt in line with their intention to move, then we clustered their responses into sentiments with positive and negative valence.
 
@@
 To confirm and demonstrate the discriminative power of the EEG features, we plotted the amplitude time course of electrode Cz between \textit{pre-move} and \textit{idle} epochs.
+Electrode Cz was chosen for exemplary presentation, as it is frequently reported in studies on the RP, since it captures neural activity originating from the sensorimotor cortex.
 Next, the slope coefficients were extracted and a paired t-test was conducted.
-Ultimately, to assess the classifier performance, we report the F1 score, i.e.
-the harmonic mean between of the precision and recall, and plot the ROC.
+Ultimately, to assess the classifier performance, we calculated the F1 score, i.e.
+the harmonic mean of precision and recall, and plotted the ROC, see figure~\ref{fig:EEG_results} a to c.
 
@@
 A long delay between the EMS trigger and the subsequent screen tap indicated that the stimulation was not strong enough in this trial to lead to muscle actuation resulting in a screen tap.
-Taken together, applying Tukey's methods to these criteria led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
+Taken together, applying Tukey's method to each of these time windows and fusing the rejected trials led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
 
@@
 Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
-To test this, we fit a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect.
-As a random effect, we considered the intercepts for participants.
+To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect and as a random effect, we considered the intercepts for participants ($time interval ~ condition + 1|participantID$).
 We obtained p-values of the coefficients by calculating likelihood-ratio tests.
-All parameters were estimated by maximum likelihood estimation~\cite[see][]{Pinheiro2006-bk}.
-Post-hoc, pairwise contrasts were computed using estimated marginal means with Tukey's methods for multiple comparisons corrections~\cite{Lenth2020-xk}.
-% ~\cite{winterLinearModelsLinear2013}
+All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
+Post hoc, we computed pairwise contrasts with the Tukey HSD (Honest Significant Differences), accounting for multiple companions with alpha correction~\cite{Lenth2020-xk}.
 
 Next, we hypothesized that subjective ratings of control over the tap movement are comparable between INTENTION and AUGMENTED and lower in the EXTERNAL condition.
-Again, we fit a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
+Again, we fitted a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
 Coefficients were assessed in the same way as for the intentional binding parameter above.
 
-In short, we tested two main hypotheses with regard to experiencing agency in our three experimental conditions: (1) participants underestimate the tone delay when acting intentionally.
+In short, we tested two main hypotheses with regard to the agency experience in our three experimental conditions: (1) participants underestimate the tone delay when acting intentionally.
 Adding EMS in line with participants' intention to move does not affect this underestimation.

### 2023-12-13 input/6_results.tex
revision

@@
 
-The grid-search over channels resulted in the BCI leveraging on average 11.5 (SD = 3.7) channels.
-\textcolor{green}{Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9, AF3, F5, F7, F2, FT8, and AF7.} The classifier cross-validation resulted in a mean F1 score of .7 (SD = .03), see figure~\ref{fig:EEG_results} c.
+The grid-search over channels resulted in the BCI leveraging on average 11.5 (SD = 3.7) channels.Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9, AF3, F5, F7, F2, FT8, and AF7.
+The classifier cross-validation resulted in a mean F1 score of .7 (SD = .03), see figure~\ref{fig:EEG_results} c.
 We set the detection threshold to 15 \% false positive rate and at that rate, observed a mean threshold of 57 \% (SD = .02).
 Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57 \% probability.
-% TODO add percentage of trials the prototype was activated before touching the screen (SD = ?)
 
@@
 Furthermore, participants reported that they found the timing of the stimulation to be unpredictable, with one participant noting, ``Stimulation was random in time.''.
-Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly involuntary, but partly as if it was following my decision / choice.''.
-One participant briefly considered the involvement of their brain waves in controlling the stimulation, but expressing doubts about this possibility, stating, ``But I don't think that is the case.''.
+Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly involuntary, but partly as if it was following my decision/choice.''.
+One participant briefly considered the involvement of their brain waves in controlling the stimulation, but expressed doubts about this possibility, stating, ``But I don't think that is the case.''.
 

### 2023-12-13 input/7_discussion.tex
revision

@@
 \section{Discussion}
-With this paper, we contributed a new way to design augmented experiences with the potential to maintain a user's SoA: a bleeding edge BCI to control EMS at moments of users' intention to interact.
+With this paper, we contributed a new way to design the experience of action augmentation augmented experiences with the potential to maintain a user's SoA: a BCI to control EMS at moments of users' intention to interact.
 By leveraging an average of 11.5 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of .7.
@@
 
-% Sandwich: 
-% Good -> [short] cool system, first of its kind setting out to augment user's action preceeding their movement in time
 Our system was built using off-the-shelf, affordable, equipment to physically augment users' actions.
-Taken together, all the tech to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 [anonymized].
+Taken together, all technical devices to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 Euros.
 While we used a 64-channel research-grade EEG system, the channel selection procedure resulted in the system ultimately using a low-density coverage of about 12 channels on average for classification.
@@
 Still, the system achieved what is considered to be a `good' F1 score of .7, and hence sometimes detected users' intention to interact.
-In our user study, this system performance affected users' performance in the time estimation task when comparing the three experimental conditions.
-We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency with respect to intention-based augmentation.
-We believe this to be a very promising finding considering the fact that our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+In our user study, this system performance affected users' tendency for temporal binding when comparing the three experimental conditions.
+We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency for our intention-based augmentation.
+We believe this to be a very promising finding because our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+
+Our system ran at a 10 Hz update rate.
+This was chosen in order to avoid buffer overflow due to the fact that the end-to-end computing latency from measuring the physiological signal to outputting the predicted class and probability took about 40 ms.
+The RP is reported to occur several hundred milliseconds prior to the physical movement~\cite{Schurger2021-vp}.
+Therefore, our augmentation should have resulted in a motion preemption, considering subtracting the computing latency and a delay that the EMS stimulation takes to affect the muscle from the RP-based intention detection.
+However, it is difficult to assess whether participants' movements were in fact pre-empted as no label exists for the `true' intention onset to compare the classifier to.
+That participants were able to act on their own volition, and hence no labelling of their movement intention could be derived, is the key distinction to related works using controlled stimulus-response paradigms.
+
+However, we learned from these paradigms that increasing participants' reaction times by about 80 ms is optimal with regards to maintaining SoA~\cite{Kasahara2019-sk}.
+On one hand, our system is capable of delivering pre-emption within this `SoA-optimal' time range.
+On the other hand, however, the performance of the sub-optimal performance of the classifier means a lot of variation due to false positive detections.
+While in some cases, as we observed in the users' comments, the stimulation may have pre-empted a user's motion, false positive stimulation may have had a significant impact on the overall impression of the system, see section~\ref{improving_classifier} below.
 
 In line with the literature, we found that participants estimated the delay between tap and tone to be lower in both conditions where they were instructed to act on their own volition as compared to the passive condition~\cite{Moore2012-dk}.
-More recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
+We point out that recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
 However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the system.
@@
 
-% \subsubsection{Opportunities}
-\begin{comment}
-% Kasahara work, Tajima, neuromodulation? reference my affordance position paper -> copy text from there!! or during painting -> scott/gehrke -> reuse image?, affordance ++ as well, 
-- relevance of research for prostheses ( see https://link.springer.com/chapter/10.1007/978-3-030-38740-2_8)
-% At this point in time, the movement plan is already in place and so <[rework] there is slack in the perceptual system that can be exploited so user's action can be altered with them noticing>. 
-- EMS augmented interaction might make sense for bigger movements (e,g.
-back muscle to support lifting) -> might have less effect (negative) effect on the feeling of control, since it is a less specific movement
-- However, brain signal might not be the best indicator in such cases (simply motion tracking indicating the start of the movement)
-
-% playful experiences: EMS
-
-\missingfigure[]{Hand weg zieh spiel foto, beispielhafte playful experience in der es interessant ist mit dem agency konzept herumzuspielen und es spielerisch zu entdecken?}
-
-- a bit on HCINT?
-% What do I want to say here?
-Just in general describe the level of integrations that exist and then say where we would place our prototype?
-Or is this more interesting for a later discussion part?
-Maybe yes, just move this to a later stage, it does not really add here and better fits in the discussion, also then to discuss the results and what it means to be integrated in a certain way
-
-%- not only physical integration but also other adpative interfaces, or control of mobile phone (see Cornelio2022 review); % say here somewhere what action augmentation is and how it differs from body and environment/outcome augmentation
-
-% briefly summarize the papers on discriminating action augementation and other such things, reference cornelio 2022 paper
-% Previous research on augmented technology and ideas to maintain sense of of agency
-% The experience of being integrated with a technology, or of being one ``human-technology assemblage'' <REF where is this from, Danry2022 used it>
-
-% Action augmentation technology using motor actuation aims to augment the user’s actions diminishing the SoA (see Cornelio 2022). 
-% - add research
-% Here I want to say where our system lies within the design space of augmentation devices and <cornelio2022> gives a good nomenclature for this.
-Then transition over to how agency is important to consider when localizing our prototpye and then in the next paragraph talk about what agency from a neuroscientific perspective
-\end{comment}
-
 % Conclusion
@@
 
-With AI technologies becoming more and more ubiquitous in our everyday life, questions about how agency is shared between us humans and these AI technologies arise on a regular basis.
+With AI technologies becoming more and more ubiquitous in our everyday lives, questions about how agency is shared between us humans and AI technologies arise regularly.
 With the likely future that these systems move directly onto our bodies, alignment with users' intentions will be the \textit{key} component driving their adoption.

### 2024-03-13 app/fastReach/ptext1.py
figure and revised CHI version

@@ -1,930 +1,0 @@
-# ptext module: place this in your import directory.
-
-# ptext.draw(text, pos=None, **options)
-
-# Please see README.md for explanation of options.
-# https://github.com/cosmologicon/pygame-text
-
-from __future__ import division, print_function
-
-from math import ceil, sin, cos, radians, exp
-from collections import namedtuple
-import pygame
-
-# Global default values
-DEFAULT_FONT_SIZE = 24
-REFERENCE_FONT_SIZE = 100
-DEFAULT_LINE_HEIGHT = 1.0
-DEFAULT_PARAGRAPH_SPACE = 0.0
-DEFAULT_FONT_NAME = None
-DEFAULT_SYSFONT_NAME = None
-FONT_NAME_TEMPLATE = "%s"
-DEFAULT_COLOR = "white"
-DEFAULT_BACKGROUND = None
-DEFAULT_SHADE = 0
-DEFAULT_OUTLINE_WIDTH = None
-DEFAULT_OUTLINE_COLOR = "black"
-OUTLINE_UNIT = 1 / 24
-DEFAULT_SHADOW_OFFSET = None
-DEFAULT_SHADOW_COLOR = "black"
-SHADOW_UNIT = 1 / 18
-DEFAULT_ALIGN = "left"  # left, center, or right
-DEFAULT_ANCHOR = 0, 0  # 0, 0 = top left ;  1, 1 = bottom right
-DEFAULT_STRIP = True
-ALPHA_RESOLUTION = 16
-ANGLE_RESOLUTION_DEGREES = 3
-DEFAULT_UNDERLINE_TAG = None
-DEFAULT_BOLD_TAG = None
-DEFAULT_ITALIC_TAG = None
-DEFAULT_COLOR_TAG = {}
-
-AUTO_CLEAN = True
-MEMORY_LIMIT_MB = 64
-MEMORY_REDUCTION_FACTOR = 0.5
-
-pygame.font.init()
-
-# Options objects encapsulate the keyword arguments to functions that take a lot of optional keyword
-# arguments.
-
-# Options object base class. Subclass for Options objects specific to different functions.
-# Specify valid fields in the _fields list. All keyword fields are optional. Unspecified fields
-# default to None, unless otherwise specified in the _defaults list.
-class _Options(object):
-	_fields = ()
-	_defaults = {}
-	def __init__(self, **kwargs):
-		fields = self._allfields()
-		badfields = set(kwargs) - fields
-		if badfields:
[…]

### 2024-03-13 signal_processing/pi_extract_bci_data_baseline.m

@@ -27,7 +27,7 @@ for pID = pIDs
     %% filter EMG
     
     EMG = pop_eegfiltnew(EEG, 20, 100); % only filter EMG channel
-    EEG = pop_eegfiltnew(EEG, .1, 15); % only filter EMG channel
+    EEG = pop_eegfiltnew(EEG, .1, 15); % EEG
     
     EEG.data(end,:) = EMG.data(end,:).^2;
     

### 2024-04-03 input/7_discussion.tex
UIST submission

@@
 \section{Discussion}
-With this paper, we contributed a new way to design the experience of action augmentation augmented experiences with the potential to maintain a user's SoA: a BCI to control EMS at moments of users' intention to interact.
-By leveraging an average of 11.5 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of .7.
-We demonstrated in eight participants that one effect of SoA, intentional binding, is maintained when using our system.
+With this paper, we contributed a new way to design the experience of action augmentation with the potential to maintain a user's SoA: A BCI to control EMS at moments of users' intention to interact.
+By leveraging an average of 11 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of about .7.
+We demonstrated in ten participants that one effect of SoA, intentional binding, is maintained when using our system.
 However, the subjectively rated level of control was lower when using the system as compared to voluntary interaction without EMS.
@@
 Taken together, all technical devices to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 Euros.
-While we used a 64-channel research-grade EEG system, the channel selection procedure resulted in the system ultimately using a low-density coverage of about 12 channels on average for classification.
+While we used a 64-channel research-grade EEG system, the channel selection procedure resulted in the system ultimately using a low-density channel coverage for classification.
 Today, many low-density EEG devices are available on consumer markets at affordable price points, see~\citet{Niso2023-ce} for a recent summary of available wireless systems.
@@
 In our user study, this system performance affected users' tendency for temporal binding when comparing the three experimental conditions.
-We conclude that taking users' sentiments and ratings of control into account, this indicates a general modulation of agency for our intention-based augmentation.
+
+We first conclude that augmentation generally impacted SoA as was evident in the users' sentiments and ratings of control.
+While we believe that our prototype did not perform at a consistently high enough performance to allow for more fine-grained inferences about the the pre-emptive gain it achieved, we maintain that in some cases it elicited an pre-emptiot that maintained agency to a certain degree.
 We believe this to be a very promising finding because our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
@@
 On one hand, our system is capable of delivering pre-emption within this `SoA-optimal' time range.
-On the other hand, however, the performance of the sub-optimal performance of the classifier means a lot of variation due to false positive detections.
+On the other hand, however, the sub-optimal performance of the classifier resulted in a lot of variation due to false positive detections.
 While in some cases, as we observed in the users' comments, the stimulation may have pre-empted a user's motion, false positive stimulation may have had a significant impact on the overall impression of the system, see section~\ref{improving_classifier} below.
@@
 On the other hand, false positive classification meant that an EMS pulse was sent in the absence of a true intention by the participant.
-Both cases, individually and jointly, had the potential to significantly impact the user experience and erode trust in the system.
+Both cases, individually and jointly, had the potential to impact the user experience and erode trust in the system.
 In the interviews, participants reported only a relatively low number of correctly detected intentions.
 This may have ultimately led to a misalignment between users' perceptions and the observed, measurable, system behavior.
+Furthermore, the necessity for a fixed block/condition order may have further contributed to this effect.
+However, keeping the order was necessary in order to first obtain training data for the BCI based on the unique EEG signals of each participant.
 
 As a consequence, the insights gained into the system's potential to preserve a sense of agency are limited.
-The quantitative metrics employed in our study, specifically those measuring the sense of control (including temporal binding and item assessment), may not have captured the nuanced temporal alignment experiences associated with EMS augmentation and user movement intentions but rather a more holistic assessment of the entire experimental block, which included trials with missed stimulation, too early stimulation, and some trials where the stimulation was in alignment with the participants' intent.
+The quantitative metrics employed in our study, specifically those measuring the sense of control (including temporal binding and item assessment), may not have captured the nuanced temporal alignment experiences associated with EMS augmentation and user movement intentions but rather a more holistic assessment of the entire experimental block, which included trials with missed stimulations, too early stimulations, and some trials where the stimulation was in alignment with the participants' intent.
 

### 2024-04-03 input/2_introduction.tex

@@
 One key change in perspective is, that \textit{integrated} users share agency with the computing machinery to execute tasks.
-This is a critical distinction as \textit{integration} technologies are designed to directly influence people’s bodies, their actions, and the resulting action outcomes.
+This is a critical distinction as \textit{integration} technologies are designed to directly influence people’s bodies, their actions, and the resulting action outcomes~\cite{Cornelio2022-aq}.
 
@@
 Unfortunately, such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/teaser.png}%{figures/bci_game.png}
+    \caption{We propose an augmentation system that maintains agency.
+When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals.
+A brain-computer interface (BCI) then predicts the user to be in either an idle (red) or pre-movement (green) state.
+In the latter case, electrical muscle stimulation (EMS) is triggered and the user's hand is moved.
+Image taken with consent from participant.}
+\end{figure}
 
@@
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\columnwidth]{figures/teaser_new.png}%{figures/bci_game.png}
-    \caption{Our augmentation system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals.
+% \begin{figure}[!h]
+%     \centering
+%     \includegraphics[width=\columnwidth]{figures/teaser_new.png}%{figures/bci_game.png}
+%     \caption{Our augmentation system: When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals.
 A brain-computer interface (BCI) predicts the data to be in either of two classes: Idle or pre-movement.
@@
 Image taken with consent from participant.}
-\end{figure}
+% \end{figure}
 
@@
 
-Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially preceding their voluntary motor command.
+Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially even preceding their voluntary motor command.
 This augmentation moved the ring finger in accordance with the user's intention to act.
 We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle.
-In our user study, a mixed-methods research approach was employed to explore whether aligning the physical impact on the user's body with their intention to move preserved their SoA.
+To get a first glimpse of how our system can maintain agency in augmented users, we conducted a user study using a mixed-methods research approach.
+In short, we explored whether aligning the physical impact on the user's body with their intention to move preserved their SoA.

### 2024-04-03 UIST_submission/source/input/0_abstract.tex
UIST submission

% To maintain a sense of agency (SoA), physical motor augmentation must align with the users' expectations. In experimental setups, this is often achieved using stimulus-response paradigms where the user behavior is predictable. Moving towards user augmentation in the everyday world where user behavior is less predictable, we developed a brain-computer interface (BCI) based augmentation using electrical muscle stimulation (EMS). By classifying readiness potentials from the user's electroencephalogram (EEG), our system controlled the user's movements at the time of their intent to interact. The system was able to discriminate pre-movement from idle EEG segments with an F1 score of 0.7. In a pilot study, we observed intentional binding, a compression of time between action and outcome, and a higher level of control when participants worked with the system instead of being passively moved. This is a step towards predictive interfaces that leverage the users’ bodies yet still feel natural as they align with the users’ intentions.

To maintain a sense of agency (SoA), physical motor augmentation must align with the users' intentions. In experiments, this is often achieved using stimulus-response paradigms. Here, the user behavior follows the stimulus and motor augmentation can be delivered in line with the user's motor plan. But, in the everyday world user's act on their own volition. Here, we designed a system to control the user's movements at the time of their intent to interact. A brain-computer interface (BCI) classified readiness potentials from the user's electroencephalogram (EEG). The BCI's output then gated a motor augmentation via electrical muscle stimulation (EMS). The system discriminated pre-movement from idle EEG segments with an F1 score of 0.7. In a pilot study, we found evidence for a maintained SoA. Participants mentally compressed the time between action and outcome, intentional binding, and reported a higher level of control when working with the system instead of being passively moved. This is a step towards leveraging the users’ bodies while maintaining SoA.

### 2024-04-03 UIST_submission/source/input/3_related_work.tex
UIST submission

--- 2023-12-13 input/3_related_work.tex
@@
 Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle-activating nerves.
-In these scenarios, the user computes no predictions about the movement and its outcome.
+Irrespective of the method applied, the user computes no predictions about the movement and its outcome in these scenarios.
 Thus, concerning the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
-With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency.
+With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- negatively impacts the experience of agency.
 
@@
 Hence, \textit{pre-empting} the user's motion can be designed to fall in line with their intention to move, thereby maintaining SoA.
-Previous work on such an augmentation scenario has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.
+Previously, ~\citet{Kasahara2019-sk} used a reaction time task in which participants had to tap a target on screen as soon as it appeared and subsequently rate their SoA.
+They showed that in such a scenario, user's actions can be pre-empted and that a pre-emption of about 80 ms best preserves agency~\cite{Kasahara2019-sk, Kasahara2021-gy}.
 
-Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
-Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.
-However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment.
+This is in line with evidence from cognitive neuroscience which indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
+Hence, after this ``point of no return'' user's struggle to assign a source other than themselves to the action initiation.
+Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention to act.
+Cruciall then, a key challenge remaining is to design systems that maintain agency when user's actions are unpredictable and where the experimenter does \textit{not} have executive control over the environment.
 In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?
@@
 Of the possible physiological signals that can be leveraged, the EEG is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain.
+
 The RP, or \textit{lateralized readiness potential}, is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}.
-The RP is reliably observed at electrodes placed over the sensorimotor cortex contralateral to the acting hand.In the 10-20 system for EEG electrode placement~\cite{Jasper1983-uw}, these are electrode C3 located over the sensorimotor cortex of the left hemisphere, and C4 vice versa.
+The RP is reliably observed at electrodes placed over the sensorimotor cortex contralateral to the acting hand.
+In the extended 10-20 system for EEG electrode placement~\cite{Jasper1983-uw}, these are electrodes C3 located over the sensorimotor cortex of the left hemisphere, and C4 vice versa.
 However, activity observed at electrode Cz is reported most frequently as it reflects neural activity originating from the sensorimotor cortex without lateralization bias.
@@
 For example, the RP can be detected in real-time using a brain-computer interface (BCI).~\citet{Schultze-Kraft2016-bx} demonstrated a prototype that detects RPs in participants ongoing EEG data and adapts an interface accordingly.
-Participants were instructed to veto their self-initiated movement whenever a red dot occurred on the screen.
+In their study, participants were instructed to veto their self-initiated movement whenever a red dot occurred on the screen.
 The red dot's appearance was controlled by the BCI.
@@
 After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot's appearance on screen.
+
+Taken together, these findings demonstrated that the RP is a reliable signal preceding voluntary movement initiation and hence, is a well-suited candidate to base (neuro--)adaptive systems on.

### 2024-04-03 UIST_submission/source/input/6_results.tex
UIST submission

\section{Results}
We first report the performance of the classification system, as the system performance greatly impacts the subjective experience of the motor augmentation.

\subsection{Classifier Performance}
Visual inspection of the amplitudes at electrode Cz revealed an increase in the difference between \textit{pre-movement} and \textit{idle} data segments towards the onset of the finger movement, see figure~\ref{fig:EEG_results}a. The slope feature for the exemplary channel Cz discriminates well between the two classes (${t_{(10)}} = 4.4, p < .001$), see figure ~\ref{fig:EEG_results}b bottom. The scalp maps in figure~\ref{fig:EEG_results}b top show the (color-coded) mean slope for each channel and each class. Central channels on the contralateral side to the moving finger on the right hand show a negative slope for the pre-movement class and a neutral slope for the idle class. Furthermore, differences in slope were also observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifested only for the idle class.

The grid-search over channels resulted in the BCI leveraging on average 11.2 (SD = 3.6) channels. Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9 and AF3. The classifier cross-validation resulted in a mean F1 score of .71 (SD = .03), see figure~\ref{fig:EEG_results}c. We set the detection threshold to 15 \% false positive rate and at that rate, observed a mean threshold of 57 \% (SD = .04). Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57 \% probability.

\subsection{Behavior: Intentional Binding}
In the intentional binding task, i.e. the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (INTENTION M = 187.5, SD = 139; EXTERNAL M = 208.6, SD = 140.1, AUGMENTED M = 196.0, SD = 138.2). The condition influenced the estimation (${\chi^{2}_{(2)}} = 17.1, p < .001$), see figure~\ref{fig:results}a. Post-hoc tests revealed that the underestimation was less pronounced in EXTERNAL as compared to INTENTION ($beta = 22.6, p < 0.001$). This was also the case in EXTERNAL, with less pronounced underestimations of the time interval between tap and tone compared to AUGMENTED ($beta = 15.3, p = 0.018$). No difference was observed between INTENTION and AUGMENTED.

\subsection{Subjective Ratings \& Reports}
The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 43.7, p < .001$), see figure~\ref{fig:results}b. Post-hoc tests revealed that participants rated their motion control higher in INTENTION compared to EXTERNAL ($beta = 4.8, p < .0001$), and higher in INTENTION compared to AUGMENTED ($beta = 3.1, p < .0001$). Further, higher control was observed in AUGMENTED compared to EXTERNAL ($beta = -2.9, p = .006$).

\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{figures/subjective_results.pdf}
    \caption{(a) Temporal binding effect over conditions. With an average presented time interval of 350ms, lower values in the estimated time interval indicate an underestimation, i.e., temporal binding. (b) Subjective ratings of control across conditions. Significance labels obtained from post-hoc tests on estimated marginal means.}
    \label{fig:results}
\end{figure}

\paragraph{Functionality.} Eight out of the ten participants made a reference to the source of the EMS in the AUGMENTED condition. In that subgroup, most participants wondered how the system worked, one remarked, ``I can't explain how it works technically.''. Another one mentioned that ``maybe the time estimation task had something to do with it.''. They found the stimulation to be sporadic, even during their own actions, leading to a belief that it was coincidental. As one participant put it, ``I have no idea what controlled the stimulation, I think it was coincidence when stimulation occurred.''. Furthermore, participants reported that they found the timing of the stimulation to be unpredictable, with one participant noting, ``Stimulation was random in time.''.

Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly involuntary, but partly as if it was following my decision/choice.''. One participant noted that they felt that ``somehow the information is in my arm''. Another one briefly considered the involvement of their brain waves in controlling the stimulation, but expressed doubts about this possibility, stating, ``But I don't think that is the case.''.

\paragraph{Guessed Percentage.} 
Some participants reported that the stimulation ``sometimes overlapped with the movement, but not often,'' or, it ``came only rarely.'' Another one noted, ``In a few cases, the stimulation came when I had already started the movement.'' For several participants, this overlap between their actions and the system's response occurred infrequently, with another user noting, ``Once, in the millisecond range between my planned movement and its execution.'' However, for some participants, there were moments of near-perfect synchronization, as described by one participant, ``In 3 cases it happened that my intention to press and the impulse of the device happened simultaneously.'' or ``in 3--4 cases it happened simultaneously that i wanted to tap and the stimulation happened''. Another one noted, ``Sometimes it really happened that they overlapped. So that I just started to move, and then the device activated,'' that user estimated an overlap of ``40 \%'' while two other users said [we worked together in] ``15 \% of trials.'' These accounts collectively highlight the varied and occasionally synchronous nature of the system's timing in relation to the users' intentions.

\paragraph{Correct Interaction.} 
In terms of valence sentiments for the cases where the stimulation aligned with participants' intentions, some participants indicated that the experience was positive, for example, participants described the experience as ``rather funny'', ``weird but funny'', ``pleasant'' or ``helped me with the execution'' and ``it was more of a collective movement.'' One participant remarked ``then it was ok to experience the stimulation, but also not more than ok'' while another one noted ``It had a bit of thinking ahead to it''. Some noted that they experienced an increase in their physical strength, remarking, ``supported my strength'', ``made me type more firmly'' and ``my typing performance was increased.'' On the other hand, some participants had negative sentiments during these moments of aligned stimulation, remarking, ``felt like I was still in competition with the system'', ``it did not feel like an acting together'' and ``on a psychological level it was a loss of control.'' Another one noted, that they ``felt determined-by-others and then tried to resist the impulse. I felt excluded from the decision to tap''.

%%% Resources and Ideas
% and the agency condition by xx (CI: ). Subsequent non-parametric comparisons revealed the significance of differences only between  
% xx (p = ).
% x\

%\missingfigure{A: Boxplot of questionnaire scores of three Blocks (block 2 grey, because not intention given) - play around with visualization of words that were used to explain control rating. B. Boxplots intentional binding (maybe integrate A and B) C. more subjective evaluation of interview / maybe world cloud / maybe sentiment analysis) }

% Did they get faster? Calculate time between hand motion onset and button press. If yes, evidence that there was acceptance for the help/augmentation the system provided

### 2024-04-03 signal_processing/pi_extract_bci_data_baseline.m

@@ -1,7 +1,7 @@
 
 
 % pID = 18;
-pIDs = [12, 14:21] ; %1:10;
+pIDs = [12, 14:23] ; %1:10;
 
 % %% config
 current_sys = "mac";

### 2024-04-03 signal_processing/pi_extract_bci_data_processed.m

@@ -1,7 +1,7 @@
 
 
 % pID = 18;
-pIDs = [12, 14:17, 19:21] ; %1:10;
+pIDs = [12, 14:17, 19:23] ; %1:10;
 
 % %% config
 current_sys = "mac";

### 2024-04-03 signal_processing/pi_extract_results_design.m

@@ -7,7 +7,7 @@ pi_bemobil_config;
 
 %% load data and parse events
 
-pIDs = [12, 14:21] ; %1:10;
+pIDs = [12, 14:23] ; %1:10;
 conds = {'Baseline', 'passive', 'agency1'}; %
 design = [];
 

### 2024-04-03 signal_processing/pi_extract_rp_for_plot.m

@@ -7,7 +7,7 @@ pi_bemobil_config;
 
 %% load data and parse events
 
-pIDs = [12, 14:17, 19:21] ; %1:10; % remove also 18 1
+pIDs = [12, 14:17, 19:23] ; %1:10; % remove also 18 1
 cond = {'Baseline'};
 cz_ix = 24;
 design_erp = [];

### 2024-04-03 signal_processing/pi_import.m

@@ -13,10 +13,10 @@ function pi_import(cfg, subject, training)
     
     % optional for dataset_description.json
     generalInfo.dataset_description.License             = 'CC BY';
-    generalInfo.dataset_description.Authors             = {'Lukas Gehrke', 'Leonie Terfurth'};
+    generalInfo.dataset_description.Authors             = 'n/a';
     generalInfo.dataset_description.Acknowledgements    = 'n/a';
-    generalInfo.dataset_description.Funding             = {''};
-    generalInfo.dataset_description.ReferencesAndLinks  = {};
+    generalInfo.dataset_description.Funding             = 'n/a';
+    generalInfo.dataset_description.ReferencesAndLinks  = 'n/a';
     generalInfo.dataset_description.DatasetDOI          = 'n/a';
     
     % general information shared across modality specific json files 

### 2024-04-03 signal_processing/pi_pipeline.m

@@ -14,7 +14,8 @@ pi_bemobil_config;
 
 % set to 1 if all files should be recomputed and overwritten
 force_recompute = 1;
-subjects = [12, 14:17, 19:21] ;
+% subjects = [12, 14:17, 19:21] ;
+subjects = [22:23];
 
 for subject = subjects
     pi_import(bemobil_config, subject, 0);

### 2024-04-03 signal_processing/pi_plot_topomaps.m

@@ -6,7 +6,7 @@ pi_bemobil_config;
 load('/Volumes/Lukas_Gehrke/fastReach/data/eeglab2python/locs.mat');
 
 % load csv scores and concatenate to one matrix
-pIDs = [12, 14:17, 19:21] ; %1:10; % remove also 18 1
+pIDs = [12, 14:17, 19:23] ; %1:10; % remove also 18 1
 
 % what is all_chans_score_order and what does it mean???
 for pID = pIDs

### 2024-09-25 submission_UIST/source/input/5_user_study_and_methods.tex
clean up

--- 2023-12-13 input/5_user_study_and_methods.tex
@@
 \section{User study \& Methods}
-With this study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when the user's own brain signals serve as the control signal to a physical end effector.
-Therefore, we conducted a user study comparing three conditions of agency experience.
+In this paper, we present a prototype that uses a user's own brain signals as the control signal to a physical end effector.
+With a user study, we wanted to find out whether the experience of agency can be preserved during physical action augmentation when using our prototype.
+
+In the user study, we compared three conditions of agency experience.
 With the first two conditions, INTENTION and EXTERNAL we queried users at the two edges of agency experience.
-In EXTERNAL participants had no intention to move and no control over their movement.
+In EXTERNAL, participants had no intention to move and no control over their movement.
 On the other hand, in INTENTION, participants were acting as they would in their day-to-day lives, fully in control.
 We then investigated in a third condition where our AUGMENTED prototype sits on the spectrum between INTENTION and EXTERNAL, in other words, whether it preserved agency and to what degree.
-To answer our question, we employed a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) interview.
+To answer our question, we employed a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) exit interview.
+
+To assist readers in replicating our prototype and experiment, we provide the necessary technical details, the complete source code for the BCI and the experiment, the collected data, as well as all the analysis scripts\footnote{[anonymized]}
 
 \subsection{Participants}
-Nine participants (M = 30.00 years, SD = 3.81) were recruited from our local institution and through the institute's online participant pool.
-All participants were right-handed.
+Eleven participants (M = 29.9 years, SD = 4) were recruited from our local institution and through the institute's online participant pool.
 Participants were compensated with course credit or 12 Euro per hour of study participation.
@@
 Precisely, they did not initiate their movements at their own volition but rather immediately at the onset of each trial, thereby violating the task instruction of waiting 2-3 s after trial onset, see task description below for more detail.
+% All participants were right-handed
 
 \subsection{Apparatus}
-The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel EMG device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected to two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
-To assist readers in replicating our experiment, we provide the necessary technical details, the complete source code for the experiment, the collected data, and the analysis scripts\footnote{[anonymized]}.
-% TODO add links after anonymization is lifted.
+The experimental setup, depicted in Figure~\ref{fig:setup}, comprised: (1) a 1-channel Electromyography (EMG) device, (2) a 64-channel EEG system, (3) a medically-compliant EMS device connected to two electrodes worn on the forearm, and (4) a tablet to run the experiment and collect behavioral responses.
 
@@
 One electrode was placed under the right eye to provide additional information about eye movements (vEOG).
-After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10k$\Omega$.
+After fitting the cap, all electrodes were filled with conductive gel to ensure proper conductivity, and electrode impedance was brought below 10k$\Omega$ where possible.
 EEG (and EMG) data were recorded with a sampling rate of 250 Hz.
@@
 \subsection{Experimental Task, Design and Procedure}
-Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED.
+Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED
 
-\begin{figure}
+\begin{figure}[h]
     \centering
@@
 \end{figure}
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figures/method.png}
+    \caption{Diagram summarizing: (a) signal processing and event labeling from muscle activity, (b) extraction of linear slope features from labeled brain signals and (c) real-time application of the BCI gating the muscle stimulation hardware.}
+    \label{fig:method}
+\end{figure*}
 
@@
 
-\begin{figure*}[!h]
-    \centering
-    \includegraphics[width=\textwidth]{figures/method.png}
-    \caption{Diagram summarizing: (a) signal processing and event labeling from muscle activity, (b) extraction of linear slope features from labeled brain signals and (c) real-time application of the BCI gating the muscle stimulation hardware.}
-    \label{fig:method}
-\end{figure*}
-
 \subsubsection{Training of EEG classifier}
@@
 Ultimately, to determine the threshold at which, during the real-time application, the classifier would trigger the EMS, we computed the Receiver-operator characteristic (ROC) and from this selected the threshold at 15 \% false positive rate.
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figures/eeg_results_new.pdf}
+    \caption{(a) Grand average event-related potential (ERP) at electrode Cz of pre-movement EEG data epochs preceding the last second before a movement, idle epochs of the same trial are plotted alongside but originate from a different time window, see section~\ref{eeg_methods}; (b) Bottom: Slope features for both idle and pre-movement classes at electrode Cz.
+Top: Scalp maps of slope values for both classes and all channels.
+(c): Mean and per participant ROC curves.
+Dotted line at 15 \% false positive rate indicate the selected threshold for the real-time application.}
+    \label{fig:EEG_results}
+\end{figure*}
 
@@
 Prior to any statistical analyses of the intentional binding measure, outlier trials were rejected.
-We applied `extreme outlier removal' using Tukey's method~\cite{Tukey1949-sl}.
-Three time intervals were considered relevant to describe `regular' behavior across trials which would manifest by: (1) Tapping the screen in a reasonable interval after the fixation cross disappeared.
+We applied `extreme outlier removal' using Tukey's method~\cite{Tukey1949-sl} on three time intervals that well describe `regular' behavior across trials: (1) Tapping the screen in a reasonable interval after the fixation cross disappeared.
 An excessively short or long delay indicated that participants either tapped the screen prematurely by accident or they were checking in with the experimenter, respectively.
@@
 A long delay between the EMS trigger and the subsequent screen tap indicated that the stimulation was not strong enough in this trial to lead to muscle actuation resulting in a screen tap.
-Taken together, applying Tukey's method to each of these time windows and fusing the rejected trials led to the exclusion of 102 trials across all participants (M = 34.00, SD = 16.46) and a final data set of 1698 trials.
+Taken together, applying Tukey's method to each of these time windows and fusing the rejected trials led to the exclusion of 110 trials across all participants (M = 12.2, SD = 9.1).
+% and a final data set of 2140 trials.
 
@@
 Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
-To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect and as a random effect, we considered the intercepts for participants ($time interval ~ condition + 1|participantID$).
-We obtained p-values of the coefficients by calculating likelihood-ratio tests.
+
+To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect and `participandID' as a random effect.
+The model was specified as `time interval $\sim$ condition + (1|participantID)' and fit using the `lme4' package~\cite{Bates2015-bh}.
+We included all single trial data points in the analyses and did not aggregate before fitting the model as fitting the random effect `participantID' corrects for the correlation of the repeated measurements within participants.
+We obtained a test statistic by calculating likelihood-ratio tests comparing the full model as specified above against the null model `time interval $\sim$ 1 + (1|participandID)'.
 All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
-Post hoc, we computed pairwise contrasts with the Tukey HSD (Honest Significant Differences), accounting for multiple companions with alpha correction~\cite{Lenth2020-xk}.
+We computed post-hoc pair-wise tests for `condition' corrected for multiple comparisons (Tukey method) using the emmeans package~\cite{Lenth2020-xk}.
 

### 2024-09-25 input/7_discussion.tex
Updates from Overleaf

@@
 \section{Discussion}
-With this paper, we contributed a new way to design the experience of action augmentation with the potential to maintain a user's SoA: A BCI to control EMS at moments of users' intention to interact.
+In this paper, we investigated if SoA can be maintained in action augmentation when the augmentation aligns with the users intention.
+We evaluated a BCI that controls EMS at moments of users' intention to interact.
 By leveraging an average of 11 EEG channels and using a simple, fast-to-compute feature, the BCI achieved a mean F1 score of about .7.
-We demonstrated in ten participants that one effect of SoA, intentional binding, is maintained when using our system.
-However, the subjectively rated level of control was lower when using the system as compared to voluntary interaction without EMS.
-This was differentiated by participants' qualitative reports: when the stimulation aligned with participants' intentions, positive sentiments outweighed negative ones.
+
+%% I Agency measures:
+
+% 1. IB Results
+In the user study with 10 participants we found no evidence for a disruption in intentional binding, hinting at a maintained SoA.
+In line with the literature, participants underestimated the delay between tap and tone in both conditions where they were instructed to act on their own volition~\cite{Moore2012-dk}.
+However, we also found that participants similarly underestimated the delay in INVOLUNTARY.
+This conflicting finding is in line with recent literature that has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
+This might be especially true for cases where one's own body is completing the action while being externally controlled.
+Interestingly, in our scenario, proprioceptive signals and other indicators of embodied actions remain in line with the user acting at their own volition.
+Hence, the underestimations we found might purely follow from multisensory causal binding.
+The temporal compression effect might therefore not be based on inferring subjective causality following from intention but solely from causally binding sensory information.
+
+We do note that the underestimation appeared to be trending smaller in INVOLUNTARY as in both, INTENTION and AUGMENTED.
+However, the effect might be smaller than what could be proven given our sample size.
+Within the AUGMENTED condition trials, we observed no difference between `EMS-controlled' --and `muscle-controlled' trials, indicating an alignment with the augmentation technology in the AUGMENTED's `EMS-controlled' trials as the stimulation did not appear to disrupt participants' performance in the temporal delay estimation task.
+
+% 2. Questionnaire & Interviews
+On the level of subjective experience, we found that participants rated their level of control lower when using the system (AUGMENTED) as compared to voluntary interaction without EMS (INTENTION).
+This is a common finding when EMS is used for muscle control.
+For example,~\citet{Lopes2015-dk} reported people frequently not feeling in control in an interaction facilitated via EMS.
+INTENTION than might serve a very high mark for ratings following AUGMENTED and a better fitting baseline comparison would make sense.
+One way to better assess agency attributions when working with the system could be a comparison to earlier works that also leveraged EMS instead of our INTENTION condition where no EMS was in use.
+Triggering EMS based on simple heuristics (or sham) and contrasting this with BCI-controlled EMS stimulation could shed light on agency attributions exclusively explained via the BCI control.
+As a next step forward, stimulus-response paradigms like those of~\citet{Kasahara2019-sk, Kasahara2021-gy} could then serve as the next step-up in evaluation, with the augmentation performance, such as \textit{pre-emptive} gain, coming into focus.
+
+%% Now use subjective reports and comments to go over into system performance and evaluation of false positive using prediction error ERPS
+From participants' qualitative reports we learned that when the stimulation aligned with participants' intentions, positive sentiments outweighed negative ones.
 When it did not, participants reported a negatively perceived loss of control.
+To situate participants comments, a differentiated look at the classification system's performance is warranted.
 
-Our system was built using off-the-shelf, affordable, equipment to physically augment users' actions.
+%% II System performance and prediction error ERPs to obtain labels
+First, we note that our system was built using off-the-shelf, affordable, equipment to physically augment users' actions.
 Taken together, all technical devices to control the users' movements, i.e., the EMS stimulation device, Arduino, and switchboard, cost less than 100 Euros.
@@
 Today, many low-density EEG devices are available on consumer markets at affordable price points, see~\citet{Niso2023-ce} for a recent summary of available wireless systems.
+Still, the system achieved what is considered to be a `good' F1 score of .7, and hence sometimes detected users' intention to interact.
 
-Still, the system achieved what is considered to be a `good' F1 score of .7, and hence sometimes detected users' intention to interact.
-In our user study, this system performance affected users' tendency for temporal binding when comparing the three experimental conditions.
+While an F1 score of .7 may be considered `good' for many classification scenarios, when aiming to elicit a feeling of control this level of performance may likely not be good enough, see~\citet{Papenmeier2022-oi} for a review.
+Balancing false positive rate (FPR) and true positive rate (TPR) appropriately may prove crucial to elicit an experience of agency.
 
-We first conclude that augmentation generally impacted SoA as was evident in the users' sentiments and ratings of control.
-While we believe that our prototype did not perform at a consistently high enough performance to allow for more fine-grained inferences about the the pre-emptive gain it achieved, we maintain that in some cases it elicited an pre-emptiot that maintained agency to a certain degree.
-We believe this to be a very promising finding because our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+\subsection{ERPs for Data Labeling}
+% To provide the full picture we first describe the impact from a signal detection theory point of view: 
+False negative classifications meant that the system did not trigger an EMS pulse in line with participants' intent to tap on the screen.
+On the other hand, false positive classification meant that an EMS pulse was sent in the absence of a true intention by the participant.
+Both cases, individually and jointly, had the potential to impact the user experience and erode trust in the system.
 
-Our system ran at a 10 Hz update rate.
-This was chosen in order to avoid buffer overflow due to the fact that the end-to-end computing latency from measuring the physiological signal to outputting the predicted class and probability took about 40 ms.
-The RP is reported to occur several hundred milliseconds prior to the physical movement~\cite{Schurger2021-vp}.
-Therefore, our augmentation should have resulted in a motion preemption, considering subtracting the computing latency and a delay that the EMS stimulation takes to affect the muscle from the RP-based intention detection.
-However, it is difficult to assess whether participants' movements were in fact pre-empted as no label exists for the `true' intention onset to compare the classifier to.
-That participants were able to act on their own volition, and hence no labelling of their movement intention could be derived, is the key distinction to related works using controlled stimulus-response paradigms.
+We investigated EEG activity at electrode FCz to ascertain presence or absence of a prediction error in response to the stimulation or movement onset, see figure~\ref{fig:erp}.
+Ultimately, classifying the EEG here could be leveraged to approximate labels for classifier validation.
+The idea is that in AUGMENTED a false positive would mean that the EMS is falsely triggered and hence the movement onset would catch the participant by surprise.
+On the other hand a false negative means that a user starts their movement on their own without EMS, potentially equally `surprising' them, or in other words not aligning with their prediction.
+In line with this theory, we found a negativity affected by the trial condition in the  150--250 ms time window after movement onset, see figure~\ref{fig:erp}a.
+A strong negativity was present in INVOLUNTARY, where the user experienced the stimulation at unpredictable times.
+A less pronounced negativity was present in AUGMENTED, with only a marginal peak present in INTENTION.
 
-However, we learned from these paradigms that increasing participants' reaction times by about 80 ms is optimal with regards to maintaining SoA~\cite{Kasahara2019-sk}.
-On one hand, our system is capable of delivering pre-emption within this `SoA-optimal' time range.
-On the other hand, however, the sub-optimal performance of the classifier resulted in a lot of variation due to false positive detections.
-While in some cases, as we observed in the users' comments, the stimulation may have pre-empted a user's motion, false positive stimulation may have had a significant impact on the overall impression of the system, see section~\ref{improving_classifier} below.
+To carve out the differences between the two trial categories `EMS-controlled', and `muscle-controlled' in AUGMENTED from the two \textit{anchoring} conditions, we subtracted each category from INVOLUNTARY and INTENTION accordingly, see figure~\ref{fig:erp}b for an explorative view on the ERP data.
 
-In line with the literature, we found that participants estimated the delay between tap and tone to be lower in both conditions where they were instructed to act on their own volition as compared to the passive condition~\cite{Moore2012-dk}.
-We point out that recent literature has questioned the validity of the intentional binding phenomenon as a correlate of agency, stating that the effect may ``merely represent a strong case of multisensory causal binding.''~\cite{Suzuki2019-pi, Gutzeit2023-ei, Hoerl2020-my}.
-However, in our study, we observed the effect of compressing the action-outcome delay to be maintained even when using the system.
-Hence, the effect was maintained for both conditions of voluntary action.
+First, we note that all `muscle-controlled' trials from AUGMENTED were false negatives, as the classifier failed to pick up on the readiness potential preceding the movement.
+By subtracting AUGMENTED, we observed that the pre-movement activity at electrode FCz did not differ from INTENTION.
+This shows that both these trial groupings exhibited a similar readiness potential at electrode FCz and the classifier failed to pick up on on it.
+Following the movement start, a trend emerged after 150--250 ms, in which `muscle-controlled' trials from AUGMENTED trends towards a stronger negativity than INTENTION trials, hence a positive difference, see figure~\ref{fig:erp}b blue line.
+This may indicate that participants always expected an EMS pulse in AUGMENTED and carrying out the movement without EMS \textit{support} violated their prediction.
 
-% Bad -> (1) situate the findings of the IB, (2) bad classification metrics not good enough to feel in control
-\subsection{Limitations}
+In the `EMS-controlled' trials in AUGMENTED, both false positive and true positive classifications overlap.
+In the contrast with INVOLUNTARY, the readiness potential which lead towards a positive classification outcome was visible in the second preceding the movement, see the negative going deflection in figure~\ref{fig:erp}b orange line.
+In the 150--250 ms time window after movement onset, a similar trend as described above was visible.
+The slight positive bump was due to a stronger negativity in `EMS-controlled' trials in AUGMENTED as compared to INVOLUNTARY.
+One possible explanation is that the trend is driven by false positive classifications, where a slight misalignment of the stimulation, i.e.
+it being too early, severely disrupts the user, eliciting a string prediction error signal.
+Taken together, we believe that contrasting and classifying single-trials could be a fruitful endeavor for approximating classification labels.
+
+\subsection{Limitations \& Future Directions}
 Two main procedural issues arose concerning the time estimation task: First, contrasting AUGMENTED with INTENTION, we noted that when EMS moved participants' fingers, they sometimes reported that their finger was pressing on the touchscreen for a longer duration as compared to their `normal' touchscreen tap.
@@
 
-\subsubsection{Improving the Prediction of the Intent to (Inter-) act}\label{improving_classifier}
-While an F1 score of .7 may be considered `good' for many classification scenarios, when aiming to elicit a feeling of control this level of performance may likely not be good enough, see~\citet{Papenmeier2022-oi} for a review.
-Balancing false positive rate (FPR) and true positive rate (TPR) appropriately may prove crucial to elicit an experience of agency.
-
-For the current study, we must note the following limitations arising from the (potentially) insufficient performance of the classification system.
-To provide a full picture we first describe the impact from a signal detection point of view: False negative classifications meant that the system did not trigger an EMS pulse in line with participants intent to tap on the screen.
-On the other hand, false positive classification meant that an EMS pulse was sent in the absence of a true intention by the participant.
-Both cases, individually and jointly, had the potential to impact the user experience and erode trust in the system.
+\subsubsection{System Performance}
 In the interviews, participants reported only a relatively low number of correctly detected intentions.
@@
 
-% Good -> This can be improved by using Gaze Dynamics, EMG and other things
-% If it would work, how can it be used? Implications -> playful experiences, adaptive interfaces
-Hence, a \textit{key} objective remains in improving the overall performance of the classification system.
+Our system ran at a 10 Hz update rate.
+This was chosen in order to avoid buffer overflow due to the fact that the end-to-end computing latency from measuring the physiological signal to outputting the predicted class and probability took about 40 ms.
+The RP is reported to occur several hundred milliseconds prior to the physical movement~\cite{Schurger2021-vp}.
+Our augmentation then should have resulted in a motion pre-emption, considering a subtraction of the computing latency and a delay that the EMS stimulation takes to affect the muscle from the RP-based intention detection.
+However, it is difficult to assess whether participants' movements were in fact pre-empted as no readily available label exists for the `true' intention onset to compare the classifier to.
+That participants were able to act on their own volition, and hence no direct label of their movement intention could be derived, is the key distinction to related works using controlled stimulus-response paradigms.
+
+While we believe that our prototype did not perform at a consistently high enough performance to allow for more fine-grained inferences about the the pre-emptive gain it achieved, we maintain that in some cases it elicited a pre-emption that maintained agency to a certain degree.
+We believe this to be a very promising finding because our prototype with low technological requirements still produces behavioral results indicating that intention can be detected and planned movements be augmented.
+However, from stimulus-response paradigms we recall that increasing participants' reaction times by about 80 ms is optimal with regards to maintaining SoA~\cite{Kasahara2019-sk}.
+On one hand, our system is capable of delivering pre-emption within this `SoA-optimal' time range.
+On the other hand, however, the sub-optimal performance of the classifier resulted in a lot of variation due to false positive detections.
+While in some cases, as we observed in the users' comments, the stimulation may have pre-empted a user's motion, false positive stimulation may have had a significant impact on the overall impression of the system.
+
+Taken together, the \textit{key} objective remains in improving the overall performance of the classification system.
 This can be accomplished by fusing several complementary models.
@@
 
-% Conclusion
+% Maybe add something here? If it would work, how can it be used? Implications -> playful experiences, adaptive interfaces
+
 \section{Conclusions}
-In this paper, we demonstrated a system to maintain users' SoA during augmented experiences using brain signals reflecting the intent to (inter-) act.
-In our user study, we found that intentional binding effects were stronger when participants worked with the system to tap on a screen compared to when they were passively moved.
-Furthermore, the level of control working with the system was higher than when being passively moved, as indicated by subjective responses.
+In this paper, we designed and investigated a system to maintain users' SoA during augmented experiences using brain signals reflecting the intent to (inter-) act.
+In our user study, we found no convincing evidence that intentional binding effects are stronger when participants work with an augmentation system compared to being passively moved.
+However, participants rated their level of control working with the system higher than when being passively moved.
 
 We believe this to be an important next step towards augmented users with full integration of the technology~\cite{Mueller2020-dl}.
-Ultimately, our closed-loop system enables novel predictive interfaces that leverage the users' body directly yet still feel \textit{natural} as they align with the users' intention: For example, altering the affordance structure of the interaction in real-time~\cite{Gehrke2022-kz, Lopes2015-ze, Nataraj2020-wm}.
+Ultimately, our closed-loop system is another step towards novel predictive interfaces that leverage the users' body directly yet still feel \textit{natural} as they align with the users' intention.
+For example, we envision that such closed-loop stimulation systems will prove useful in altering the affordance structure of an interaction in real-time~\cite{Gehrke2022-kz, Lopes2015-ze, Nataraj2020-wm}.
 

### 2024-10-04 submission_JNE/source/input/2_introduction.tex
page and jne submission

--- 2024-04-03 input/2_introduction.tex
@@
 Such action augmentations can be realized purely on a software integration level, for example by an AI pair programmer.
-When it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of one's body.
+When it is designed to happen on a hardware level, further challenges emerge, specifically due to shared agency, which in this case means handing over control of one's own body.
 Unfortunately, such augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.
 
-\begin{figure}[h]
+\begin{figure}[!h]
     \centering
     \includegraphics[width=\columnwidth]{figures/teaser.png}%{figures/bci_game.png}
-    \caption{We propose an augmentation system that maintains agency.
+    \caption{We propose an augmentation system that aligns with the user's agency.
 When participants feel the spontaneous urge to move, readiness potentials (RPs) are picked up in the user's brain signals.
 A brain-computer interface (BCI) then predicts the user to be in either an idle (red) or pre-movement (green) state.
-In the latter case, electrical muscle stimulation (EMS) is triggered and the user's hand is moved.
+In the latter case, electrical muscle stimulation (EMS) is triggered and moves the user's hand.
 Image taken with consent from participant.}
+    \label{fig:teaser}
 \end{figure}
 
-Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us.
+Having an SoA means experiencing control over our own voluntary actions, instead of them feeling as randomly happening to us, for a recent review see~\cite{Wen2022-bm}.
 It has been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}.
@@
 
-\subsection{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
-In this paper, we present a prototype to overcome the disruption in experiencing agency during action augmentation.
-We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector.
-The augmentation system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+% \subsection*{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
+With this paper, we strive towards an `agency-aligned' augmentation system.
+We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector, see figure~\ref{fig:teaser}.
+The closed-loop augmentation system (from hereon referred to simply as \textit{system}) controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.
+Ultimately, such an augmentation can cue the user's movements, increase their strength, and might also preempt their action, i.e., increase their speed.
 
@@
 
-Upon predicting a \textit{pre-movement} state, the system augmented the user's action, potentially even preceding their voluntary motor command.
-This augmentation moved the ring finger in accordance with the user's intention to act.
+Upon predicting a \textit{pre-movement} state, the system augments the user's action, potentially even preceding their voluntary motor command.
+This augmentation moves the ring finger in accordance with the user's intention to act.
 We achieved the movement by leveraging electrical muscle stimulation (EMS) applied to the user's forearm flexor muscle.
-To get a first glimpse of how our system can maintain agency in augmented users, we conducted a user study using a mixed-methods research approach.
-In short, we explored whether aligning the physical impact on the user's body with their intention to move preserved their SoA.
+To get a first glimpse of how our system can maintain agency in augmented users, we conducted a user study using both quantitative and qualitative research methods.
+In short, we explored whether aligning the physical impact on the user's body with their intention to move, preserved their SoA.

### 2024-10-04 submission_JNE/source/input/3_related_work.tex
page and jne submission

--- 2024-04-03 UIST_submission/source/input/3_related_work.tex
@@
 \section{Related Work}
-Our research draws inspiration from neuroscience, especially in assessing the SoA qualitatively and quantitatively, and from engineering work on BCIs as well as on physical user augmentation.
+Our research draws inspiration from neuroscience and from engineering work on BCIs as well as on physical user augmentation.
+In order to situate our findings, we briefly review the literature on SoA specifically focusing on what it means to act at one's own volition, being a passive observer, and when acting integrated with a technology.
 
-\subsection{Theories of Sense of Agency}
-The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: When we intentionally perform an action, the brain generates sensory predictions about the action outcome.
-These predictions are constantly compared to the actual sensory data available through the sensory system during the execution of the action.
+\subsection{Theories on Sense of Agency}
+The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: When we act at our own volition and intentionally perform an action, the brain generates sensory predictions about the action outcome.
+These predictions are constantly compared to the actual sensory data available during the execution of the action.
 These include continuous signals such as proprioceptive and visual monitoring of the ongoing movement as well as higher level predictions about the semantic outcome of the action~\cite{Clark2013-ah, Haggard2003-ff, Haggard2017-uv}.
-If no sensorimotor incongruency arises, a SoA manifests.
+If no sensorimotor incongruency arises, and further, the brain attributes subjective causality over the action outcome, SoA manifests.
 
@@
 Subsequently, the tone generated by the key press is evaluated against auditory predictions.
-On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key~\cite{Pangratz2023-ew}.
+On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key, and in general aligns with the subjective goal of the keypress~\cite{Pangratz2023-ew}.
 If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced.
@@
 
-In human-computer interaction (HCI) research, these constructs are categorized using a different terminology.
-Often, the term \textit{pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements.
+In human-computer interaction (HCI), these constructs are often categorized in slightly different terms.
+\textit{Pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements.
 At higher levels of the cognitive hierarchy, \textit{reflective}, i.e.
-conscious, experience is used to refer to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.
+conscious, experience refers to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.
 
-As opposed to intended actions using our own body, movement augmentation hardware allows moving a user's body without their intention.
-Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e., exoskeletons, a user's body can be moved by applying forces to the extremities.
-Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS).
+In order to measure SoA, both explicit and implicit methods have been developed.
+Explicit methods directly query participants to report their subjective experiences using questionnaires.
+Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}.
+In most cases such questionnaires aim at a higher-level, reflective, judgment of agency.
+
+On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
+Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, e.g.
+a sound, participants mentally compress the delay~\cite{Haggard2002-sz}.
+In the theories original formulation, this temporal compression was assumed to only occur following movements that were intended: The action outcome is mentally \textit{bound} to the intention.
+To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay.
+
+More recently it has been shown that this `temporal binding' also manifests when participants are merely a bystander witness in action-outcome scenarios~\cite{Suzuki2019-pi, Gutzeit2023-ei}.
+Among others, Suzuki et al.~\citep{Suzuki2019-pi} showed that temporal binding manifests as an effect of multisensory causal binding unrelated to intention or agency, e.g.
+binding effects were shown in scenarios where one is witnessing a replay of one's own earlier actions.
+Hence, it remains interesting to see how such binding manifests in \textit{integrated} user's.
+
+% integration
+As opposed to acting at one's own volition using one's own body, movement augmentation hardware allows moving a user's body even without their intention.
+Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e., exoskeletons, a user's body can be moved by applying forces to the extremities~\cite{Kuhn2013-ls}.
+Another possibility is to stimulate the brain directly~\cite{Haggard2002-sz}, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS).
 Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle-activating nerves.
-Irrespective of the method applied, the user computes no predictions about the movement and its outcome in these scenarios.
-Thus, concerning the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}.
+Irrespective of the method applied, these technologies allow to move a user's body without the user having generated any predictions about the movement and its outcome.
+However, proprioceptive, visual, and other signals indicate that one's own body is moving.
+Hence, re-afference signals are present without an efference command and copy.
+Thus, concerning the comparator model, a prediction error will arise, negatively impacting SoA~\cite{Wen2020-dk}.
+
+% none or a decreased SoA will be observed in the case of externally controlled actions, for example by brain stimulation or when the body is moved by another person~\cite{Kuhn2013-ls}.
 With something or somebody else moving our body to perform, e.g., a key press on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- negatively impacts the experience of agency.
 
-\subsubsection{Measuring Sense of Agency}
-Both explicit and implicit methods have been developed to evaluate the sense of agency.
+% \subsubsection{Measuring Sense of Agency}
+% Both explicit and implicit methods have been developed to evaluate the sense of agency.
 These methods provide the basis to investigate the effects of action augmentation technology on agency experience.
-Explicit methods directly query participants to report their subjective experiences using questionnaires.
+
+% Explicit methods directly query participants to report their subjective experiences using questionnaires.
 Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}.
-However, in most cases such questionnaires aim at a higher-level, reflective, judgment of agency.
+In most cases such questionnaires aim at a higher-level, reflective, judgment of agency.
 
-On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
+% On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously perceived~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}.
 Seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}.
-Critically, this temporal compression only occurs following movements that were intended.
-The action outcome is mentally \textit{bound} to the intention.
+In the theories original formulation, this temporal compression was assumed to only occur following movements that were intended: The action outcome is mentally \textit{bound} to the intention.
 To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay~\cite{Barlas2018-bq}.
-Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.
+
+% % introduce temporal binding -- intentional binding as a measure frequently used in neuroscience and what is the state-of-the-art of the separation of these phenomena
+% % ~\cite{Wen2020-dk} - prediction errors
+
+% Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.
 
@@
 
-Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA frequently use highly controlled `stimulus-response' paradigms.
+Experimental setups to investigate new `on-body' augmentation technologies that aim to preserve the user's SoA, frequently use highly controlled `stimulus-response' paradigms.
 For example, scenarios where participants are instructed to tap on a touchscreen in response to a presented stimulus on the screen.
@@
 
-This is in line with evidence from cognitive neuroscience which indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
-Hence, after this ``point of no return'' user's struggle to assign a source other than themselves to the action initiation.
+This is in line with evidence from cognitive neuroscience which indicates that from around 200 ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}.
+After this ``point of no return'' user's struggle to assign a source other than themselves to the action initiation.
 Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention to act.
-Cruciall then, a key challenge remaining is to design systems that maintain agency when user's actions are unpredictable and where the experimenter does \textit{not} have executive control over the environment.
+Crucially then, a key challenge remaining is to design systems that maintain agency when user's actions are unpredictable and where the experimenter does \textit{not} have executive control over the environment.
 In other words, how can a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions be designed?
@@
 
-Taken together, these findings demonstrated that the RP is a reliable signal preceding voluntary movement initiation and hence, is a well-suited candidate to base (neuro--)adaptive systems on.
+Taken together, these findings demonstrated that the RP is a reliable signal preceding voluntary movement initiation and hence, is a well-suited candidate to base (neuro)--adaptive systems on.

### 2024-10-04 submission_JNE/source/input/5_user_study_and_methods.tex
page and jne submission

--- 2024-09-25 submission_UIST/source/input/5_user_study_and_methods.tex
@@
 
-In the user study, we compared three conditions of agency experience.
-With the first two conditions, INTENTION and EXTERNAL we queried users at the two edges of agency experience.
-In EXTERNAL, participants had no intention to move and no control over their movement.
-On the other hand, in INTENTION, participants were acting as they would in their day-to-day lives, fully in control.
-We then investigated in a third condition where our AUGMENTED prototype sits on the spectrum between INTENTION and EXTERNAL, in other words, whether it preserved agency and to what degree.
-To answer our question, we employed a mixed methods approach including a psychometric test of intentional binding, one standardized question, and a qualitative (phenomenological) exit interview.
+We compared three conditions of agency experience.
+With the first two conditions, INTENTION and INVOLUNTARY we queried users at the two edges of agency experience.
+In INVOLUNTARY, participants had no intention to move and no control over their movement, much like in a bystander scenario.
+On the other hand, in INTENTION, participants were acting as they would in their day-to-day lives, fully in control and with volition.
+We then investigated in a third condition where our AUGMENTED prototype sits on the spectrum between INTENTION and INVOLUNTARY, in other words, whether it preserved agency and to what degree.
+To answer our question, we employed a mix of qualitative and quantitative methods, including a psychometric test of intentional binding, one standardized question, a qualitative (phenomenological) exit interview, and an exploratory EEG analyses of prediction errors following muscle stimulation.
 
-To assist readers in replicating our prototype and experiment, we provide the necessary technical details, the complete source code for the BCI and the experiment, the collected data, as well as all the analysis scripts\footnote{[anonymized]}
+% To assist readers in replicating our prototype and experiment, we provide the necessary technical details, the complete source code for the BCI and the experiment, the collected data, as well as all the analysis scripts\footnote{[anonymized]}
+% https://github.com/lukasgehrke/2021-fastReach
 
@@
 Prior to their participation, they were informed of the nature of the experiment, recording, and anonymization procedures and signed a consent form.
-The experiment was approved by the ethics committee of [anonymized].
+The experiment was approved by the local ethics committee of the Department of Psychology and Ergonomics at the TU Berlin (Ethics protocol approval code: BPN\_GEH\_2\_230130).
 One participant had to be excluded from further data analyses due to significant deviations from the instructions in the execution of the task.
 Precisely, they did not initiate their movements at their own volition but rather immediately at the onset of each trial, thereby violating the task instruction of waiting 2-3 s after trial onset, see task description below for more detail.
-% All participants were right-handed
 
@@
 \indent\textbf{(4) Experiment Presentation and Collection of Behavioral Responses.} An Acer Group (Acer Inc, Taipeh, Taiwan) tablet was used to present the task to participants and record their behavioral responses.
-In addition to the tablet, we used an external keyboard to allow users to input their timing judgments.
+In addition to the tablet, we used an INVOLUNTARY keyboard to allow users to input their timing judgments.
 
 \subsection{Experimental Task, Design and Procedure}
-Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, EXTERNAL, and AUGMENTED
+Participants performed a simple tapping task in three conditions with 75 trials each: INTENTION, INVOLUNTARY, and AUGMENTED
 
@@
     \includegraphics[width=\textwidth]{figures/method.png}
-    \caption{Diagram summarizing: (a) signal processing and event labeling from muscle activity, (b) extraction of linear slope features from labeled brain signals and (c) real-time application of the BCI gating the muscle stimulation hardware.}
+    \caption{Diagram summarizing(a) signal processing and event labeling from muscle activity, (b) extraction of linear slope features from labeled brain signals, and (c) real-time application of the BCI gating the muscle stimulation hardware.}
     \label{fig:method}
@@
 During INTENTION, participants were equipped with EMG sensors instead of EMS electrodes.
-Participants' reaction times in INTENTION were used to select stereotypical reaction times for the EXTERNAL condition.
+Participants' reaction times in INTENTION were used to select stereotypical reaction times for the INVOLUNTARY condition.
 At the end of INTENTION, EMG electrodes were exchanged for EMS electrodes at the identical location on the forearm.
 
-\indent\textbf{EXTERNAL.} The task structure was identical to INTENTION, however, participants were now instructed to hold and wait for the muscle stimulation to move their finger thereby eliciting the screen tap.
+\indent\textbf{INVOLUNTARY.} The task structure was identical to INTENTION, however, participants were now instructed to hold and wait for the muscle stimulation to move their finger thereby eliciting the screen tap.
 The timing for the EMS trigger was taken by randomly choosing a time between the 5th and 95th percentile of their actual individual reaction time in the INTENTION condition.
@@
 
-The order of the conditions was not pseudo-randomized since training data obtained in INTENTION was required for both EXTERNAL and AUGMENTED.
+The order of the conditions was not pseudo-randomized since training data obtained in INTENTION was required for both INVOLUNTARY and AUGMENTED.
 Furthermore, AUGMENTED was always the last condition, thereby allowing for a prolonged interview immediately after the experience with the prototype.
@@
 The data obtained in INTENTION was used to train the BCI.
-We utilized the [anonymized] and the EEGLAB~\cite{Delorme2004-sn} toolbox inside the MATLAB (The MathWorks Inc.
-Natick, MA, USA) environment for preprocessing both EEG and EMG data.
+For processing EEG and EMG data from INTENTION, we utilized the EEGLAB~\cite{Delorme2004-sn} toolbox with wrapper functions from BeMoBIL-pipeline~\cite{Klug2022-lc} running in the MATLAB (The MathWorks Inc.
+Natick, MA, USA) environment.
 First, to generate behavioral labels at a high temporal resolution for the EEG-based classifier, we leveraged EMG data from the flexor digitorum profundus.
@@
 Next, to label the time of movement onset, the EMG data was averaged across trials for the second preceding the screen tap.
-From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset, see~\ref{fig:method} a.
-% TODO remove anonymization bemobil pipeline
+From this averaged data, the first sample where the EMG amplitude exceeded the 95th percentile was selected as the time of movement onset, see~\ref{fig:method}a.
 
-% TODO/Ignore Next, noisy data segments were determined. 
+% Ignore Next, noisy data segments were determined. 
 % A segment was rejected as an outlier using Matlab's \textit{isoutlier} function using default parameters.
@@
 As single-trial features, the (linear) slope coefficient was obtained for both \textit{idle} and \textit{pre-movement} segments by fitting a linear regression.
-The classifier was then trained using the slope coefficient of the selection of the most discriminative channels as features, generating a feature vector in the dimension of channels that were kept for classification, see~\ref{fig:method} b.
+The classifier was then trained using the slope coefficient of the selection of the most discriminative channels as features, generating a feature vector in the dimension of channels that were kept for classification, see~\ref{fig:method}b.
 We purposefully constrained the dimensionality of the feature vector to avoid over-fitting and decrease the computational load.
@@
 
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figures/eeg_results_new.pdf}
-    \caption{(a) Grand average event-related potential (ERP) at electrode Cz of pre-movement EEG data epochs preceding the last second before a movement, idle epochs of the same trial are plotted alongside but originate from a different time window, see section~\ref{eeg_methods}; (b) Bottom: Slope features for both idle and pre-movement classes at electrode Cz.
-Top: Scalp maps of slope values for both classes and all channels.
-(c): Mean and per participant ROC curves.
-Dotted line at 15 \% false positive rate indicate the selected threshold for the real-time application.}
-    \label{fig:EEG_results}
-\end{figure*}
-
 \subsubsection{Real-time application and EMS control}
@@
 This procedure ran at an update rate of 10 Hz, hence every 100 ms a new prediction was obtained from the classifier.
-To smooth the prediction output to reduce false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging current and the preceding prediction with a weighting (weight of .3 for the preceding, and .5 for the current prediction).
+To smooth the prediction output to reduce false predictions due to unlikely peaks, the predicted probability for the \textit{pre-movement} class was smoothed by averaging the current and the preceding prediction with a weighting (weight of .3 for the preceding, and .5 for the current prediction).
 The weights were obtained through trial-and-error during piloting.
-Then, with a 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s, see~\ref{fig:method} c.
+Then, with a 10 Hz update rate, this smoothed probability, as well as the predicted class for the current frame were gating the EMS switch: when the probability exceeded the threshold and the currently predicted class was \textit{pre-movement}, the switch was opened for 0.5 s, see figure~\ref{fig:method}c.
 
@@
 
-\subsection{Measures of Agency Experience: Intentional Binding, Question \& Interview}
-To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated.
-After each condition we prompted participants to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
+\subsubsection{Post-hoc evaluation of real-time performance}
+To not break the user's focus on the task at hand, (subjective) labels were not obtained during the real-time application in AUGMENTED.
+In other words, participants were not directly queried, e.g.
+via questionnaires, to judge whether the stimulation in the current trial was in line with their intention or not.
+Without these labels, a post-hoc analyses of the binary classifiers performance was not possible, i.e.
+to ascertain for example false positive stimulations.
+However, we explored an alternate approach to estimate (subjective) labels that does not break the user's task immersion.
+We investigated prediction errors in response to the movement onset through event-related potentials (ERPs) at fronto-central electrode FCz.
+Previously, prediction error ERPs at, among others, electrode FCz have been shown to be one suitable candidate feature to detected breaks in (task) immersion, such as when perceiving glitches in VR~\cite{Gehrke2019-og, Gehrke2022-kz, Gehrke2024-xq, Terfurth2024-kh, Si-mohammed2020-ru}.
+In BCIs, these ERPs are frequently leveraged to correct system errors~\cite{Zander2016-ed}.
+In their work,~\citet{Zander2016-ed} demonstrated a BCI to decode a user’s intended cursor movement direction on a 6 × 6 grid.
+The system regularly probed the user by observing the EEG response to random cursor movements.
+How severely the random dot movement violated the user’s intention was directly reflected in the ERP activity.
+We hypothesized to find similar ERP signatures when the stimulation misaligned with the user's prediction.
+
+Since this analyses was conducted post-hoc, it allowed for more signal processing in comparison to the feature extraction from INTENTION when participants were waiting to start the next block.
+With the goal to best recover the prediction error ERP, we again applied `BeMoBIL-pipeline'~\cite{Klug2022-lc} wrapper functions of EEGLAB~\cite{Delorme2004-sn}: After removing non-experiment segments at the beginning and end of the concatenated recording from all three conditions, EEG data was re-sampled to $250~Hz$.
+Next, bad channels were detected using the `FindNoisyChannel' function, which selects bad channels by amplitude, the signal-to-noise ratio, and correlation with other channels~\cite{Bigdely-Shamlo2015-ds}.
+Rejected channels were then interpolated while ignoring the EOG channel, and finally re-referenced to the average of all channels, including the original reference channel FCz.
+After applying a high-pass filter at 1.5 Hz, time-domain cleaning and outlier removal were performed using AMICA auto rejection~\cite{Palmer2011-zs}.
+Eye artifacts were removed using the ICLabel toolbox applied to the results from an AMICA~\cite{Pion-Tonachini2019-fy}.
+For this, ICLabel's popularity classifier was used, meaning that all components having the highest probability for the eye class were projected out of the sensor data.
+
+For all three conditions, ERPs were extracted from band-pass filtered (0.1 Hz to 15 Hz) activity at electrode \textit{FCz}.
+These ERPs were obtained from -1000 ms to 500 ms around the movement onset.
+Trials were excluded in line with the removal for classifier training, see section~\ref{eeg_methods}.
+For all muscle-controlled trials without EMS, movement onset was defined as the time of the tap on the touchscreen minus the `EMG-delay' defined in section~\ref{BCI}.
+For all EMS-controlled trials, movement onset was defined as EMS onset.
+To keep only EMS-controlled trials where the stimulation resulted in an `immediate' screen tap, first all trials were rejected where there was no screen tap in the 350 ms following EMS.
+Next, `extreme outlier removal' was conducted on the delta between EMS and the screen tap using Tukey's method based on the inter-quartile range ~\cite{Tukey1949-sl}.
+In total, 32.9 (SD = 30.3) trials were rejected on average per participant.
+
+% To analyze prediction error ERPs following the movement onset, potentially reflecting a disruptive experience impacting SoA, a linear mixed effects model was fit at each time point of the ERP using the `lme4' package~\cite{Bates2015-bh}.
+The model `ERP\_sample $\sim$ condition + (1|participantID)' was fit with condition reflecting trials belonging to INTENTION, INVOLUNTARY, or AUGMENTED.
+Prior to model fitting, all trials were aggregated per participant and condition.
+A test statistic was obtained by calculating likelihood-ratio tests comparing the full model as specified above against the null model `ERP\_sample $\sim$ 1 + (1|participantID)'.
+All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
+P-values were corrected for multiple comparisons using \textit{false discovery rate}~\cite{Benjamini1995-cw} at alpha = .05.
+
+% Since the AUGMENTED condition contains both, `EMS-controlled' trials, where the classifier detected an intent to interact, as well as `muscle-controlled' trials, where the classifier failed to detect the user's intent and the user carried out the tap without EMS, we split these two trial categories.
+We hypothesized that `EMS-controlled' trials in AUGMENTED \textit{differ} from trials in INVOLUNTARY.
+The latter should elicit a stronger prediction error than the former since no impact of volition is present in INVOLUNTARY, which could potentially moderate the ERP.
+Similarly, we hypothesized that `muscle-controlled' trials in AUGMENTED should \textit{not} differ from INTENTION trials.
+To test these hypotheses, permutation t-tests were conducted on aggregated data using MNE-python~\cite{Gramfort2013-fa} contrasting `EMS-controlled' trials in AUGMENTED with INVOLUNTARY as well as `muscle-controlled' trials in AUGMENTED with INTENTION.
+The tests were conducted for the time window from 150-250 ms post movement-- or EMS-onset event.
+In short, these tests were conducted to investigate the moderating role of the user acting at their own volition in AUGMENTED.
+
+\subsection{Measures of Agency Experience: Intentional Binding, Question \& Interview}% adapt title?
+% To assess if SoA was preserved, an intentional binding measure, a standardized question, and an interview following the AUGMENTED block were investigated. 
+After each condition participants were prompted to rate their experienced agency on a 7-point Likert scale with the statement ``It felt like I was in control of the movements during the task.'', the item was copied from~\cite{Bergstrom2022-fb}.
 Following AUGMENTED we interviewed users about their experience working with the system.
@@
 
-% add back in after anonymization: All interviews were manually transcribed and then automatically translated using }
 We analyzed the interviews by loosely following~\citet{Mayring2015-pp}.
@@
 Ultimately, to assess the classifier performance, we calculated the F1 score, i.e.
-the harmonic mean of precision and recall, and plotted the ROC, see figure~\ref{fig:EEG_results} a to c.
+the harmonic mean of precision and recall, and plotted the ROC, see figure~\ref{fig:EEG_results}a to c.
 
@@
 In line with the literature on intentional binding, we hypothesized that the time intervals should be underestimated for the INTENTION and the AUGMENTED conditions.
-This should not be the case for the EXTERNAL condition where users had no intention to move.
+This should not be the case for the INVOLUNTARY condition where users had no intention to move.
 Hence, when binding occurs, the intervals should be underestimated, hinting at a higher SoA.
 
-To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, EXTERNAL, AUGMENTED) as a fixed effect and `participandID' as a random effect.
+To test this, we fitted a linear mixed effects model with \textit{condition} (INTENTION, INVOLUNTARY, AUGMENTED) as a fixed effect and `participantID' as a random effect on aggregated data.
 The model was specified as `time interval $\sim$ condition + (1|participantID)' and fit using the `lme4' package~\cite{Bates2015-bh}.
-We included all single trial data points in the analyses and did not aggregate before fitting the model as fitting the random effect `participantID' corrects for the correlation of the repeated measurements within participants.
-We obtained a test statistic by calculating likelihood-ratio tests comparing the full model as specified above against the null model `time interval $\sim$ 1 + (1|participandID)'.
+A test statistic was obtained by calculating likelihood-ratio tests comparing the full model as specified above against the null model `time interval $\sim$ 1 + (1|participantID)'.
 All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
-We computed post-hoc pair-wise tests for `condition' corrected for multiple comparisons (Tukey method) using the emmeans package~\cite{Lenth2020-xk}.
+We computed post-hoc pair-wise tests for `condition' corrected for multiple comparisons (Tukey method) using the emmeans package~\cite{Lenth2020-xk}
 
-Next, we hypothesized that subjective ratings of control over the tap movement are comparable between INTENTION and AUGMENTED and lower in the EXTERNAL condition.
+% We included all single trial data points in the analyses and did not aggregate before fitting the model as fitting the random effect `participantID' corrects for the correlation of the repeated measurements within participants.
+
+Next, we hypothesized that subjective ratings of control over the tap movement are comparable between INTENTION and AUGMENTED and lower in the INVOLUNTARY condition.
 Again, we fitted a linear mixed effects model with \textit{condition} as a fixed effect and participant ID as a random effect.
@@
 (2) The subjective feeling of control is comparable between INTENTION and AUGMENTED conditions.
-EXTERNAL should decrease the feeling of control significantly.
+INVOLUNTARY should decrease the feeling of control significantly.
 Additionally, we report the clustered interviews anecdotally.
+
+To analyze prediction error ERPs following the movement onset, potentially reflecting a disruptive experience impacting SoA, a linear mixed effects model was fit at each time point of the ERP using the `lme4' package~\cite{Bates2015-bh}.
+The model `ERP\_sample $\sim$ condition + (1|participantID)' was fit with condition reflecting trials belonging to INTENTION, INVOLUNTARY, or AUGMENTED.
+Prior to model fitting, all trials were aggregated per participant and condition.
+A test statistic was obtained by calculating likelihood-ratio tests comparing the full model as specified above against the null model `ERP\_sample $\sim$ 1 + (1|participantID)'.
+All parameters were estimated by maximum likelihood estimation~\cite{Pinheiro2006-bk}.
+P-values were corrected for multiple comparisons using \textit{false discovery rate}~\cite{Benjamini1995-cw} at $\alpha = .05$.
+
+Since the AUGMENTED condition contains both, `EMS-controlled' trials, where the classifier detected an intent to interact, as well as `muscle-controlled' trials, where the classifier failed to detect the user's intent and the user carried out the tap without EMS, we split these two trial categories.
+We hypothesized that `EMS-controlled' trials in AUGMENTED \textit{differ} from trials in INVOLUNTARY.
+The latter should elicit a stronger prediction error than the former since no impact of volition is present in INVOLUNTARY, which could potentially moderate the ERP.
+Similarly, we hypothesized that `muscle-controlled' trials in AUGMENTED should \textit{not} differ from INTENTION trials.
+To test these hypotheses, permutation t-tests were conducted on aggregated data using MNE-python~\cite{Gramfort2013-fa} contrasting `EMS-controlled' trials in AUGMENTED with INVOLUNTARY as well as `muscle-controlled' trials in AUGMENTED with INTENTION.
+The tests were conducted for the time window from 150-250 ms post movement-- or EMS-onset event.
+In short, these tests were conducted to investigate the moderating role of the user acting at their own volition in AUGMENTED.
 

### 2024-10-04 submission_JNE/source/input/6_results.tex
page and jne submission

--- 2024-04-03 UIST_submission/source/input/6_results.tex
@@
 \section{Results}
-We first report the performance of the classification system, as the system performance greatly impacts the subjective experience of the motor augmentation.
+% We first report the performance of the classification system, as the system performance greatly impacts the subjective experience of the motor augmentation.
 
-\subsection{Classifier Performance}
-Visual inspection of the amplitudes at electrode Cz revealed an increase in the difference between \textit{pre-movement} and \textit{idle} data segments towards the onset of the finger movement, see figure~\ref{fig:EEG_results}a.
-The slope feature for the exemplary channel Cz discriminates well between the two classes (${t_{(10)}} = 4.4, p < .001$), see figure ~\ref{fig:EEG_results}b bottom.
-The scalp maps in figure~\ref{fig:EEG_results}b top show the (color-coded) mean slope for each channel and each class.
-Central channels on the contralateral side to the moving finger on the right hand show a negative slope for the pre-movement class and a neutral slope for the idle class.
-Furthermore, differences in slope were also observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifested only for the idle class.
+% \subsection{Behavior: Intentional Binding}
+In the intentional binding task, i.e.
+the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (INTENTION M = -160.7 ms, SD = 88.1; INVOLUNTARY M = -135.3, SD = 110.3, AUGMENTED M = -162.4, SD = 113.8).
+The underestimation was not affected by the condition, see~\ref{fig:ib_task}.
 
-The grid-search over channels resulted in the BCI leveraging on average 11.2 (SD = 3.6) channels.
-Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9 and AF3.
-The classifier cross-validation resulted in a mean F1 score of .71 (SD = .03), see figure~\ref{fig:EEG_results}c.
-We set the detection threshold to 15 \% false positive rate and at that rate, observed a mean threshold of 57 \% (SD = .04).
-Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57 \% probability.
-
-\subsection{Behavior: Intentional Binding}
-In the intentional binding task, i.e.
-the estimation of the time interval between tap and tone, participants generally underestimated the average real delay (350ms) in all three conditions (INTENTION M = 187.5, SD = 139; EXTERNAL M = 208.6, SD = 140.1, AUGMENTED M = 196.0, SD = 138.2).
-The condition influenced the estimation (${\chi^{2}_{(2)}} = 17.1, p < .001$), see figure~\ref{fig:results}a.
-Post-hoc tests revealed that the underestimation was less pronounced in EXTERNAL as compared to INTENTION ($beta = 22.6, p < 0.001$).
-This was also the case in EXTERNAL, with less pronounced underestimations of the time interval between tap and tone compared to AUGMENTED ($beta = 15.3, p = 0.018$).
+% The condition influenced the estimation (${\chi^{2}_{(2)}} = 17.1, p < .001$), see figure~\ref{fig:results}a.
+Post-hoc tests revealed that the underestimation was less pronounced in INVOLUNTARY as compared to INTENTION ($beta = 22.6, p < 0.001$).
+This was also the case in INVOLUNTARY, with less pronounced underestimations of the time interval between tap and tone compared to AUGMENTED ($beta = 15.3, p = 0.018$).
 No difference was observed between INTENTION and AUGMENTED.
-
-\subsection{Subjective Ratings \& Reports}
-The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 43.7, p < .001$), see figure~\ref{fig:results}b.
-Post-hoc tests revealed that participants rated their motion control higher in INTENTION compared to EXTERNAL ($beta = 4.8, p < .0001$), and higher in INTENTION compared to AUGMENTED ($beta = 3.1, p < .0001$).
-Further, higher control was observed in AUGMENTED compared to EXTERNAL ($beta = -2.9, p = .006$).
 
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/subjective_results.pdf}
-    \caption{(a) Temporal binding effect over conditions.
+    \includegraphics[width=\columnwidth]{figures/ib_task.pdf}
+    \caption{Difference in time estimation from real delay in intentional binding task for the three experimental conditions (left).
+Negative values indicate a temporal compression, i.e., temporal binding.
+Right side: Trials in the AUGMENTED condition split in EMS-- and self-executed trials.}
+    \label{fig:ib_task}
+\end{figure}
+
+\subsection{Subjective Ratings \& Reports}
+The subjective rating of control differed between conditions (${\chi^{2}_{(2)}} = 43.7, p < .001$), see figure~\ref{fig:loc}b.
+Post-hoc tests revealed that participants rated their level of control higher in INTENTION compared to INVOLUNTARY ($beta = 4.8, p < .0001$), and higher in INTENTION compared to AUGMENTED ($beta = 3.1, p < .0001$).
+Further, higher control was observed in AUGMENTED compared to INVOLUNTARY ($beta = -2.9, p = .006$).
+
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\columnwidth]{figures/subjective_results.pdf}
+%     \caption{(a) Temporal binding effect over conditions.
 With an average presented time interval of 350ms, lower values in the estimated time interval indicate an underestimation, i.e., temporal binding.
@@
 Significance labels obtained from post-hoc tests on estimated marginal means.}
-    \label{fig:results}
+%     \label{fig:results}
+% \end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.5\columnwidth]{figures/level_of_control.pdf}
+    \caption{Subjective ratings of control across conditions. Significance labels obtained from post-hoc tests on estimated marginal means.}
+    \label{fig:loc}
 \end{figure}
@@
 
-Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly involuntary, but partly as if it was following my decision/choice.''.
+Some perceived the stimulation as externally triggered, yet partially responsive to their choices, resulting in a sentiment captured by, ``I think that the stimulation in the third block was partly random, but partly as if it was following my decision/choice.''.
 One participant noted that they felt that ``somehow the information is in my arm''.
@@
 
+\subsection{Classifier Performance}
+Visual inspection of the amplitudes at electrode Cz revealed an increase in the difference between \textit{pre-movement} and \textit{idle} data segments towards the onset of the finger movement, see figure~\ref{fig:EEG_results}a.
+The slope feature for the exemplary channel Cz discriminates well between the two classes (${t_{(10)}} = 4.4, p < .001$), see figure ~\ref{fig:EEG_results}b bottom.
+The scalp maps in figure~\ref{fig:EEG_results}b top show the (color-coded) mean slope for each channel and each class.
+Central channels on the contralateral side to the moving finger on the right hand show a negative slope for the pre-movement class and a neutral slope for the idle class.
+Furthermore, differences in slope were also observed at frontal electrodes over the left hemisphere as well as parietal electrodes, were a positive slope manifested only for the idle class.
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figures/eeg_results_new.pdf}
+    \caption{(a) Grand average event-related potential (ERP) at electrode Cz of pre-movement EEG data epochs preceding the last second before a movement, idle epochs of the same trial are plotted alongside but originate from a different time window, see section~\ref{eeg_methods}; (b) Bottom: Slope features for both idle and pre-movement classes at electrode Cz.
+Top: Scalp maps of slope values for both classes and all channels.
+(c): Mean and per participant ROC curves.
+Dotted line at 15 \% false positive rate indicate the selected threshold for the real-time application.}
+    \label{fig:EEG_results}
+\end{figure*}
+
+The grid-search over channels resulted in the BCI leveraging on average 11.2 (SD = 3.6) channels.
+Besides channels C3, C4, and Cz, that were always included, other common channels (retained for at least 3 participants) included FT9 and AF3.
+The classifier cross-validation resulted in a mean F1 score of .71 (SD = .03), see figure~\ref{fig:EEG_results}c.
+We set the detection threshold to 15 \% false positive rate and at that rate, observed a mean threshold of 57 \% (SD = .04).
+Hence, on average, the classifier switched on the EMS when it predicted class \textit{pre-movement} with 57 \% probability.
+
+\subsubsection{Post-hoc analyses of classifier performance}
+
+In line with the readiness potential a negative going deflection over the last second preceding the movement onset was present in both INTENTION and AUGMENTED, but not in INVOLUNTARY, see figure~\ref{fig:erp}a.
+Furthermore, ERPs differed between the three conditions in the time window from 210 ms -- 250 ms following movement onset ($\chi^{2}_{(2} = 10.6, p = .03$ at 220 ms).
+INVOLUNTARY exhibited a strong negativity peak (strongest among the three conditions), peaking at around -10 $\mu$V at 210 ms after movement onset.
+A weaker peak was observed for AUGMENTED (-6 $\mu$V at 210 ms) and even more so for INTENTION (-2 $\mu$V at 150 ms).
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/PE_erp.pdf}
+    \caption{ERPs at electrode FCz from -1000 before, to 600 ms after the tapping movement onset.
+Top (a): INTENTION, INVOLUNTARY and AUGMENTED trials.
+Red bar on top indicate main effect at $p<.05$.
+Bottom (b): Difference ERPs at FCz of the `muscle-controlled' trials in AUGMENTED minus INTENTION as well as EMS trials in AUGMENTED minus INVOLUNTARY.}
+    \label{fig:erp}
+\end{figure}
+
+When correcting for the influence of the EMS by subtracting INVOLUNTARY from EMS trials in AUGMENTED an INTENTION from `muscle-controlled' trials in AUGMENTED we observed no differences.
+Visually, it appears that trials in both INVOLUNTARY and INTENTION condition exhibit a stronger negativity in the post movement time window of interest than respective EMS and `muscle-controlled' trials in AUGMENTED, see the positive deflections peaking at 300 ms post movement onset in figure~\ref{fig:erp}b.
+
 %%% Resources and Ideas
